### 2018-02-07 doc/ancestralPCA.tex
Add important existing files

\documentclass{article}[12pt]

\usepackage{epsfig} \usepackage[margin=1.5in]{geometry}
\usepackage[semicolon,authoryear]{natbib} \bibliographystyle{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{cleveref}
\usepackage{verbatim}

\newcommand{\bO}{\mathcal{O}}

\title{Computational Methods for Principle Component Analysis Applied to Ancestry Estimation}

\author{David Zhang}

\begin{document}

\maketitle

\section{Principle Component Analysis (PCA)}

\subsection{What is PCA}

Let there be $n$ individuals and $p$ features.
Moreover, let these points be centered.
Then we have $n$ points in a $p$-dimensional space,
with the origin being the center of mass of these points.
For any direction, we can project the points to it and find the variance of the projected points.
We are interested in the direction of the greatest variance.
This direction is the linear combination of the features that distinguishes the individuals the most,
which we call the first principle component (PC1).
We repeat the process in the quotient space induced by PC1 to find the second principle component (PC2) and so forth.
By using the first $k$ PC's, we can reduce the $p$-dimensional data to $k$-dimensional at the least cost of losing information.

\subsection{How to find the PCs}
To find the PC's, we record the information of this group in a $p \times n$ matrix $X$ and conduct singular value decomposition (SVD) to it.
SVD gives us the spectrum decomposition $X = UDV^T$, where $U$ is $p \times n$, $D$ and $V$ are $n \times n$.
Here $U$ and $V$ are unitary matrices,
and their columns are the eigenvectors of $XX^T$ and $X^T X$, respectively.
And $D$ is a diagonal matrix,
whose diagonal entries are the square roots of the nonzero eigenvalues of $X X^T$ and $X^T X$.
In other words, $X X^T = U D^2 U^T$ and $X^T X = V D^2 V^T$.
For PCA, the columns of $U$ are the first $n$ PC's,
the diagonal entries of $D$ are the standard deviations of the $n$ points projected to the $n$ PC's,
and the rows of $V$ are the $n$ points' standardized PC scores on the $n$ PC's,
which can be seen as $n$ points' coordinates in the space spanned by the $n$ PC's.

For more information on PCA, see \cite{jolliffe}.

\section{Application of PCA to Genotyping Data}

\subsection{The format of genotyping data}

PCA can be applied to genotyping data.
In particular, it is used to analyze single nucleotide polymorphisms (SNPs).
A SNP is a variation in a single nucleotide.
For example, most individuals may have an A base at a specific base position, but a minority of individuals may have a C base at this site.
Here we say that there is a SNP at this base position,
and A and C are said to be its major and minor alleles, respectively.
Since chromosomes come in pairs,
at each base position the minor allele may appear zero, one, or two times.
Thus by looking at certain base positions,
we receive a data of the form of a string of $0$, $1$, and $2$,
of length as large as hundreds of thousands.

\subsection{The usefulness of PCA on genotyping data}

The format of genotyping data makes it suitable for PCA.
A group of individuals is first selected as the reference group.
A testing individual is combined with the reference individuals to find its PC scores.
By using the first few PCs (often times four PCs are informative enough), 
we are able to convert the data from the form $3^p$ with $p \approx 10^6$ to $\mathbb{R}^k$ with $k \approx 4$,
which significantly reduces the complicatedness of the data and make it more intuitive to comprehend.
For example, we can find how similar is the testing individual to any of the reference individuals simply by eyeballing their PC1 vs PC2 plot.

\subsection{Ancestry estimation by PCA}

By applying PCA to genotyping data, we can retrieve information about many different aspects of an individual.
In this research project, we focus on using PCA to estimate an individual's ancestry history from his or her genotyping data.
Ancestry history is often a confounder for the association between certain gene and the trait being studied, such as having typej-II diabetes.
By estimating an individual's ancestry history,
researchers can eliminate the bias in the association between the gene and the trait caused by ancestry differences.
It has been found that individuals with the same ancestry history score similarly on the first four PCs of their genotyping data (see \cite{wang}).
Thus PCA provides a way to convert a person's ancestry history into real numbers,
which is useful for numerically estimating the previously mentioned confounding effect induced by ancestry history.

\section{Computational Methods for PCA}

There are at least two methods for finding the testing individual's PC scores.
Here we call them the ``once-for-everyone'' method and the ``once-for-all'' method.

\subsection{The ``once-for-everyone'' method}

In this method, we append the testing individual's genotyping data to the reference group's genotype matrix and compute the PC scores of the amended matrix.
This is the most standard and reliable computational method for PCA.
However, this method is relatively slow in computation.
Here all the reference individuals' PC scores must be recomputed when a new testing individual arrives.

\subsubsection{Standard implementation}

One of such implementation is the TRACE program by Wang et al.
An important addition in this program to the description above is Procrustes analysis.
Given two sets of points that are of the same number but live in spaces of different dimensions,
Procrustes analysis helps us to find a projection
that transforms the higher-dimensional set to the lower-dimensional set
such that the distance between the points after the transformation is the least.
This transformation improves the accuracy of the estimated ancestry.

The detailed algorithm of TRACE runs as follows.
Let $X$ be the $p \times n$ reference genotyping data matrix where each column represents an individual.
Moreover, let $y$ be the $p \times 1$ matrix containing the testing individual's genotyping data.
\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$ (columns of $D$ and $V$ are sorted by PC variance with the first column corresponding to the greatest PC variance).
  ($\bO[n^3]$.)
\item Append $X^T y$ and its transpose to the last column and row of $X^T X$, respectively.
  The entry of the bottom-right corner is $y^T y$.
  This is exactly the same as computing $X'^T X'$,
  where $X' = (X, y)$ is the reference group with the testing individual added.
  ($\bO[np]$.)
\item Apply eigendecomposition to $X'^T X'$ to get $X'^T X' = V' D'^2 V'^T$.
  ($\bO[n^3]$.)
\item Project the first $k_2$ columns of $V'$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the testing individual (the last row of $V$),
  where $k_1 < k_2 << n$.
  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
  \item Go to step 3 for the next testing individual.
\end{enumerate}

This is a simplified version of the TRACE algorithm.
Some minor adjustments in the program are not mentioned above.
TRACE actually uses the unstandardized PC scores $VD$ rather than the standardized PC scores $V$.
The two treatments of the PC scores have the same computational complexity and the results are equivalent.
Moreover, TRACE has a slightly more sophisticated treatment of missing data.
In this report, we assume that no missing data exists to simply the comparison of the three methods.

By adding up the computational complexity of each step, we can see that the total computational complexity of TRACE is $\bO[n^2p + m(np + n^3)]$,
where $m$ is the number of testing individuals.

For more information about TRACE,
see \cite{wang} and TRACE's manual.

\subsubsection{Improvments}

The algorithm above can be speeded up by using a different way to calculate the PC scores, that is, matrix $V'$.
Notice that TRACE only uses the first $k_2$ columns of $V'$.
Thus we take advantage of this and use a method called online SVD that gives a close approximation of only some of the columns of $V'$ and saves computational complexity.
The detailed algorithm is:

\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
  ($\bO[n^3]$.)
\item Calculate $U_3 = X V_3 D_3^{-1}$,
  where $V_3$ is the first $k_3$ columns of $V$,
  and $D_3$ is the diagonal matrix of the first $k_3$ diagonal entries of $D$.
  ($\bO[k_3 n p]$.)
\item Calculate 
  \[
    L = U_3^T y \quad \text{and} \quad K = y^T H,
  \]
  where $H$ is the normalized  $y - U_3L$
  ($\bO[k_3 p]$.)
\item Calculate $Q^T Q$, where
  \[
    Q = 
    \begin{bmatrix}
      D_3 & L \\
      0 & K
    \end{bmatrix}.
  \]
  ($\bO[k_3^3]$.)
\item Apply eigendecomosition to $Q^T Q$ to get $Q^T Q = V''_3 D''^2_3 V''^T_3$.
  ($\bO[k_3^3]$.)
\item Calculate
  \[
    V'_3 =
    \begin{bmatrix}
      V_3 & 0 \\
      0 & 1
    \end{bmatrix}
    V''_3.
  \]
  ($\bO[nk_3^2]$.)
\item Project the first $k_2$ columns of $V'_3$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the testing individual,
  where $k_1 < k_2  < k_3 << n$.
  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
  \item Go to step 4 for the next testing individual.
\end{enumerate}

The total computational complexity for $m$ testing individuals is $\bO[n^2p + n^3 + k_3np + m(k_3p + k_3^3 + nk_3^2])$, which is approximately $\bO[n^2p + m(k_3p + k_3^2n)]$ since $k_3 << n << p$.
This is a big save on the runtime compared to the TRACE algorithm,
if $k_3$ is small compared to $n$.

For more details of the online SVD method, see \cite{brand}.

\subsection{The ``once-for-all'' method}

In the ``once-for-everyone'' method,
we have to decompose the data matrix of both the reference group and the testing individual for every single testing individual.
Alternatively, we can find the PCs of the reference group and project the testing individual's genotyping data to that direction.
This way, eigendecomposition or SVD is done only once,
which significantly reduces computational complexity.
However, this method tends to underestimate the magnitude of the testing individual's PC scores when $n << p$,
which is often the case for genotyping data,
as discussed in \cite{dey}.
This is called the ``shrinking bias''.
This problem has been dealt with by the R package HDPCA by Dey.
The algorithm runs as follows:

\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
  ($\bO[n^3]$.)
\item Calculate $W = (Y^T X) V D^{-1}$, where $Y$ is the data matrix for all the testing individuals. ($\bO[mpn + mn^2]$.)
  \item Input $D$ and $W$ to the pc\_adjust function in the HDPCA function to obtain the PC scores adjusted for the shrinking bias. ($\bO[p]$ for finding the number of distant spikes (function select.nspike), $\bO[n]$ for adjusting for the shrinking bias (function hdpc\_est))
\end{enumerate}

Then the total computational complexity is $\bO[n^2p + mn(p + n) + p + n]$,
which is approximately equal to $\bO[n^2p + mn(p+n)]$,
given that $m$ or $n$ is large.
The comparison of computational complexity of the three methods is shown in \Cref{tbl:cplx}.

\begin{table} 
  \centering
  \begin{tabular}{|l|l|}
    \hline
    Method & Computational Complexity \\ 
    \hline
    TRACE & $\bO[n^2 p + mn(p + n^2)]$ \\
    \hline
    TRACE + online SVD & $\bO[n^2 p + mk_3(p + k_3 n)]$ \\
    \hline
    HDPCA & $\bO[n^2p + mn(p+n)]$ \\
    \hline
  \end{tabular}
  \caption{
    Comparison of computational complexity for the three methods.
    Here $p$ is the number of SNPs,
    $n$ is the number of individuals in the reference group,
    and $m$ is the number of individuals in the  testing group.
    For Online TRACE, $k_3$ is the number of PCs calculated by online SVD, which is usually much less than $n$.}
  \label{tbl:cplx}
\end{table}

\section{Runtime Comparison}

We now do an empirical test to compare the efficiency of the three methods.

\subsection{Dataset}

We use the datasets from the Human Genome Diversity Project (HGDP) and the 1000 Genomes Project (1000GP).
The HGDP dataset has genotyping data about 600k SNPs on 900 individuals with different ancestry history.
The 1000GP dataset contains most of the SNPs in the HGDP dataset as well as many others.
Its sample size is about 2300.

For our test, we use the 1000GP dataset but only the SNPs that also appear in the HGDP dataset,
which are as many as about 570k.
Moreover, we remove the individuals who has at least one parent that is already in the dataset.

\subsection{Data Format}

We use TRACE's geno format.
Each row contains the genotype of an individual.
The first two columns are the individual's ID and ancestral group.
Starting at the third column, each column represents a SNP and is either 0, 1, or 2, which is the number of mutations, or -9, which is the value for missing data.
Each SNP's reading is centered and normalized before applying the PCA methods,
with the missing data replaced by the mean.

The vcf format is also acceptable, since TRACE provides the vcf2geno program that converts vcf files to geno files.

\subsection{Setting}

\section{Result}

We use the 1000Genomes dataset for testing the performance of the three methods.
Only the SNP's that have a minor allele frequency greater than 0.05 and appear in the HGDP dataset are selected.
Individuals whose parents are in the dataset are removed.
This filtering leaves us 570859 SNP's and 2481 individuals.
For the testing group, we use the 100 individuals at the bottom of the dataset.
For the reference group, we use the $n$ individuals at the top of the dataset,
where $n$ is set to 500, 1000, 1500, 2000, and 2300 to compare the runtimes.

\subsection{Estimations}

The PC scores calculated by the three methods are almost identical,
with the only exception of the PC3 and PC4 scores calculated by HDPCA for $n = 500$ being slightly different from those calculated by the other two methods.
(See the pdf files in the Google Drive folder for the plots.)

\subsection{Runtimes}

The three methods' runtimes for the learning phase are not different from each other by much.
However, for the analysis phase, Online PCA and HDPCA are significantly more efficient as $n$ increases.
This empirical result is consistent with the theoretical computational complexity.
In \Cref{tbl:runLearn} and \Cref{tbl:runAna},
we show the runtimes for different reference sizes.
We break the runtimes into two phases: learning and analysis.
The learning phase is required for only once,
and the analysis phase is required for every testing individual.

\begin{table} 
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    nTrain & 500 & 1000 & 1500 & 2000 & 2300 \\ 
    \hline
    TRACE & 138 & 482 & 1022 & 62557 & 85913 \\
    \hline
    TRACE + online SVD & 152 & 505 & 1407 & 36824 & 79027 \\
    \hline
    HDPCA & 139 & 477 & 1368 & 35825 & 76700 \\
    \hline
  \end{tabular}
  \caption{Empirical runtimes (sec) for the learning phase for different reference sizes}
  \label{tbl:runLearn}
\end{table}

\begin{table} 
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    nTrain & 500 & 1000 & 1500 & 2000 & 2300 \\ 
    \hline
    TRACE & 59 & 247 & 729 & 8531 & 7851 \\
    \hline
    TRACE + online SVD & 135 & 131 & 132 & 138 & 140 \\
    \hline
    HDPCA & 79 & 119 & 194 & 390 & 463 \\
    \hline
  \end{tabular}
  \caption{Empirical runtimes (sec) for the analysis phase for different reference sizes}
  \label{tbl:runAna}
\end{table}

\section{Discussion}

\begin{comment}

\section{Old}

Alternatively, we can find the PCs (directions of the greatest variances) of the reference group,
and then project the testing individual's data to them to find the PC scores.
The advantage of this method is that the PCs for the reference group only need to be calculated once for all testing individuals.
This method, though fast, tends to underestimate the magnitude of the PC scores when compared to the results given by the first method.
This ``shrinking bias'' is discussed in \cite{dey}.

This research project studies how either method can be improved.
The first method can be speeded up by alternating the SVD algorithm,
so that the computational complexity can be reduced at little cost of the accuracy of the result.
For the second method, the accuracy of the estimation can be improved by calculating and adjusting for the shrinking bias.
In this report, we compare the accuracy and runtimes of different ways to improve the two methods.
\subsection{TRACE}

TRACE has the most conceptually simple algorithm among the three methods.
Let $X$ be the $p \times n$ matrix ($n << p$) that contains the reference genotype,
where each row corresponds to a SNP and each column corresponds to an individual.
Moreover, let $y$ be the $p \times 1$ matrix of the genotype of the testing individual. The algorithm runs as follows:

\subsection{Online PCA}

The Online PCA method has the same algorithm except for calculating the testing individual's PC scores.
Notice that in TRACE, only the first $k_2$ PC's (that is, the first $k_2$ columns of $V'$) are used for the testing individual.
The Online PCA method can compute for the first $k$ PC scores of a data matrix and skip the rest.
The result is an approximation, but the runtime is significantly reduced.

More specifically, Online PCA takes the following steps:

\subsection{HDPCA}

A simple way to compute the PC scores for the testing individual is $w^T = y^T U$.
However, the resulted PC scores tend to be smaller than the PC scores obtained from eigendecomposing $X'^T X$ where $X^T = (X, y)$.
This shrinking bias can be adjusted by the HDPCA package in R.
The algorithm is as follows:

\end{comment}

\newpage

\begin{thebibliography}{}

\bibitem[Brand(2002)]{brand} Brand, M. (2002). Incremental singular value decomposition of uncertain data with missing values. {\it Computer Vision--ECCV 2002}, 707-720.

\bibitem[Dey \& Lee(2016)]{dey} Dey, R., \& Lee, S. (2016). Asymptotic properties of Principal Component Analysis and shrinkage-bias adjustment under the Generalized Spiked Population model. {\it arXiv preprint arXiv:1607.08647}.

\bibitem[Jolliffe(2002)]{jolliffe} Jolliffe, I. (2002). {\it Principal component analysis}. John Wiley \& Sons, Ltd.
Chicago

 \bibitem[Wang et al.(2015)]{wang} Wang, C., Zhan, X., Liang, L., Abecasis, G. R., \& Lin, X. (2015). Improved ancestry estimation for both genotyping and sequencing data using projection procrustes analysis and genotype imputation. {\it The American Journal of Human Genetics, 96}(6), 926-937.

\end{thebibliography}

\end{document}

### 2018-02-07 doc/paper.tex
Add important existing files

\documentclass{article}[12pt]

\usepackage{epsfig} \usepackage[margin=1.5in]{geometry}
\usepackage[semicolon,authoryear]{natbib} \bibliographystyle{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{cleveref}
\usepackage{verbatim}
\usepackage{graphicx}

\newcommand{\bO}{\mathcal{O}}
\newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg\;min}}\;}

\title{Principle Component Analysis Computational Methods
  for Estimating Ancestral Histroy from Genotype Data}

\begin{document}

\maketitle

\section{Introduction}

\section{Method}

\subsection{Model}

\begin{enumerate}
\item Data matrices:\\
  The data matrix contains the genotypes for the individuals.
  Each row represents a SNP, and each column represents an individual.
  Each entry is $0$, $1$, or $2$.
  The genotypes of the reference group are stored in the $p \times n$ matrix $X$,
  and those of the study group are stored in the $p \times m$ matrix $Y$.
  For most genetic data, $n \ll p$ and $m \ll p$.
\item Principle component analysis \\
  Given a multi-dimensional data, principle component analysis (PCA) looks for the directions, which are called the principle components (PC), in which the data has the greatest variance.
  The PC's form an orthonormal basis for the sample space,
  and we can ignore all but the first few PC's to reduce the dimension of the data.
  In particular, let $X=UDV^T$ be the singular value decomposition (SVD) of $X$,
  where $U$ is a $p \times n$ matrix whose columns are the PC's,
  $D$ is an $n \times n$ diagonal matrix storing the standard deviations of the PC's,
  and $V$ an $n \times n$ matrix such that each row contains the standardized PC scores of an individual.
  (Alternatively, we can also use eigendecomposition to find the matrices since $X^T X = VD^2V^T$ and $X X^T = UD^2U^T$.)
  We use $D$ to sort out the most influential $k$ PC's ($k \ll n$) and look at the coresponding columns of $V$ to find their PC scores.
  For estimating ancestral histroy from genotypes, often the first $4$ PC's will be sufficient for providing the information we need.

  PCA allows us to use the reference data to find the PC scores of the study data.
  The standard method is to append a study individual's genotype to the reference matrix and then conduct a SVD of the augmented matrix.
  The corresponding row in the matrix $V$ is the study individual's PC scores.
\end{enumerate}

\subsection{Computational Methods}

The usefulness of PCA for genetic data largely depends on its computational cost.
The computational complexity of SVD for a $p \times n+1$ matrix is $\min((n+1)^2p, (n+1)p^2) = n^2p $ for $1 \ll n \ll p$.
Thus the total computational complexity for performing PCA on $m$ individuals by using the standard method is $mn^2p$.
This value increases quadratically as we increase the reference sample size to increase the accuracy of the PC scores.
Alternative computational methods have been developed to improve the efficiency of PCA.
In this study, we compare four such methods: TRACE, projection, online TRACE, and HDPCA projection.

\subsubsection{TRACE}

TRACE implements the standard PCA method, with additional procedures for handling minor issues such as missing data.
An important improvement in TRACE on the standard method is the inclusion of Procrustes analysis.
Given two data sets with the same sample size and possibly different dimensions,
Procrustes analysis finds the linear transformation that maps one data matrix as close as possible to the other data matrix.
In our contex, the PC scores of the reference matrix is computed and we obtain the $n \times n$ matrix $V$.
For a study individal with genotype $y$, we decompose $\tilde{X} = (X, y)$ into $\tilde{X} = \tilde{U}\tilde{D}\tilde{V}^T$.
(TRACE uses eigendecomposition to find the ``$V$'' matrix to reduce computational complexity since we do not need the ``$U$'' matrix here.)
Let
\[
  V =
  \begin{bmatrix}
    V_{\text{ref}} & W
  \end{bmatrix},
  \quad
  \tilde{V} =
  \begin{bmatrix}
    \tilde{V}_{\text{ref}} & \tilde{W}_\text{ref} \\
    \tilde{V}_\text{stu} & \tilde{W}_\text{stu}
  \end{bmatrix},
\]
where the dimensions of $V_\text{ref}$, $\tilde{V}_\text{ref}$, and $\tilde{V}_\text{stu}$ are $n \times k_1$, $n \times k_2$, and $1 \times k_2$, respectively, with $ 1 \leq k_1 \leq k_2 \leq n+1$. 
We use Procrustes analysis to find
\[
  A, b = \argmin{A, b} d(V_\text{ref}, \tilde{V}_{\text{ref}}A+b)
\]
for some metric $d$.
Finally, we get the Procrustes-adjusted top $k_1$ PC scores for the study individual $V_\text{pro} = \tilde{V}_\text{stu} A + b$.
The algorithm is summarized as follows:
\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$ (columns of $D$ and $V$ are sorted by PC variance with the first column corresponding to the greatest PC variance).
  ($\bO[n^3]$.)
\item Append $X^T y$ and its transpose to the last column and row of $X^T X$, respectively.
  The entry of the bottom-right corner is $y^T y$.
  This is exactly the same as computing $\tilde{X}^T \tilde{X}$,
  where $\tilde{X} = (X, y)$ is the reference group with the testing individual added.
  ($\bO[np]$.)
\item Apply eigendecomposition to $\tilde{X}^T \tilde{X}$ to get $\tilde{X}^T \tilde{X} = \tilde{V} \tilde{D}^2 \tilde{V}^T$.
  ($\bO[n^3]$.)
\item Project the first $k_2$ columns of $\tilde{V}$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the testing individual (the last row of $V$),
  where $k_1 < k_2 \ll n$.
  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
  \item Go to step 3 for the next testing individual.
\end{enumerate}

Total Computational complexity: $\bO[n^2p + m(np + n^3)]$.

TRACE is the slowest among the computational methods we study.
However, its result is also the most reliable.

\subsubsection{Projection}
An exchange for runtimes with accuracy is the projection method.
For this method, we simply find the PC's of the reference matrix and project the study individual's genotype onto them,
as shown below:

\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
  ($\bO[n^3]$.)
\item Calculate $W = Y^T (X V D_1^{-1})$, where $Y$ is the data matrix for all the testing individuals, $D_1$ the first $k_1$ columns of $D$. ($\bO[mpk_1 + pnk_1]$.)
\end{enumerate}

Total computational complexity: $\bO[n^2p + n^3 + mpk_1 + pnk_1)] \approx \bO[n^2p + mk_1p)]$,
given that $k_1 \ll n \ll p$.

As the reference sample size increases,
the computational complexity of the projection method only increases quadratically,
(except for decomposing the reference matrix, which is done once for all study individuals,)
while that of TRACE increases cubically.
However, it is known that the projection method tends to underestimates the magnitude of the PC scores when $n \ll p$,
which is known as the ``shrinking bias''.

\subsubsection{Online TRACE}

A remedy for TRACE is to reduce the computational complexity withou much sacrifice of the accuracy.
This is achieved by the online TRACE method.
Notice that in TRACE, only the top $k_2$ PC's are used.
Online TRACE calculates a close approximation of the first $k_3$ columns of $\tilde{V}$ for some $k_2 \leq k_3 \ll n$ and keeps everything else the same as in TRACE, including Procrustes analysis. 
The algorithm is
\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
  ($\bO[n^3]$.)
\item Calculate $U_3 = X V_3 D_3^{-1}$,
  where $V_3$ is the first $k_3$ columns of $V$,
  and $D_3$ is the diagonal matrix of the first $k_3$ diagonal entries of $D$.
  ($\bO[k_3 n p]$.)
\item Calculate 
  \[
    L = U_3^T y \quad \text{and} \quad K = y^T H,
  \]
  where $H$ is the normalized  $y - U_3L$
  ($\bO[k_3 p]$.)
\item Calculate $Q^T Q$, where
  \[
    Q = 
    \begin{bmatrix}
      D_3 & L \\
      0 & K
    \end{bmatrix}.
  \]
  ($\bO[k_3^3]$.)
\item Apply eigendecomosition to $Q^T Q$ to get $Q^T Q = \ddot{V}_3 \ddot{D}^2_3 \ddot{V}^T_3$.
  ($\bO[k_3^3]$.)
\item Calculate
  \[
    \tilde{V}_3 =
    \begin{bmatrix}
      V_3 & 0 \\
      0 & 1
    \end{bmatrix}
    \ddot{V}_3.
  \]
  ($\bO[nk_3^2]$.)
\item Project the first $k_2$ columns of $\tilde{V}_3$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the testing individual,
  where $k_1 < k_2  < k_3 \ll n$.
  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
  \item Go to step 4 for the next testing individual.
\end{enumerate}

Total computational complexity: $\bO[n^2p + n^3 + k_3np + m(k_3p + k_3^3 + nk_3^2]) \approx \bO[n^2p + m(k_3p + k_3^2n)]$, provided $k_3 \ll n \ll p$.

Here we see that the computational complexity of online TRACE for the study individuals increases quadratically with respect to the reference sample size.
This makes the runtimes of online TRACE close to those of the projection method.
The closeness between the results given by online TRACE and TRACE will be empirically shown later.

\subsubsection{HDPCA Projection}

As online TRACE reduces the computational complexity of TRACE,
we have HDPCA projection that improves the accuracy of projection without increasing the runtimes.
HDPCA reads the PC scores and variances calculated by projection and estimate the shrinkage factor to restore the unshrinked scores. 
The algorithm is
\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
  ($\bO[n^3]$.)
\item Calculate $W = (Y^T X) V D^{-1}$, where $Y$ is the data matrix for all the testing individuals. ($\bO[mpn + mn^2]$.)
  \item Input $D$ and $W$ to the pc\_adjust function in the HDPCA function to obtain the PC scores adjusted for the shrinking bias. ($\bO[p]$ for finding the number of distant spikes (function select.nspike), $\bO[n]$ for adjusting for the shrinking bias (function hdpc\_est))
\end{enumerate}

Total computational complexity: $\bO[n^2p + mn(p + n) + p + n] \approx \bO[n^2p + mn(p+n)]$,
given that $m$ or $n$ is large.

The computational complexities between projection and HDPCA projection are the same.
The accuracy of HDPCA projection will be empirically shown later.

\subsubsection{Computational Complexity Summary}
A comparison of computatinal complexities of the four PCA methods are shown in \Cref{tbl:cplx}.

\begin{table} 
  \centering
  \begin{tabular}{|l|l|}
    \hline
    Method & Computational Complexity \\ 
    \hline
    TRACE & $\bO[n^2 p + mn(p + n^2)]$ \\
    \hline
    Projection & $\bO[n^2p + mk_1p)]$ \\
    \hline
    Online TRACE & $\bO[n^2 p + mk_3(p + k_3 n)]$ \\
    \hline
    HDPCA Projection & $\bO[n^2p + mn(p+n)]$ \\
    \hline
  \end{tabular}
  \caption{
    Comparison of computational complexity for the three methods.
    Here $p$ is the number of SNPs,
    $n$ is the number of individuals in the reference group,
    and $m$ is the number of individuals in the  testing group.
    For projection, $k_1$ is the number of PCs needed
    For Online TRACE, $k_3$ is the number of PCs calculated by online SVD, which is usually much less than $n$.}
  \label{tbl:cplx}
\end{table}

\subsection{Data}

In this study, we used both simulated data and real-world data.

\subsubsection{Simulation Studies}

We used the GGS software to simulate individuals migrating on a grid.
The parameters we chose to use are as follows:
\begin{enumerate}
\item L: Number of loci in each genealogy. (Fixed to $1000$.)
\item G: Number of genealogies. (Equal to $p / L$)
\item K: Side length of the grid. (Fixed to $2$.)
\item c: Number of haploids at each cell of the grid. (Equal to $2n / K^2$ and
$2m / K^2$)
\item M: Migration rate. (Varies)
\end{enumerate}

\subsubsection{Real-World Datasets}

The real-world datasets used in our study are the HGDP and 1000 Genomes datasets.
They contain the genotype readings of individuals from different ethnic groups over the world.
For better comparison, we select only the SNPs that are included in both datasets,
the number of which is approximately $6 \times 10^5$.
Moreover, we removed individuals who have at least parent also in the dataset.
The samples size is approximately $900$ for HGDP and $2300$ for 1000 Genomes.

\section{Results (Simulation Study)}

We now present the results of our study.

\subsection{Parameter Setting}

\begin{enumerate}
\item $n$ fixed, $m$ changing
\begin{enumerate}
\item $p = 100000$
\item $n = 600$
\item $m$: $\min = 1000$, $\max = 5000$, $\text{increment} = 500$ 
\item $migration = 100$
\end{enumerate}
\item $m$ fixed, $n$ changing
\begin{enumerate}
\item $p = 100000$
\item $n$: $\min = 1000$, $\max = 5000$, $\text{increment} = 500$ 
\item $m = 200$
\item $migration = 100$
\end{enumerate}
\end{enumerate}

\subsection{Computation Time}

\includegraphics[width=0.98\textwidth]{runtimes}

\subsection{Accuracy}

Next, we show the accuracy of each method.
We have two ways to measure the accuracy.
First, we can use the information from the simulation given by GGS to categorize the reference individuals into $K^2$ clusters and find the mean of each cluster,
which will be treated as the center of that cluster.
We then find the Euclidean distance between the study individuals and the centers,
again using the true locations in the grid to categorize the study individuals.

Second, we can use Online TRACE as the standard and find the Euclidean distance between the study individuals' PC scores calculated by Online TRACE and another method.

The parameter setting is the same as in comparing the runtimes.

\includegraphics[width=0.98\textwidth]{err_refcenter.pdf}

\includegraphics[width=0.98\textwidth]{err_online.pdf}

\section{Discussion}

\newpage

\begin{thebibliography}{}

\bibitem[Brand(2002)]{brand} Brand, M. (2002). Incremental singular value decomposition of uncertain data with missing values. {\it Computer Vision--ECCV 2002}, 707-720.

\bibitem[Dey \& Lee(2016)]{dey} Dey, R., \& Lee, S. (2016). Asymptotic properties of Principal Component Analysis and shrinkage-bias adjustment under the Generalized Spiked Population model. {\it arXiv preprint arXiv:1607.08647}.

\bibitem[Jolliffe(2002)]{jolliffe} Jolliffe, I. (2002). {\it Principal component analysis}. John Wiley \& Sons, Ltd.
Chicago

 \bibitem[Wang et al.(2015)]{wang} Wang, C., Zhan, X., Liang, L., Abecasis, G. R., \& Lin, X. (2015). Improved ancestry estimation for both genotyping and sequencing data using projection procrustes analysis and genotype imputation. {\it The American Journal of Human Genetics, 96}(6), 926-937.

\end{thebibliography}

    % \begin{enumerate}
    %     \item $X$ ($p$ by $n$): 
    %         Genotype readings for the reference group.
    %     \item $Y = (y_1 \cdots y_m)$ ($p$ by $m$): 
    %         Genotype readings for the study group.
    % \end{enumerate}

% \begin{enumerate}
% \item HGDP
%   The HGDP dataset contains individuals from different ethnic groups in the world. It includes about $6 \times 10^5$ SNPs and 
%   \begin{enumerate}
%   \item Individuals from different countries in the world.
%   \item We only use the SNPs that are also included in the 1000
%     Genomes dataset.
%   \item $p \approx 6 \times 10^5, n \approx 900$.
%   \end{enumerate}
% \item 1000 Genomes
%   \begin{enumerate}
%   \item Individuals from different countries in the world.
%   \item We only use the SNPs that are also included in the HGDP
%     dataset.
%   \item $p \approx 6 \times 10^5, n \approx 2300$.
%   \end{enumerate}
% \item Data filtering
%   \begin{enumerate}
%   \item We only used the SNPs that are included in both the HGDP and
%     1000 Genomes datasets.
%   \item We have removed the individuals with at least one parent in
%     the dataset.
%   \end{enumerate}
% \end{enumerate}

\end{document}

### 2018-02-07 doc/presentation.Rmd
Add important existing files

---
title: "Computational Methods for Principle Component Analysis"
author: "David (Daiwei) Zhang"
date: "November 10, 2017"
header-includes:
   - \usepackage{amsmath, amssymb, amsfonts}
   - \setbeamertemplate{navigation symbols}{}
output:
  beamer_presentation:
    theme: "AnnArbor"
    colortheme: "dolphin"
    fonttheme: "structurebold"
    fig_caption: false
    incremental: false
---

## Principle Component Analysis

- Ancestral history
    - Often a confounder for the association between genes and diseases
- Principle Component Analysis
    - Estimate and quantify ancestral history 
    - Used to eliminate bias caused by ancestral history
- How PCA works
    - Find the directions of the greatest variance
    - Intuitively, rotate the coordination system to fit the directions of the data
    - PCA in 3D (https://www.joyofdata.de/public/pca-3d/)

## PCA on the HGDP Dataset

- Plot by Chaolong Wang et. al.

![](pcaPlot_wangchaolong.png)

## Singular Value Decomposition
- SVD gives us everything we need for PCA
\[
\underset{p \times n}{X} = \underset{p \times n}{U} \; \underset{n \times n}{D} \; \underset{n \times n}{V^T}
\]
    - $p$: # Features (SNPs)
    - $n$: # Individuals
    - $U$: PCs
    - $D$: Diagonal matrix of standard deviations (descending)
    - $V$: PC scores
- Can also use eigen decomposition
\[
X^T X = VD^2V^T, \quad X X^T = U D^2 U^T
\]

## Prediction: Augment and Decompose (A&D)

- Reference group: $\underset{p \times n}{X}$
- New individual: $\underset{p \times 1}{y}$
- Augment & decompose: $\underset{p \times (n+1)}{\tilde{X}} = (X, y) = \tilde{U}\tilde{D}\tilde{V}^T$
- PC scores for the new individual: The last row of $\underset{(n+1) \times (n+1)}{\tilde{V}}$
- The standard way of doing PCA

## Improving A&D: Procrustes Analysis

- So far we have $X = UDV^T$ and $\tilde{X} = \tilde{U}\tilde{D}\tilde{V}^T$
- Define
\[
  \underset{n \times n}V =
  \begin{bmatrix}
    \underset{n \times k_1}{V_{\text{ref}}} & \text{Trash}
  \end{bmatrix}
  \quad \text{and} \quad
  \underset{(n+1) \times (n+1)}{\tilde{V}} =
  \begin{bmatrix}
    \underset{n \times k_2}{\tilde{V}_\text{ref}} & \text{Trash}\\
    \underset{1 \times k_2}{\tilde{V}_\text{stu}} & \text{Trash}\\
  \end{bmatrix}
\]
    - $k_1$: # PCs we need ($1 \leq k_1 \leq k_2 \ll n$)
- Look for a linear transformation $f(\tilde{V}_\text{ref}) = \tilde{V}_\text{ref}A + b$ that minimizes the "distance" between $f(\tilde{V}_\text{ref})$ and $V_\text{ref}$.
\[
  A, b = \underset{A, b}{\operatorname{arg\;min}}\; d(V_\text{ref}, \tilde{V}_{\text{ref}}A+b)
\]
    - e.g. $d(A,B) = [\sum_{i,j}(A-B)_{ij}^2]^\frac{1}{2}$
- Procrustes-transformed PC scores for the new individual
\[
\underset{1 \times k_1}{\tilde{V}_\text{pro}} = \tilde{V}_\text{stu} A + b
\]

## Prediction: Simple Projection

- A&D + Procrustes is implemented in the software TRACE
- Accurate (golden standard), but slow
- A different way to do PCA: Simple projection
- Once finish $X = UDV^T$ and have $y$:
\[
\underset{1 \times n}{V_\text{stu}} = y^T U D^{-1}
\]
- Fast but has a "shrinking" bias
    - PC scores are less in magnitude compared to TRACE
    - Shrinkage increases as $p/n \to \infty$

## TRACE Improved: Online SVD

- Recall in TRACE
\[
  \underset{n \times n}V =
  \begin{bmatrix}
    \underset{n \times k_1}{V_{\text{ref}}} & \text{Trash}
  \end{bmatrix}
  \quad \text{and} \quad
  \underset{(n+1) \times (n+1)}{\tilde{V}} =
  \begin{bmatrix}
    \underset{n \times k_2}{\tilde{V}_\text{ref}} & \text{Trash}\\
    \underset{1 \times k_2}{\tilde{V}_\text{stu}} & \text{Trash}\\
  \end{bmatrix}
\]
where $1 \leq k_1 \leq k_2 \ll n$.

- Only the first $k_2$ PCs are needed for Procrustes analysis
- Online SVD: Only calculate for the first $k_3$ columns of $V$ ($k_2 \leq k_3 \ll n$)
- Make TRACE much faster

## Online SVD Algorithm

\begin{enumerate}
\item Calculate $U_3 = X V_3 D_3^{-1}$,
  where $V_3$ is the first $k_3$ columns of $V$,
  and $D_3$ is the diagonal matrix of the first $k_3$ diagonal entries of $D$.
\item Calculate 
  \[
    L = U_3^T y \quad \text{and} \quad K = y^T H,
  \]
  where $H$ is the normalized  $y - U_3L$
\item Calculate $Q^T Q$, where
  \[
    Q = 
    \begin{bmatrix}
      D_3 & L \\
      0 & K
    \end{bmatrix}.
  \]
\item Apply eigendecomosition to $Q^T Q$ to get $Q^T Q = \ddot{V}_3 \ddot{D}^2_3 \ddot{V}^T_3$.
\item Calculate
  \[
    \tilde{V}_3 =
    \begin{bmatrix}
      V_3 & 0 \\
      0 & 1
    \end{bmatrix}
    \ddot{V}_3.
  \]
\end{enumerate}

## Simple Projection Improved: HDPCA

- Input
    - Eigenvalues of the reference matrix
    - PC scores predicted by simple projection
- Output
    - Magnitude and angle of the shrinkage
    - Adjusted PC scores
- Computational Complexity
    - Higher than simple projection, lower than TRACE

## Comparison of Computational Complexity

\begin{table} 
  \centering
  \begin{tabular}{|l|l|}
    \hline
    Method & Computational Complexity \\ 
    \hline
    TRACE & $\mathcal{O}[n^2 p + mn(p + n^2)]$ \\
    \hline
    Projection & $\mathcal{O}[n^2p + mkp]$ \\
    \hline
    Online TRACE & $\mathcal{O}[n^2 p + mk(p + kn)]$ \\
    \hline
    HDPCA Projection & $\mathcal{O}[n^2p + mkp]$ \\
    \hline
  \end{tabular}
\end{table}

- $p$: # SNPs
- $n$: # Reference individuals 
- $m$: # Study individuals 
- $k_3$: # PCs calculated in Online SVD

## Simulation Study

- GGS: Simulates populations migrating on a $H \times H$ grid
- Parameters for GGS
    - L: Number of loci in each genealogy. (Fixed to $1000$.)
    - G: Number of genealogies. (Equal to $p / L$)
    - H: Side length of the grid. (Fixed to $2$.)
    - c: Number of haploids at each cell of the grid. (Equal to $2n / K^2$ and $2m / H^2$)
    - M: Migration rate. (Fixed to 100)
- Setting for simulation
    - $p = 100,000$
    - Setting 1
        - $n = 600$
        - $m$: $\min = 1000$, $\max = 5000$, $\text{increment} = 500$ 
    - Setting 2
        - $n$: $\min = 1000$, $\max = 5000$, $\text{increment} = 500$ 
        - $m = 200$

## Results ($n = 600$, $m = 1000$)

![](..//data/cluster/ggsim/ggsim_100000_600_1000_2_1_100.pdf)

## Results ($n = 600$, $m = 3000$)

![](..//data/cluster/ggsim/ggsim_100000_600_3000_2_1_100.pdf)

## Results ($n = 1000$, $m = 200$)

![](..//data/cluster/ggsim/ggsim_100000_1000_200_2_1_100.pdf)

## Results ($n = 3000$, $m = 200$)

![](..//data/cluster/ggsim/ggsim_100000_3000_200_2_1_100.pdf)

## Runtimes

![](runtimes.pdf)

## Accuracy (Distance from Reference Centers)

![](err_refcenter.pdf)

## Accuracy (Distance from Online TRACE)

![](err_online.pdf)

## Conclusion

\begin{table} 
  \centering
  \begin{tabular}{|l|c|c|}
    \hline
    Method & Accuracy & Speed (wrt $n$)\\ 
    \hline
    TRACE & Golden Standard & Low ($\mathcal{O}[n^3]$) \\
    \hline
    Projection & Low (when $n \ll p$) & High ($\mathcal{O}[1]$) \\
    \hline
    Online TRACE & High & High ($\mathcal{O}[n]$) \\
    \hline
    HDPCA Projection & High & High ($\mathcal{O}[1]$) \\ 
    \hline
  \end{tabular}
\end{table}

Note: The computational complexity is for each new study individual and $p$ is assumed to be fixed.

### 2018-02-07 genoPCA.py

@@ -1,0 +1,335 @@
+import numpy as np
+import subprocess as sp
+import os
+from os import path as op
+import math
+
+
+import procdata as pd
+
+# User input
+KK = 10
+L = 100
+M = 0.01
+GGS_DIR = '../data/ggs'
+DATA_DIR = '../data/ggs_out'
+
+# Program setting
+INT_TYPE = np.int8
+INT_FMT = '%d'
+FLOAT_TYPE = np.float32
+AXIS = 1  # rows are SNPs, cols are indvs
+DELIM = '\t'
+
+# For my convenience
+TRACE_DIR = '../data/laser'
+TRACE_VER = 'ori'
+
+# Process the info above
+GGS_DIR = op.realpath(GGS_DIR)
+GGS_EXE = GGS_DIR + '/' + 'ggs'
+TRACE_DIR = op.realpath(TRACE_DIR)
+TRACE_EXE = TRACE_DIR + '/' + 'trace' + '.' + TRACE_VER
+
+
+def ggs(outFile, KK, c, G, L, M, exe=GGS_EXE):
+    outFile = op.realpath(outFile)
+    exe = op.realpath(exe)
+    cmd = [
+        exe,
+        '-o', outFile,
+        '-K', str(KK),
+        '-c', str(c),
+        '-G', str(G),
+        '-L', str(L),
+        '-M', str(M)
+    ]
+    # Use devnull to suppress output
+    devnull = open(os.devnull, 'w')
+    sp.call(cmd, stdout=devnull)
+
+
+def ggs_smart(prefix, p, n, KK=KK, L=L, M=M, data_dir=DATA_DIR):
+    data_dir = op.realpath(data_dir)
+    outFile = data_dir + '/' + prefix + '_' + str(p) + '_' + str(n) + '.ggs'
+    assert(p % L == 0)
+    G = p // L
+    assert(n % (KK * KK) == 0)
+    c = n // (KK * KK) * 2
+    ggs(outFile, KK, c, G, L, M)
+
+
+def ggs2dup(ggsFile, dupFile):
+    ''' Convert the haploid genotype simulated by ggs to a duploid genotype. '''
+    # print('Creating duploid genotype...')
+    nCols = pd.getNCols(ggsFile) - 1  # The first column is the row names
+    assert(nCols % 2 == 0)
+    with open(ggsFile, 'r') as ggsF, open(dupFile, 'w') as dupF:
+        next(ggsF)  # skip the header
+        for i, line in enumerate(ggsF):
+            lst = line.split()
[…]

### 2018-02-07 genoPCA_tester.py

@@ -1,0 +1,420 @@
+# Testing functions in genoPCA
+
+# # Remove all user-defined variables
+# import sys
+# this = sys.modules[__name__]
+# for n in dir():
+#     if n[0] != '_':
+#         delattr(this, n)
+
+from os import path as op
+import matplotlib.pyplot as plt
+import subprocess as sp
+import importlib
+import numpy as np
+import filecmp
+
+import genoPCA as gp
+import procdata as pd
+
+importlib.reload(gp)
+
+PREFIX = 'test'
+P = 1000
+N = 100
+
+NPLUSM = N * 2
+K = 4
+DATA_DIR = '../data/ggs_out'
+INT_TYPE = np.int8
+INT_FMT = '%d'
+FLOAT_TYPE = np.float32
+DELIM = '\t'
+
+TRACE_DIR = '../data/laser'
+TRACE_DIR = op.realpath(TRACE_DIR)
+TRACE_EXE = TRACE_DIR + '/' + 'trace.ori'
+
+# sp.call(['rm', op.realpath(DATA_DIR) + '/test*'])
+
+
+def test_ggs():
+    print('Testing ggs...')
+    K = 2
+    c = 2
+    G = 3
+    L = 4
+    M = 10
+    data_dir = '../data/ggs_out'
+    prefix = 'test'
+    p = 1000
+    n = 100
+    outFile = data_dir + '/' + prefix
+    outFile += '_' + str(p) + '_' + str(n)
+    outFile += '.ggs'
+    gp.ggs(outFile, K, c, G, L, M)
+    print('Passed.')
+
+
+def test_ggs_smart():
+    print('Testing smart ggs...')
+    prefix = 'test'
+    p = 1000
+    n = 100
+    gp.ggs_smart(prefix, p, n)
+    print('Passed.')
+
+
+def test_ggs2dup():
+    print('Testing ggs2dup...')
+    data_dir = '../data/ggs_out'
+    prefix = 'test'
+    p = 1000
+    n = 100
+    prefix = data_dir + '/' + prefix + '_' + \
+        str(p) + '_' + str(n)
+    ggsFile = prefix + '.ggs'
+    dupFile = prefix + '.geno'
+    gp.ggs2dup(ggsFile, dupFile)
+    print('Passed.')
+
+
+def test_ggs2dup_smart():
+    print('Testing ggs2dup_smart')
+    prefix = 'test'
+    p = 1000
+    n = 100
[…]

### 2018-02-07 ggs2geno.py

@@ -1,0 +1,9 @@
+import numpy as np
+import makeDataFile as mdf
+import subprocess
+
+    
+
+
+
+

### 2018-02-07 makeDataFile.py

@@ -1,0 +1,524 @@
+
+# def normArray(x):
+#     y = np.array(x)
+#     n = len(y)
+#     xbar = np.mean(y)
+#     xdev = np.std(y) 
+#     y = y - xbar
+#     if xdev > 0:
+#         y = y / xdev
+#     return(y)
+
+# chromoInfoFile = '../data/HGDP_Map'
+# chromoInfoFileCol = 1
+# chromoInfo = getChromoInfo(chromoInfoFile, chromoInfoFileCol)
+
+
+# chromoNum = "21"
+# subsetSNPsFile = snpsFile + "_chromo" + chromoNum
+# getSNPsOnChromo(chromoNum, snpsFile, chromoInfo, subsetSNPsFile)
+
+# popFile = "../data/hgdp_population"
+# popFileCol = 3
+# testIdx, trainIdx = getUnqPops(popFile, popFileCol)
+
+
+
+# pIdx = np.random.choice(pAll, p, replace=False) 
+# nIdx = np.random.choice(nAll, n, replace=False)
+
+# idxConvertFile = "../data/idxConvert"
+# toIdx2 = np.loadtxt(idxConvertFile, dtype="int")
+# toIdx2 = toIdx2 - 1 # start counting by zero
+# nIdx2 = toIdx2[nIdx]
+
+# def getTrainTest(allFile, trainIdx, testIdx, trainFile, testFile):
+
+#     print("Dividing matrix into training and testing matrices by column...")
+#     # clear the output file
+#     open(trainFile, 'w').close()
+#     open(testFile, 'w').close()
+
+#     counter = 0
+#     with open(allFile, "r") as all, open(trainFile, "ab") as train, open(testFile, "ab") as test:
+#         for line in all:
+#             x = strToArray(line)
+#             x = np.array(x)
+#             xTest = x[testIdx]
+#             xTest = np.matrix(xTest)
+#             xTrain = x[trainIdx]
+#             xTrain = np.matrix(xTrain)
+#             np.savetxt(train, xTrain, fmt='%10.5f')
+#             np.savetxt(test, xTest, fmt='%10.5f')
+
+#             counter += 1
+#             if counter % 10000 == 0:
+#                 print("Finished processing line " + str(counter))
+
+#     print("Done!")
+
+
+
+# def getTrainTestLargeMemUse(allFile, trainIdx, testIdx, trainFile, testFile):
+
+#     print("Dividing matrix into training and testing matrices...")
+
+#     # read the SNP file(p * n) for individuals
[…]

### 2018-02-07 pcSummary.R

@@ -1,0 +1,464 @@
+rm(list=ls())
+
+plot.m <- function(pref.std, pref.rand, items){
+  idxs <- c(items.idxs[, items])
+  ## idxs <- c(12:15)
+  p <- 1e5
+  n <- 600
+  ver.std <- "comb"
+  # Y.std <- X[X$pref==pref.std & X$ver==ver.std & X$n==n & X$m >= 1000 & X$m <= 3000,]
+  Y.std <- X[X$pref==pref.std,]
+  Y.std <- Y.std[order(Y.std$m),]
+  ver.rand <- "rand"
+  # Y.rand <- X[X$pref==pref.rand & X$ver==ver.rand & X$n==n & X$m >= 1000 & X$m <= 3000,]
+  Y.rand <- X[X$pref==pref.rand,]
+  Y.rand <- Y.rand[order(Y.rand$m),]
+  main <- paste0(items, ", p=", toString(p), ", n=", toString(n), ", mig=100")
+  pch = c(rep(1,4), rep(2,4))
+  lty = 1
+  col = c(1:4, 1:4)
+  cex = 1.5
+  ylab = items
+  Y.cbind <- cbind(Y.std[, idxs], Y.rand[, idxs])
+  matplot(Y.std$m, Y.cbind, type="b", lty=lty, pch=pch, col=col, xlab="m", ylab=ylab, cex=cex, main=main)
+  legend("topleft",
+        legend = methods,
+          pch = pch,
+          col = col,
+        cex = cex,
+  )
+}
+
+plot.n <- function(pref.std, pref.rand, items){
+  idxs <- c(items.idxs[, items])
+  ## idxs <- c(12:15)
+  p <- 1e5
+  m <- 200
+  ver.std <- "comb"
+  # Y.std <- X[X$pref==pref.std & X$ver==ver.std & X$n==n & X$m >= 1000 & X$m <= 3000,]
+  Y.std <- X[X$pref==pref.std,]
+  Y.std <- Y.std[order(Y.std$n),]
+  ver.rand <- "rand"
+  # Y.rand <- X[X$pref==pref.rand & X$ver==ver.rand & X$n==n & X$m >= 1000 & X$m <= 3000,]
+  Y.rand <- X[X$pref==pref.rand,]
+  Y.rand <- Y.rand[order(Y.rand$n),]
+  main <- paste0(items, ", p=", toString(p), ", m=", toString(m), ", mig=100")
[…]

### 2018-02-07 plot.R

@@ -1,0 +1,348 @@
+rm(list=ls())
+
+## Get the run date for TRACE
+## Output in number of seconds after 1970 UTC (POSIXct)
+get.tracedate <- function(trace.log){
+    trace.log.lineskip <- 5
+    trace.log.charskip <- 17
+    # tzone <- "UTC"
+    tformat <- "%b %d %H:%M:%S %Y"
+    con <- file(trace.log, open="r")
+    readLines(con, n=trace.log.lineskip)
+    ourLine <- readLines(con, n=1)
+    # print(ourLine)
+    close(con)
+    rundate <- substring(ourLine, first=trace.log.charskip)
+    rundate <- as.POSIXct(rundate, format=tformat) 
+    # attr(rundate, "tzone") <- tzone
+    # rundate <- as.numeric(rundate)
+    rundate <- strftime(rundate, format = "%Y-%m-%d_%H:%M:%S")
+    return(rundate)
+}
+
+# read testing pcs from trace's output file
+load.trace <- function(fin) {
+    if(!file.exists(fin)){
+         return(NaN)
+    }
+    result <- read.table(fin, header = TRUE)
+    return(result)
+}
+# one way to define the norm of a matrix
+mat.norm <- function(x){
+    m <- nrow(x)
+    n <- ncol(x)
+    # depends on which verison you want
+    norm <- sqrt(sum(x^2) / (m * n))
+    # norm <- sqrt(sum(x^2))
+    return(norm)
+}
+    
+# the difference between two matrices
+mat.err <- function(x, y){
+    mat.norm(x - y)
+}
+
+# Create a named list of the properties of the methods
+# Each element is a list of the same length
+pcprops <- list()
+
+# Input the display names
+display.names <- c(
+    "ref_trace",
+    "test_trace",
+    "test_onl",
+    "test_proj",
+    "test_hdpca"
+)
+pcprops$display.names <- display.names
+meths.n <- length(display.names)
+
+# Read arguements from user
+args <- commandArgs(trailingOnly = TRUE)
+if(identical(args, character(0))){
+    print("Using testing args...")
+    args <- c("rand", "ggsim", "1000", "300", "100", '2', '1', '100')
+    print(args)
+}
+ver <- args[1]
+pref <- args[2]
+pref <- paste(pref, pref, sep = "/")
+p <- args[3]
+n <- args[4]
+m <- args[5]
+gridsize <- args[6]
[…]

### 2018-02-07 plot_simp.R

@@ -1,0 +1,352 @@
+rm(list=ls())
+
+require(graphics)
+
+## Get the run date for TRACE
+## Output in number of seconds after 1970 UTC (POSIXct)
+get.tracedate <- function(trace.log){
+    trace.log.lineskip <- 5
+    trace.log.charskip <- 17
+    tzone <- "UTC"
+    tformat <- "%b %d %H:%M:%S %Y"
+    con <- file(trace.log, open="r")
+    readLines(con, n=trace.log.lineskip)
+    ourLine <- readLines(con, n=1)
+    print(ourLine)
+    close(con)
+    rundate <- substring(ourLine, first=trace.log.charskip)
+    rundate <- as.POSIXct(rundate, format=tformat) 
+    attr(rundate, "tzone") <- tzone
+    rundate <- as.numeric(rundate)
+    return(rundate)
+}
+
+# read testing pcs from trace's output file
+load.trace <- function(fin) {
+    if(!file.exists(fin)){
+         return(NaN)
+    }
+    result <- read.table(fin, header = TRUE)
+    return(result)
+}
+# one way to define the norm of a matrix
+mat.norm <- function(x){
+    m <- nrow(x)
+    n <- ncol(x)
+    # depends on which verison you want
+    norm <- sqrt(sum(x^2) / (m * n))
+    # norm <- sqrt(sum(x^2))
+    return(norm)
+}
+    
+# the difference between two matrices
+mat.err <- function(x, y){
+    mat.norm(x - y)
+}
+
+# Create a named list of the properties of the methods
+# Each element is a list of the same length
+pcprops <- list()
+
+# Input the display names
+display.names <- c(
+    "ref_trace",
+    "test_trace",
+    "test_onl",
+    "test_proj",
+    "test_hdpca"
+)
+pcprops$display.names <- display.names
+meths.n <- length(display.names)
+
+# Read arguements from user
+args <- commandArgs(trailingOnly = TRUE)
+if(identical(args, character(0))){
+    print("Using testing args...")
+    ## args <- c("rand", "ggsim", "ggsim_1000_200_2_1_100_0_ggsim_1000_100_2_1_100_1", "1000 ", "200", "100")
+    args <- c("comb", "ukb", "kgn_allChr_ukb_orphans_ukb_1k", "145282", "2492", "1000")
+    print(args)
+}
[…]

### 2018-02-07 procdata.py

@@ -1,0 +1,317 @@
+import numpy as np
+import subprocess
+import math
+
+DELIM = '\t'
+P = 1000
+Q = 50000
+
+
+def getNCols(fileName):
+    ''' find the number of columns in a file '''
+    with open(fileName, 'r') as text:
+        line = text.readline()
+        n = len(str(line).split())
+    return(n)
+
+
+def getRows(inFile, rowIdxs, outFile, otherFile):
+    '''Get rows from inFile by rowIdxs as index and save in outFile'''
+    print("Getting rows from index...")
+    with open(inFile, 'r') as inf, open(outFile, 'w') as outf, open(otherFile, 'w') as othf:
+        for i, line in enumerate(inf):
+            if (i in rowIdxs):
+                outf.write(line)
+            else:
+                othf.write(line)
+            if i % 100000 == 0:
+                print('Finished line ' + str(i))
+    print("Done!")
+
+
+def getCols(inFile, idxs, outFile, otherFile):
+    with open(inFile, 'r') as inf, open(outFile, 'w') as outf, open(otherFile, 'w') as othf:
+        for i, line in enumerate(inf):
+            lst = str(line).split()
+            n = len(lst)
+            outlst = [lst[i] for i in idxs]
+            othlst = [lst[i] for i in range(n) if i not in idxs]
+            outst = '\t'.join(outlst) + '\n'
+            othst = '\t'.join(othlst) + '\n'
+            outf.write(outst)
+            othf.write(othst)
+
+
+def getRandRows(inFile, n, nn, outFile, otherFile):
+    print("Getting " + str(n) + " random rows out of " + str(nn) + " rows ...")
+    rowIdxs = np.random.choice(nn, size=n, replace=False)
+    getRows(inFile, rowIdxs, outFile, otherFile)
+
+
+def getEvenRows(inFile, nn, outFile, otherFile):
+    print("Getting every other row of " + str(nn) + " rows ...")
+    n = nn / 2
+    rowIdxs = list(range(n))
+    rowIdxs = [i * 2 for i in rowIdxs]
+    getRows(inFile, rowIdxs, outFile, otherFile)
+
+
+def getEqSpacedRows(inFile, nplusm, n, outFile, otherFile):
+    print('Getting equally spaced ' + str(n) +
+          ' rows out of ' + str(nplusm) + 'rows ...')
+    k = nplusm // n
+    rowIdxs = list(range(n))
+    rowIdxs = [i * k for i in rowIdxs]
+    print(rowIdxs)
[…]

### 2018-02-07 runHdpca.R

@@ -1,0 +1,51 @@
+# Run the hdpca package's function to predict the PC scores
+
+require(hdpca)
+require(MASS)
+
+print("Running hdpca...")
+args <- commandArgs(trailingOnly = TRUE)
+if(identical(args, character(0))){
+    print("Using testing args...")
+	args <- c("comb", "ggsim", "1000", "100", "100", "2", "1", "10")
+}
+
+ver <- args[1]
+pref <- args[2]
+pref <- paste(pref, pref, sep = "/")
+p <- args[3]
+n <- args[4]
+m <- args[5]
+k <- args[6]
+s <- args[7]
+mig <- args[8]
+name <- paste(pref, p, n, m, k, s, mig, sep = "_")
+name <- paste0("../data/", name, ".", ver)
+
+vproj.file <- paste0(name, ".hdpca.vproj")
+vproj.df <- read.table(vproj.file, header = TRUE)
+vproj <- as.matrix(vproj.df[,-(1:5)])
+
+d.file <- paste0(name, ".hdpca.d")
+d <- c(as.matrix(read.table(d.file)))
+
+timing <- (system.time({
+	print("Running HDPCA (pc_adjust)...")
+	vpred <- pc_adjust(d^2, as.numeric(p), as.numeric(n), vproj, method = "d.gsp", n.spikes.max = 20)
+}))
+timing.idx <- 3
+hdpca.timing <- unname(timing[timing.idx])
+
+runtimes.file <- paste0(name, ".runtimes")
+runtimes <- read.table(runtimes.file, header=TRUE)
+runtimes$elapse[runtimes$method=="test_hdpca"] <- runtimes$elapse[runtimes$method=="test_hdpca"] + hdpca.timing
+write.table(runtimes, runtimes.file, sep = "\t", quote=FALSE, row.names=FALSE)
+
+popID <- vproj.df$popID
+indivID <- vproj.df$indivID
+vpred.df <- data.frame(popID, indivID, L=NaN, K=NaN, t=NaN, vpred)
+pred.file <- paste0(name, ".hdpca.vpred")
+write.matrix(vpred, pred.file, sep = "\t")
+write.table(vpred.df, pred.file, sep="\t", quote=FALSE, row.names=FALSE)
+print("Done!")
+

### 2018-02-07 runHdpca_simp.R

@@ -1,0 +1,45 @@
+# Run the hdpca package's function to predict the PC scores
+
+require(hdpca)
+require(MASS)
+
+print("Running hdpca...")
+args <- commandArgs(trailingOnly = TRUE)
+if(identical(args, character(0))){
+    print("Using testing args...")
+	args <- c("comb", "ggsim", "1000", "100", "100", "2", "1", "10")
+}
+
+ver <- args[1]
+direc <- args[2] 
+name <- args[3]
+p <- args[4]
+n <- args[5]
+name <- paste0("../data/", direc, "/", name)
+
+vproj.file <- paste0(name, ".", ver, ".hdpca.vproj")
+vproj.df <- read.table(vproj.file, header = TRUE)
+vproj <- as.matrix(vproj.df[,-(1:5)])
+
+d.file <- paste0(name, ".", ver, ".hdpca.d")
+d <- c(as.matrix(read.table(d.file)))
+timing <- (system.time({
+	print("Running HDPCA (pc_adjust)...")
+	vpred <- pc_adjust(d^2, as.numeric(p), as.numeric(n), vproj, method = "d.gsp", n.spikes.max = 20)
+}))
+timing.idx <- 3
+hdpca.timing <- unname(timing[timing.idx])
+
+runtimes.file <- paste0(name, ".", ver, ".runtimes")
+runtimes <- read.table(runtimes.file, header=TRUE)
+runtimes$elapse[runtimes$method=="test_hdpca"] <- runtimes$elapse[runtimes$method=="test_hdpca"] + hdpca.timing
+write.table(runtimes, runtimes.file, sep = "\t", quote=FALSE, row.names=FALSE)
+
+popID <- vproj.df$popID
+indivID <- vproj.df$indivID
+vpred.df <- data.frame(popID, indivID, L=NaN, K=NaN, t=NaN, vpred)
+pred.file <- paste0(name, ".", ver, ".hdpca.vpred")
+write.matrix(vpred, pred.file, sep = "\t")
+write.table(vpred.df, pred.file, sep="\t", quote=FALSE, row.names=FALSE)
+print("Done!")
+

### 2018-02-07 run_getHalfRows.py

@@ -1,0 +1,11 @@
+import procdata as pd
+import sys
+
+inFile = sys.argv[1]
+n = int(sys.argv[2])
+nplusm = int(sys.argv[3])
+outFile = sys.argv[4]
+otherFile = sys.argv[5]
+# pd.getEqSpacedRows(inFile, nplusm, n, outFile, otherFile)
+# pd.getEvenRows(inFile, nplusm, outFile, otherFile)
+pd.getRandRows(inFile, n, nplusm, outFile, otherFile)

### 2018-02-07 run_getWeightedCols.py

@@ -1,0 +1,10 @@
+import procdata as pd
+import numpy as np
+import sys
+
+inFile = sys.argv[1]
+outFile = sys.argv[2]
+othFile = sys.argv[3]
+s = int(sys.argv[4])
+weight = np.array([1.0, 1.0 / s, 1.0 / s, 1.0 / s])
+pd.getWeightedCols(inFile, outFile, othFile, weight)

### 2018-02-07 run_ggs2trace.py

@@ -1,0 +1,28 @@
+import procdata as pd
+import sys
+
+pref = sys.argv[1]
+p = int(sys.argv[2])
+n = int(sys.argv[3])
+k = int(sys.argv[4])
+s = int(sys.argv[5])
+mig = int(sys.argv[6])
+
+# print('Using testing parameters for ggs2trace')
+# pref = '../data/ggsim/ggsim'
+# p = 1000
+# n = 100 + 100
+# s = 1
+# h = 2 * 2
+
+h = k * k
+n_reduced = n // h + n // (h * s) * (h - 1)
+ggsFile = pref + '_' + str(p) + '_' + str(n) + '_' + \
+    str(k) + '_' + str(mig) + '.ggs'
+genoFile = pref + '_' + str(p) + '_' + str(n_reduced) + \
+    '_' + str(k) + '_' + str(s) + '_' + str(mig) + '.geno'
+weight = [1.0] + [1.0 / s] * (h - 1)
+# Something went wrong with weight after transpose_block is used
+# Temporarily disabled weight option
+# pd.ggs2trace(ggsFile, genoFile, weight)
+pd.ggs2trace(ggsFile, genoFile)

### 2018-02-07 run_ggsChild2geno.py

@@ -1,0 +1,11 @@
+import procdata as pd
+import sys
+
+fp = sys.argv[1]
+n = int(sys.argv[2])
+m = int(sys.argv[3])
+nChild = m // (n // 2)
+ggsFile = fp + "_" + str(n) + ".ggs"
+genoFile = fp + "_" + str(m) + ".geno"
+pd.ggsChild2geno(ggsFile, genoFile, nChild)
+

### 2018-02-07 trace.comb.cpp

@@ -1,0 +1,2476 @@
+/*  
+	TRACE: fasT and Robust Ancestry Coordinate Estimation
+    Copyright (C) 2013-2016  Chaolong Wang
+
+    This program is free software: you can redistribute it and/or modify
+    it under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    This program is distributed in the hope that it will be useful,
+    but WITHOUT ANY WARRANTY; without even the implied warranty of
+    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+    GNU General Public License for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with this program.  If not, see <http://www.gnu.org/licenses/>.
+*/
+
+#include <iostream>
+#include <iomanip>
+#include <fstream>
+#include <string>
+#include <map>
+#include <vector>
+#include <gsl/gsl_rng.h>
+#include <gsl/gsl_randist.h>
+#include <gsl/gsl_statistics_double.h>
+#include <cmath>
+#include <algorithm>
+#include <time.h>
+#include <stdio.h>
+#include <stdlib.h>
+#define  ARMA_DONT_USE_WRAPPER
+#include "armadillo"
+#include <list>
+using namespace arma;
+using namespace std;
+extern "C" void openblas_set_num_threads(int num_threads);
+
+
+// Print a variable with its name
+#define printer(name) PRINTER(#name, (name))
+void PRINTER(string name, mat X) {cout << endl << name << endl << X << endl;}
+void PRINTER(string name, fmat X) {cout << endl << name << endl << X << endl;}
+void PRINTER(string name, int X) {cout << endl << name << endl << X << endl;}
+void PRINTER(string name, float X) {cout << endl << name << endl << X << endl;}
+void PRINTER(string name, double X) {cout << endl << name << endl << X << endl;}
+
+const string ARG_PARAM_FILE = "-p";
+const string ARG_STUDY_FILE = "-s";
+const string ARG_GENO_FILE = "-g";
+const string ARG_COORD_FILE = "-c";
+const string ARG_OUT_PREFIX = "-o";
+const string ARG_DIM = "-k";
[…]

### 2018-02-07 trace.notr.cpp

@@ -1,0 +1,2476 @@
+/*  
+	TRACE: fasT and Robust Ancestry Coordinate Estimation
+    Copyright (C) 2013-2016  Chaolong Wang
+
+    This program is free software: you can redistribute it and/or modify
+    it under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    This program is distributed in the hope that it will be useful,
+    but WITHOUT ANY WARRANTY; without even the implied warranty of
+    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+    GNU General Public License for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with this program.  If not, see <http://www.gnu.org/licenses/>.
+*/
+
+#include <iostream>
+#include <iomanip>
+#include <fstream>
+#include <string>
+#include <map>
+#include <vector>
+#include <gsl/gsl_rng.h>
+#include <gsl/gsl_randist.h>
+#include <gsl/gsl_statistics_double.h>
+#include <cmath>
+#include <algorithm>
+#include <time.h>
+#include <stdio.h>
+#include <stdlib.h>
+#define  ARMA_DONT_USE_WRAPPER
+#include "armadillo"
+#include <list>
+using namespace arma;
+using namespace std;
+extern "C" void openblas_set_num_threads(int num_threads);
+
+
+// Print a variable with its name
+#define printer(name) PRINTER(#name, (name))
+void PRINTER(string name, mat X) {cout << endl << name << endl << X << endl;}
+void PRINTER(string name, fmat X) {cout << endl << name << endl << X << endl;}
+void PRINTER(string name, int X) {cout << endl << name << endl << X << endl;}
+void PRINTER(string name, float X) {cout << endl << name << endl << X << endl;}
+void PRINTER(string name, double X) {cout << endl << name << endl << X << endl;}
+
+const string ARG_PARAM_FILE = "-p";
+const string ARG_STUDY_FILE = "-s";
+const string ARG_GENO_FILE = "-g";
+const string ARG_COORD_FILE = "-c";
+const string ARG_OUT_PREFIX = "-o";
+const string ARG_DIM = "-k";
[…]

### 2018-02-07 trace.ori.cpp

@@ -1,0 +1,1986 @@
+/*  
+	TRACE: fasT and Robust Ancestry Coordinate Estimation
+    Copyright (C) 2013-2016  Chaolong Wang
+
+    This program is free software: you can redistribute it and/or modify
+    it under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    This program is distributed in the hope that it will be useful,
+    but WITHOUT ANY WARRANTY; without even the implied warranty of
+    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+    GNU General Public License for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with this program.  If not, see <http://www.gnu.org/licenses/>.
+*/
+
+#include <iostream>
+#include <iomanip>
+#include <fstream>
+#include <string>
+#include <map>
+#include <vector>
+#include <gsl/gsl_rng.h>
+#include <gsl/gsl_randist.h>
+#include <gsl/gsl_statistics_double.h>
+#include <cmath>
+#include <algorithm>
+#include <time.h>
+#include <stdio.h>
+#include <stdlib.h>
+#define  ARMA_DONT_USE_WRAPPER
+#include "armadillo"
+using namespace arma;
+using namespace std;
+extern "C" void openblas_set_num_threads(int num_threads);
+
+const string ARG_PARAM_FILE = "-p";
+const string ARG_STUDY_FILE = "-s";
+const string ARG_GENO_FILE = "-g";
+const string ARG_COORD_FILE = "-c";
+const string ARG_OUT_PREFIX = "-o";
+const string ARG_DIM = "-k";
+const string ARG_DIM_HIGH = "-K";
+const string ARG_MIN_LOCI = "-l";
+const string ARG_REF_SIZE = "-N";
+const string ARG_FIRST_IND = "-x";
+const string ARG_LAST_IND = "-y";
+const string ARG_ALPHA = "-a";
+const string ARG_THRESHOLD = "-t";
+const string ARG_MASK_PROP = "-m";
+const string ARG_TRIM_PROP = "-M";
+const string ARG_EXCLUDE_LIST = "-ex";
+const string ARG_PROCRUSTES_SCALE = "-rho";
+const string ARG_RANDOM_SEED= "-seed";
+//const string ARG_NUM_THREADS = "-nt";
+
[…]

### 2018-02-07 trace.rand.cpp

@@ -1,0 +1,2481 @@
+/*  
+	TRACE: fasT and Robust Ancestry Coordinate Estimation
+    Copyright (C) 2013-2016  Chaolong Wang
+
+    This program is free software: you can redistribute it and/or modify
+    it under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    This program is distributed in the hope that it will be useful,
+    but WITHOUT ANY WARRANTY; without even the implied warranty of
+    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+    GNU General Public License for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with this program.  If not, see <http://www.gnu.org/licenses/>.
+*/
+
+#include <iostream>
+#include <iomanip>
+#include <fstream>
+#include <string>
+#include <map>
+#include <vector>
+#include <gsl/gsl_rng.h>
+#include <gsl/gsl_randist.h>
+#include <gsl/gsl_statistics_double.h>
+#include <cmath>
+#include <algorithm>
+#include <time.h>
+#include <stdio.h>
+#include <stdlib.h>
+#define  ARMA_DONT_USE_WRAPPER
+#include "armadillo"
+#include <list>
+using namespace arma;
+using namespace std;
+extern "C" void openblas_set_num_threads(int num_threads);
+
+
+// Print a variable with its name
+#define printer(name) PRINTER(#name, (name))
+void PRINTER(string name, mat X) {cout << endl << name << endl << X << endl;}
+void PRINTER(string name, fmat X) {cout << endl << name << endl << X << endl;}
+void PRINTER(string name, int X) {cout << endl << name << endl << X << endl;}
+void PRINTER(string name, float X) {cout << endl << name << endl << X << endl;}
+void PRINTER(string name, double X) {cout << endl << name << endl << X << endl;}
+
+const string ARG_PARAM_FILE = "-p";
+const string ARG_STUDY_FILE = "-s";
+const string ARG_GENO_FILE = "-g";
+const string ARG_COORD_FILE = "-c";
+const string ARG_OUT_PREFIX = "-o";
+const string ARG_DIM = "-k";
[…]

### 2018-02-07 trace.rand.cpp
Change randSvd dim to DIM_SUPER in trace.rand.cpp. Change python to python3 in ggs2trace. Change runTrace to run TRACe here.

@@ -46,6 +46,17 @@ void PRINTER(string name, int X) {cout << endl << name << endl << X << endl;}
 void PRINTER(string name, float X) {cout << endl << name << endl << X << endl;}
 void PRINTER(string name, double X) {cout << endl << name << endl << X << endl;}
 
+template<class Matrix>
+void pm(Matrix matrix) {
+  cout << endl << size(matrix) << endl;
+  Matrix submat = matrix.submat(0,0,4,4);
+  submat.print(std::cout);
+  cout << endl;
+}
+template void pm<arma::mat>(arma::mat matrix);
+template void pm<arma::fmat>(arma::fmat matrix);
+
+
 const string ARG_PARAM_FILE = "-p";
 const string ARG_STUDY_FILE = "-s";
 const string ARG_GENO_FILE = "-g";
@@ -805,7 +816,7 @@ int main(int argc, char* argv[]){
 		cout << "Performing PCA on reference individuals (using rand)..." << endl;
 		foutLog << endl << asctime (timeinfo);
 		foutLog << "Performing PCA on reference individuals (using rand)..." << endl;
-		randSvd(RefD_raw, V_rand_flt, d_rand_flt, k = DIM);
+		randSvd(RefD_raw, V_rand_flt, d_rand_flt, k = DIM_SUPER);
 		// V.save("/home/david/research/ref_V", raw_ascii);
 		// V_rand_flt.save("/home/david/research/ref_V_rand", raw_ascii);
 		// printer(V.submat(0, 0, 4, 4));

### 2018-02-07 trace.rand.cpp
Change randSvd dim to REF_SIZE

@@ -816,7 +816,7 @@ int main(int argc, char* argv[]){
 		cout << "Performing PCA on reference individuals (using rand)..." << endl;
 		foutLog << endl << asctime (timeinfo);
 		foutLog << "Performing PCA on reference individuals (using rand)..." << endl;
-		randSvd(RefD_raw, V_rand_flt, d_rand_flt, k = DIM_SUPER);
+		randSvd(RefD_raw, V_rand_flt, d_rand_flt, k = REF_SIZE);
 		// V.save("/home/david/research/ref_V", raw_ascii);
 		// V_rand_flt.save("/home/david/research/ref_V_rand", raw_ascii);
 		// printer(V.submat(0, 0, 4, 4));
@@ -2462,8 +2462,7 @@ int randSvd(const fmat &XX, fmat &U, fvec &d, unsigned k, unsigned nIter){
   // cout << "X" << accu(X) << endl << X.submat(0,0,3,3) << endl;
   standardize_rsvd(X);
   // cout << "X after standarization" << accu(X) << endl << X.submat(0,0,3,3) << endl;
-  unsigned l = k;
-  fmat R(p, k + l, fill::randn);
+  fmat R(p, k, fill::randn);
   // cout << "R" << accu(R) << endl << R.submat(0,0,3,3) << endl;
   // cout << "X before multiplication" << accu(X) << endl << X.submat(0,0,3,3) << endl;
   fmat Y = X * R;

### 2018-02-08 procdata.py
More precise timing for trace and eliminate trash files from transpose_block.
makefile: Added comb inside each if branch
procdata: Auto delete the trash middle files for block transposing.
trace.comb.cpp: More precise timing for online procrustes
trace.rand.cpp: More precise timing for online procrustes. Disable
eigen decomposition on study individuals and use random numbers as place
holders to save time.

@@ -286,6 +286,8 @@ def transpose_block(inFile, outFile, p, q):
             qq = N % q
         concat_col(outRowBlocks, outRow, qq, M, m)
     concat(outRows, outFile)
+    subprocess.call(["rm", inFile + "_*"])
+    subprocess.call(["rm", outFile + "_*"])
 
 
 def concat(inFiles, outFile):

### 2018-02-08 trace.comb.cpp

@@ -995,6 +995,7 @@ int main(int argc, char* argv[]){
         double augEigen_time = 0.0;
         double procrust_time = 0.0;
         double onlsvd_time = 0.0;
+        double onlprocrust_time = 0.0;
         double randsvd_time = 0.0;
         double proj_time = 0.0;
         mat V_proj(LAST_IND, REF_SIZE); 
@@ -1177,6 +1178,7 @@ int main(int argc, char* argv[]){
                                 onlsvd_time += time_entry;
 
 				//=================  Procrustes Analysis (Online) =======================
+                                t2 = clock();
 				refPC_rot.clear();
 				t = 0;
 				rho = 0;
@@ -1191,6 +1193,8 @@ int main(int argc, char* argv[]){
 				refPC_new.clear();
 				refPC_rot.clear();
 				rotPC_one = rho*PC_one*A+b;
+        time_entry = (clock() - t2) * 1.0 / CLOCKS_PER_SEC;
+        onlprocrust_time += time_entry;
 
 				//================= Output Procrustes Results (Online) ===================				
 				fout2 << Info1 << "\t" << Info2 << "\t" << (LOCI-Lm) << "\t" << DIM_HIGH << "\t" << t << "\t";	
@@ -1282,6 +1286,9 @@ int main(int argc, char* argv[]){
         name_entry = "onlSVD";
         time_entry = onlsvd_time;
         runtimes[name_entry] = time_entry;
+        name_entry = "onlProcrust";
+        time_entry = onlprocrust_time;
+        runtimes[name_entry] = time_entry;
         // name_entry = "randSVD";
         // time_entry = randsvd_time;
         // runtimes[name_entry] = time_entry;
@@ -1331,7 +1338,7 @@ int main(int argc, char* argv[]){
 	runtimes_sum["ref_proj"] = runtimes_sum["ref_trace"];
 	runtimes_sum["ref_hdpca"] = runtimes_sum["ref_proj"];
 	runtimes_sum["test_trace"] = runtimes["augCov"] + runtimes["augEigen"] + runtimes["procrustes"];
-	runtimes_sum["test_onl"] = runtimes["onlSVD"] + runtimes["procrustes"];
+	runtimes_sum["test_onl"] = runtimes["onlSVD"] + runtimes["onlProcrust"];
 	runtimes_sum["test_proj"] = runtimes["proj"];
 	runtimes_sum["test_hdpca"] = runtimes["proj"];
 	cout << "Runtime breakdown (sec): " << endl;

### 2018-02-08 trace.rand.cpp

@@ -792,22 +792,22 @@ int main(int argc, char* argv[]){
                 time_entry = (clock() - t2) * 1.0 / CLOCKS_PER_SEC;
                 runtimes[name_entry] = time_entry;
 	}else{
-		t2 = clock();
-		time ( &rawtime );
-		timeinfo = localtime ( &rawtime );
-		cout << endl << asctime (timeinfo);
-		cout << "Performing PCA on reference individuals ..." << endl;
-		foutLog << endl << asctime (timeinfo);
-		foutLog << "Performing PCA on reference individuals ..." << endl;	
 		rowvec PCvar = zeros<rowvec>(DIM);
 		rowvec PCvar_rand = zeros<rowvec>(DIM);
-		// pca_cov(RefM, DIM, refPC, PCvar);   // Perform PCA
-		// These two functions is a breakdown of pca_cov. It makes d and V available.
-		eig_descend(d, V, RefM);
-		eigDes2pcaCov(d, V, DIM, refPC, PCvar);
-		name_entry = "calc_ref_pca_coord";
-		time_entry = (clock() - t2) * 1.0 / CLOCKS_PER_SEC;
-		runtimes[name_entry] = time_entry;
+		// t2 = clock();
+		// time ( &rawtime );
+		// timeinfo = localtime ( &rawtime );
+		// cout << endl << asctime (timeinfo);
+		// cout << "Performing PCA on reference individuals ..." << endl;
+		// foutLog << endl << asctime (timeinfo);
+		// foutLog << "Performing PCA on reference individuals ..." << endl;	
+		// // pca_cov(RefM, DIM, refPC, PCvar);   // Perform PCA
+		// // These two functions is a breakdown of pca_cov. It makes d and V available.
+		// eig_descend(d, V, RefM);
+		// eigDes2pcaCov(d, V, DIM, refPC, PCvar);
+		// name_entry = "calc_ref_pca_coord";
+		// time_entry = (clock() - t2) * 1.0 / CLOCKS_PER_SEC;
+		// runtimes[name_entry] = time_entry;
 
 		t2 = clock();
 		time ( &rawtime );
@@ -1010,6 +1010,7 @@ int main(int argc, char* argv[]){
         double augEigen_time = 0.0;
         double procrust_time = 0.0;
         double onlsvd_time = 0.0;
+        double onlprocrust_time = 0.0;
         double randsvd_time = 0.0;
         double proj_time = 0.0;
[…]

### 2018-02-08 trace.rand.cpp
Change to the right ref matrix for procrustes in rand

@@ -606,7 +606,6 @@ int main(int argc, char* argv[]){
 	string *RefInfo2 = new string [REF_SIZE];
 	fmat RefD(REF_SIZE, LOCI);
 	mat refPC = zeros<mat>(REF_SIZE,DIM);
-	mat refPC_rand = zeros<mat>(REF_SIZE,DIM);
 	mat V = zeros<mat>(REF_SIZE, REF_SIZE);
 	vec d = zeros<vec>(REF_SIZE);
 	unsigned DIM_SUPER = 2 * DIM_HIGH;
@@ -829,7 +828,7 @@ int main(int argc, char* argv[]){
 		d_rand = conv_to<mat>::from(d_rand_flt);
     V = V_rand;
     d = d_rand;
-		eigDes2pcaCov(d_rand, V_rand, DIM, refPC_rand, PCvar_rand);
+		eigDes2pcaCov(d_rand, V_rand, DIM, refPC, PCvar_rand);
 		name_entry = "calc_ref_pca_coord_rand";
 		time_entry = (clock() - t2) * 1.0 / CLOCKS_PER_SEC;
 		runtimes[name_entry] = time_entry;

### 2018-02-21 accuracy.R
Improve plotting clarity and settings for FLUX

@@ -25,7 +25,8 @@ print("Calculating accuracy...")
 args <- commandArgs(trailingOnly = TRUE)
 if(identical(args, character(0))){
     print("Using testing args.")
-    args <- "../data/tmp/tmp_100000_1000_100_2_1_10.rand"
+    ## args <- "../data/tmp/tmp_100000_1000_100_2_1_10.rand"
+    args <- "../data/ukb/kgn_allChr_ukb_orphans_ukb_1k.rand"
 }
 
 # Info about data file format
@@ -59,6 +60,9 @@ trace.ref.ctr <- aggregate(trace.ref[,ref.pc.idx], by = list(trace.ref$popID), F
 colnames(trace.ref.ctr) <- c("popID", "PC1.ctr", "PC2.ctr")
 trace.test.ctr <- merge(trace.test[,info.idx], trace.ref.ctr, by = "popID")
 ref.ctr <- as.matrix(trace.test.ctr[,-info.idx])
+if(nrow(ref.ctr) == 0){
+  ref.ctr <- matrix(-99, nrow(trace.test), 2)
+}
 
 # Create reference matrix (gold)
 ref.gold <- as.matrix(trace.test[,test.pc.idx])
@@ -104,31 +108,34 @@ print("Done!")
 ## Plot pc scores
 ref.popu <- as.integer(trace.ref$popID)
 popu.n <- length(unique(ref.popu))
-meth.col <- rainbow(meth.n)
+ref.col <- rainbow(popu.n)[ref.popu]
+ref.pch <- 0
+meth.col <- rep("gray40", meth.n)
+meth.pch <- c(1:meth.n)
 pdf(pdf.file, width = 8, height = 8)
 plot(ref.mat[,1], ref.mat[,2],
-     pch = ref.popu+1, col = "gray",
+     pch = ref.pch, col = ref.col,
      xlab = "PC1", ylab = "PC2",
      main = out.pref)
 for(i in 1:meth.n){
-  points(test.mat[[i]][,1], test.mat[[i]][,2], pch = 1, col = meth.col[i])
+  points(test.mat[[i]][,1], test.mat[[i]][,2], pch = meth.pch[i], col = meth.col[i])
 }
 legend("topleft",
        title = "method",
        legend = meth.name,
-       pch = 1,
+       pch = meth.pch,
        col = meth.col
 )
 legend("bottomleft",
        title = "err.center",
        legend = err.ctr,
-       pch = 1,
+       pch = meth.pch,
[…]

### 2018-02-21 chain.sh

@@ -4,13 +4,16 @@ sim=$1
 ver=$2
 
 host=`hostname`
-if [ ${host:0:3} == "flux" ]; then
+if [[ "${host:0:3}" == "nyx" || "${host:0:4}" == "flux" ]]; then
+    echo "Loading modules for flux..."
     module load python-dev 
     module load numpy-dev 
     module load mkl/11.3.3 # Dependency for armadillo
     module load gcc/5.4.0 # Dependency for armadillo
     module load armadillo
     module load R
+    export LD_LIBRARY_PATH=/home/daiweiz/gsl/lib:$LD_LIBRARY_PATH;
+    echo "Done."
 fi
 
 if [ ${sim} == "1" ]; then

### 2018-02-21 ggs2trace.sh

@@ -10,6 +10,6 @@ k=$4
 s=$5
 mig=$6
 
-python3 run_ggs2trace.py ../data/${pref}/${pref} ${p} ${n} ${k} ${s} ${mig}
+python run_ggs2trace.py ../data/${pref}/${pref} ${p} ${n} ${k} ${s} ${mig}
 rm ../data/${pref}/${pref}_${p}_${n}_${k}_${s}_${mig}.geno_*
 rm ../data/${pref}/${pref}_${p}_${n}_${k}_${s}_${mig}.geno.tr_*

### 2018-02-21 simGeno.sh

@@ -27,9 +27,7 @@ if [[ -f ${geno_file} && -f ${study_file} ]]; then
     echo "Using existing data files."
 else
     echo "Generating data with GGS..."
-    if [ ! ../data/${pref}/ ]; then
-        mkdir ../data/${pref}
-    fi
+    mkdir -p ../data/${pref}
     bash runGgs.sh ${pref} ${p} ${nplusmHigh} ${k} ${mig}
     date
     bash ggs2trace.sh ${pref} ${p} ${nplusmHigh} ${k} ${s} ${mig}

### 2018-02-21 chain.sh
Add Python virtual environment for Local

@@ -14,6 +14,11 @@ if [[ "${host:0:3}" == "nyx" || "${host:0:4}" == "flux" ]]; then
     module load R
     export LD_LIBRARY_PATH=/home/daiweiz/gsl/lib:$LD_LIBRARY_PATH;
     echo "Done."
+elif [ ${host} == "david-XPS-13-9343" ]; then
+    echo "Using Python virtualenv for local."
+    source virtualenvwrapper.sh
+    workon research
+    python --version
 fi
 
 if [ ${sim} == "1" ]; then

### 2018-02-21 plot.R

@@ -66,7 +66,7 @@ meths.n <- length(display.names)
 args <- commandArgs(trailingOnly = TRUE)
 if(identical(args, character(0))){
     print("Using testing args...")
-    args <- c("rand", "tmp", "100000", "1000", "200", '2', '1', '100')
+    args <- c("rand", "tmp", "100000", "1000", "100", '2', '1', '10')
     print(args)
 }
 ver <- args[1]

### 2018-02-21 simGeno.sh

@@ -23,7 +23,7 @@ ggs_file=../data/${pref}/${pref}_${p}_${nplusmHigh}_${k}_${mig}.ggs
 geno_both_file=../data/${pref}/${pref}_${p}_${nplusmHigh}_${k}_${s}_${mig}.geno
 dup_file=${geno_both_file}.tr
 
-if [[ -f ${geno_file} && -f ${study_file} ]]; then
+if [[ -s ${geno_file} && -s ${study_file} ]]; then
     echo "Using existing data files."
 else
     echo "Generating data with GGS..."

### 2018-02-21 ggs2geno.py
Move old files to old

@@ -1,9 +1,0 @@
-import numpy as np
-import makeDataFile as mdf
-import subprocess
-
-    
-
-
-
-

### 2018-02-22 doc/paper.tex
Create mergePlotR and add paper.tex

@@
-\documentclass{article}[12pt]
+\documentclass{article}
 
@@
 
-\title{Principle Component Analysis Computational Methods
-  for Estimating Ancestral Histroy from Genotype Data}
+\title{Computational Methods for Principle Component Analysis Applied to
+  Genotype Data}
+\date{}
 
@@
 
+Ancestral history can be a influential confounder for the association between
+genotypes and phenotypes (\cite{price}).
+Thus stratification is needed to eliminate this bias.
+Principle component analysis (PCA) has been proven to be useful in estimating
+ancestral history (\cite{reich}).
+The standard way of doing PCA is to find the principle components (PC) of the
+reference data and project the data points of the study data onto them.
+We call this method simple projection.
+However, when the number of features are greatly larger than the size of the
+reference sample,
+the estimated PC scores by this method are known to be biased and less
+than the populational PC scores.
+One way of solving this issue is presented by \cite{wang}.
+Their solution is to combine one study individual with the reference set and
+find the PC scores of this augmented data set by singular value decomposition (SVD).
+The study individual's PC scores is then retrieved from the PC score matrix and
+mapped to the reference PC's by a Procrustes transformation.
+We call this method ``augment, decompose, and transform (ADT).
+This method has been shown to be useful in eliminating the shrinking bias of
+study PC scores, but its disadvantage is computational inefficiency.
+
+Two methods have been developed to solve this dillema of accuracy verses efficiency.
+The magnitude and angle of shrinkage in simple projection has been calculated by
+\cite{dey} and the result is implemented into the High Dimensional Principle
+Component Analysis (HDPCA) package in R.
+We call this method ``adjusted projection'' (AP).
+The computational complexity of AP is significantly lower than that of ADT.
+Moreover, an online SVD algorithm has been developed to speed up SVD by calculating an approximation of the top few PC scores only.
+This is useful for our purpose since in genetic studies usually only the first few PC scores are needed. 
+We call this method ``online augment-decompose-transform'' (OADT).
+
+In addition, we aim at solving a problem with finding the PC's of the reference set.
+The previously mentioned four methods all need to find the SVD of the reference data first.
+This requires loading the covariance matrix of the reference data into the
+computer's memory. 
+However, when the size of the reference set is large,
+the resulted covariance matrix will far exceed the size of a normal personal
+computer's memory.
+To solve this issue, a randomized SVD method has been developed that does not
+require finding the covariance matrix.
+
+In this paper, we compare the theoretical and empirical properties of simple
+projection, augment-decompose-transform, adjusted projection, and online augment-decompose-transform. In particular, we are
+interested in their computational complexity as well as empirical speed and
+accuracy.
+Secondly, we compare using standard SVD and randomized SVD for the reference set
+and examine the difference in accuracy and runtimes.
+We found that HDPCA projection and OADT have both achieved the accuracy of ADT
+and the efficiency of simple projection.
+In addition, randomized SVD gives results similar to standard SVD and only
+increase the computational time for the reference data.
+
 \section{Method}
@@
 
-\begin{enumerate}
-\item Data matrices:\\
-  The data matrix contains the genotypes for the individuals.
-  Each row represents a SNP, and each column represents an individual.
-  Each entry is $0$, $1$, or $2$.
+\subsubsection{Data Format}
+As mentioned previously, there are two datasets: a reference set and a study
+set.
+They are both represented as matrices.
+A data matrix contains the genotypes for the individuals.
+Each row represents a single-nucleotide polymorphism (SNP),
+while each column represents an individual.
+Each entry of the matrix is $0$, $1$, or $2$,
+representing the number of mutations observed at the SNP.
+In our study, we only use the SNPs that are observed in both the reference and
+the study group and have the same alleles.
+Thus we have the same number of SNPs for both groups.
 The genotypes of the reference group are stored in the $p \times n$ matrix $X$,
 and those of the study group are stored in the $p \times m$ matrix $Y$.
-  For most genetic data, $n \ll p$ and $m \ll p$.
-\item Principle component analysis \\
-  Given a multi-dimensional data, principle component analysis (PCA) looks for the directions, which are called the principle components (PC), in which the data has the greatest variance.
+For genotype data, it is common for $p$ to be significantly larger than $n$.
+Thus we assume this to be true except for the part where we discuss the randomized SVD algorithm,
+which is developed to handle datasets with exceedingly large $n$.
+
+\subsubsection{Principle Component Analysis}\label{pca-intro}
+  Given a multi-dimensional data, principle component analysis (PCA) looks for the directions in which the data has the greatest variance.
+These unit vectors are called the principle components (PC's).
 The PC's form an orthonormal basis for the sample space,
 and we can ignore all but the first few PC's to reduce the dimension of the data.
-  In particular, let $X=UDV^T$ be the singular value decomposition (SVD) of $X$,
+In particular, let $X=UDV^T$ be the SVD of $X$,
 where $U$ is a $p \times n$ matrix whose columns are the PC's,
@@
 and $V$ an $n \times n$ matrix such that each row contains the standardized PC scores of an individual.
-  (Alternatively, we can also use eigendecomposition to find the matrices since $X^T X = VD^2V^T$ and $X X^T = UD^2U^T$.)
-  We use $D$ to sort out the most influential $k$ PC's ($k \ll n$) and look at the coresponding columns of $V$ to find their PC scores.
-  For estimating ancestral histroy from genotypes, often the first $4$ PC's will be sufficient for providing the information we need.
+(Alternatively, we can use eigendecomposition to find the matrices since $X^T X = VD^2V^T$ and $X X^T = UD^2U^T$.)
+We sort the diagonal elements of $D$ in descending order and rearrange the
+columns of $U$ and $V$ accordingly.
+We are interested in the top $k$ PC's (where $k \ll n$) and look at the top $k$
+columns of $V$ for their PC scores.
 
-  PCA allows us to use the reference data to find the PC scores of the study data.
-  The standard method is to append a study individual's genotype to the reference matrix and then conduct a SVD of the augmented matrix.
-  The corresponding row in the matrix $V$ is the study individual's PC scores.
+\subsubsection{Estimating PC scores}
+
+We assume that the reference data and the study data are generated from the same
+population.
+When the genotypes of all individuals of the population are given (which is very
+unlikely in global genetic studies),
+we use the method in \Cref{pca-intro} to find the PC scores of the individuals
+we are interested in.
+In practice, we have a reference dataset and a study dataset assumed to be generated from the
+same population and use the former to estimate the populational PC scores of the latter.
+There are several computational methods one can use to do this estimation.
+In \Cref{sec:compu}, we compare the performance of four of them.
+
+\subsection{Computational Methods}\label{sec:compu}
+
+\subsubsection{Simple Projection}
+The standard way for estimating PC scores is by projecting the study
+individuals' data points to the PCs of the reference individuals.
+We call this method ``simple projection'' (SP).
+The algorithm goes as follows.
+
+\begin{enumerate}
+\item Compute $X^T X$.
+  (Computational complexity: $\bO[n^2p]$.)  
+\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
+  ($\bO[n^3]$.)
+\item Calculate $W = Y^T (X V D_1^{-1})$, where $Y$ is the data matrix for all the study individuals, $D_1$ the first $k_1$ columns of $D$. ($\bO[mpk_1 + pnk_1]$.)
 \end{enumerate}
 
-\subsection{Computational Methods}
+Total computational complexity: $\bO[n^2p + n^3 + mpk_1 + pnk_1)] \approx \bO[n^2p + mk_1p]$,
+given that $k_1 \ll n \ll p$.
 
-The usefulness of PCA for genetic data largely depends on its computational cost.
-The computational complexity of SVD for a $p \times n+1$ matrix is $\min((n+1)^2p, (n+1)p^2) = n^2p $ for $1 \ll n \ll p$.
-Thus the total computational complexity for performing PCA on $m$ individuals by using the standard method is $mn^2p$.
-This value increases quadratically as we increase the reference sample size to increase the accuracy of the PC scores.
-Alternative computational methods have been developed to improve the efficiency of PCA.
-In this study, we compare four such methods: TRACE, projection, online TRACE, and HDPCA projection.
+As the reference sample size increases,
+the computational complexity of SP stays as a constant,
+except for decomposing the reference matrix, which is done once for all study individuals.
+However, it is known that the projection method tends to underestimates the magnitude of the PC scores when $n \ll p$,
+which we call the ``shrinking bias''. (See \cite{dey}.)
+This limits the accuracy of SP for high dimensional data.
 
-\subsubsection{TRACE}
+\subsubsection{Augment, Decompose, and Transform}
 
-TRACE implements the standard PCA method, with additional procedures for handling minor issues such as missing data.
-An important improvement in TRACE on the standard method is the inclusion of Procrustes analysis.
-Given two data sets with the same sample size and possibly different dimensions,
-Procrustes analysis finds the linear transformation that maps one data matrix as close as possible to the other data matrix.
-In our contex, the PC scores of the reference matrix is computed and we obtain the $n \times n$ matrix $V$.
-For a study individal with genotype $y$, we decompose $\tilde{X} = (X, y)$ into $\tilde{X} = \tilde{U}\tilde{D}\tilde{V}^T$.
-(TRACE uses eigendecomposition to find the ``$V$'' matrix to reduce computational complexity since we do not need the ``$U$'' matrix here.)
-Let
+One way to solve the shrinking bias of SP is developed by \cite{wang}.
+In this method, each study individual's data point is appended to the reference
+data matrix.
+Then SVD is applied to the augmented matrix,
+and PC scores in this augmented sample are found for both the reference
+individuals and the study individual.
+The the PC scores of the reference individuals in this augmented sample is
+compared with the PC scores of the reference individuals in the reference
+sample,
+which is calculated in the beginning,
+and a linear (Procrustes) map is found to transform the latter into the former
+with the smallest error.
+This transformation is then applied to the PC scores of the study individual in
+the augmented sample.
+The resulted PC scores are used as an estimate for the study individual's
+population PC scores.
+Specifically, let
 \[
@@
 where the dimensions of $V_\text{ref}$, $\tilde{V}_\text{ref}$, and $\tilde{V}_\text{stu}$ are $n \times k_1$, $n \times k_2$, and $1 \times k_2$, respectively, with $ 1 \leq k_1 \leq k_2 \leq n+1$. 
-We use Procrustes analysis to find
+Next, we use Procrustes analysis to find
 \[
@@
   This is exactly the same as computing $\tilde{X}^T \tilde{X}$,
-  where $\tilde{X} = (X, y)$ is the reference group with the testing individual added.
+  where $\tilde{X} = (X, y)$ is the reference group with the study individual added.
   ($\bO[np]$.)
@@
   ($\bO[n^3]$.)
-\item Project the first $k_2$ columns of $\tilde{V}$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the testing individual (the last row of $V$),
+\item Project the first $k_2$ columns of $\tilde{V}$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual (the last row of $V$),
   where $k_1 < k_2 \ll n$.
   (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
-  \item Go to step 3 for the next testing individual.
+  \item Go to step 3 for the next study individual.
 \end{enumerate}
@@
 
-TRACE is the slowest among the computational methods we study.
-However, its result is also the most reliable.
+Although accurate, ADT is the slowest among the other three computational
+methods.
+Its computational complexity increases cubically as we increase the reference sample size to increase the accuracy of the PC scores.
+The ADT method is has been implemented in the TRACE software by \cite{wang}.
 
-\subsubsection{Projection}
-An exchange for runtimes with accuracy is the projection method.
-For this method, we simply find the PC's of the reference matrix and project the study individual's genotype onto them,
-as shown below:
+\subsubsection{Adjusted Projection}
 
+A more efficient way to solve the shrinking bias of SP is to estimate the
+magnitude and angle of the shrinkage and adjust the study individuals' PC scores accordingly.
+This method has been developed by \cite{dey} and implemented in the High
+Dimension Principal Component Analysis (HDPCA) in the R language.
+HDPCA reads the PC scores and variances calculated by projection and estimate the shrinkage factor to restore the unshrinked scores. 
+The algorithm is
 \begin{enumerate}
@@
   ($\bO[n^3]$.)
-\item Calculate $W = Y^T (X V D_1^{-1})$, where $Y$ is the data matrix for all the testing individuals, $D_1$ the first $k_1$ columns of $D$. ($\bO[mpk_1 + pnk_1]$.)
+\item Calculate $W = Y^T (X V D_1^{-1})$, where $Y$ is the data matrix for all the study individuals, $D_1$ the first $k_1$ columns of $D$. ($\bO[mpk_1 + pnk_1]$.)
+  \item Input $D$ and $W$ to the pc\_adjust function in the HDPCA function to obtain the PC scores adjusted for the shrinking bias.
+($\bO[p]$ for finding the number of distant spikes (function select.nspike), $\bO[n]$ for adjusting for the shrinking bias (function hdpc\_est))
 \end{enumerate}
 
-Total computational complexity: $\bO[n^2p + n^3 + mpk_1 + pnk_1)] \approx \bO[n^2p + mk_1p)]$,
-given that $k_1 \ll n \ll p$.
+Total computational complexity: $\bO[n^2p + mk_1(p + n) + p + n] \approx \bO[n^2p + mk_1p]$,
+given that $m$ or $n$ is large. The computational complexities between simple projection and adjusted projection are the same.
+The accuracy of adjusted projection will be empirically shown later.
 
-As the reference sample size increases,
-the computational complexity of the projection method only increases quadratically,
-(except for decomposing the reference matrix, which is done once for all study individuals,)
-while that of TRACE increases cubically.
-However, it is known that the projection method tends to underestimates the magnitude of the PC scores when $n \ll p$,
-which is known as the ``shrinking bias''.
+\subsubsection{Online ADT}
 
-\subsubsection{Online TRACE}
-
-A remedy for TRACE is to reduce the computational complexity withou much sacrifice of the accuracy.
-This is achieved by the online TRACE method.
-Notice that in TRACE, only the top $k_2$ PC's are used.
-Online TRACE calculates a close approximation of the first $k_3$ columns of $\tilde{V}$ for some $k_2 \leq k_3 \ll n$ and keeps everything else the same as in TRACE, including Procrustes analysis. 
+A remedy for ADT is to reduce the computational complexity withou much sacrifice of the accuracy.
+This is achieved by the online ADT (OADT) method.
+Notice that in OADT, only the top $k_2$ PC's are used.
+OADT uses the online SVD method, which calculates a close approximation of the first $k_3$ columns of $\tilde{V}$ for some $k_2 \leq k_3 \ll n$ and keeps everything else the same as in ADT, including Procrustes analysis.
 The algorithm is
@@
   ($\bO[nk_3^2]$.)
-\item Project the first $k_2$ columns of $\tilde{V}_3$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the testing individual,
+\item Project the first $k_2$ columns of $\tilde{V}_3$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual,
   where $k_1 < k_2  < k_3 \ll n$.
   (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
-  \item Go to step 4 for the next testing individual.
+  \item Go to step 4 for the next study individual.
 \end{enumerate}
@@
 
-Here we see that the computational complexity of online TRACE for the study individuals increases quadratically with respect to the reference sample size.
-This makes the runtimes of online TRACE close to those of the projection method.
-The closeness between the results given by online TRACE and TRACE will be empirically shown later.
+Here we see that the computational complexity of OADT for the study individuals increases linearly with respect to the reference sample size.
+This makes the runtimes of OADT close to those of SP.
+The closeness between the results given by OADT and ADT will be empirically shown later.
 
-\subsubsection{HDPCA Projection}
+A comparison of computatinal complexities of the four PCA methods are shown in \Cref{tbl:cplx}.
 
-As online TRACE reduces the computational complexity of TRACE,
-we have HDPCA projection that improves the accuracy of projection without increasing the runtimes.
-HDPCA reads the PC scores and variances calculated by projection and estimate the shrinkage factor to restore the unshrinked scores. 
-The algorithm is
-\begin{enumerate}
-\item Compute $X^T X$.
-  (Computational complexity: $\bO[n^2p]$.)  
-\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
-  ($\bO[n^3]$.)
-\item Calculate $W = (Y^T X) V D^{-1}$, where $Y$ is the data matrix for all the testing individuals. ($\bO[mpn + mn^2]$.)
-  \item Input $D$ and $W$ to the pc\_adjust function in the HDPCA function to obtain the PC scores adjusted for the shrinking bias.
-($\bO[p]$ for finding the number of distant spikes (function select.nspike), $\bO[n]$ for adjusting for the shrinking bias (function hdpc\_est))
-\end{enumerate}
-
-Total computational complexity: $\bO[n^2p + mn(p + n) + p + n] \approx \bO[n^2p + mn(p+n)]$,
-given that $m$ or $n$ is large.
-
-The computational complexities between projection and HDPCA projection are the same.
-The accuracy of HDPCA projection will be empirically shown later.
-
-\subsubsection{Computational Complexity Summary}
-A comparison of computatinal complexities of the four PCA methods are shown in \Cref{tbl:cplx}.
+\subsubsection{Randomized SVD}
+In addition to estimating the PC scores of the study individuals,
+we have another algorithm that aims at improving the decomposition of the
+training data.
+For SP, AP, ADT, and OADT, we need to find the covariance matrix of the training
+data, which has dimension $n \times n$, and conduct eigendecomosition on it.
+However, when $n$ is exceedingly large,
+the covariance matrix cannot be loaded into the memory of an average personal
+computer.
+A randomized SVD (RSVD)algorithm has been developed to handle this issue.
+RSVD is slower than standard SVD,
+but it does not require loading the whole covariance matrix into the memory.
 
@@
     \hline
-    TRACE & $\bO[n^2 p + mn(p + n^2)]$ \\
+    Simple Projection (SP) & $\bO[n^2p + mk_1p]$ \\
     \hline
-    Projection & $\bO[n^2p + mk_1p)]$ \\
+    Augment-Decompose-Transform (ADT) & $\bO[n^2 p + mn(p + n^2)]$ \\
     \hline
-    Online TRACE & $\bO[n^2 p + mk_3(p + k_3 n)]$ \\
+    Adjusted Projection (AP) & $\bO[n^2p + mk_1p]$ \\
     \hline
-    HDPCA Projection & $\bO[n^2p + mn(p+n)]$ \\
+    Online Augment-Decompose-Transform (OADT) & $\bO[n^2 p + mk_3(p + k_3 n)]$ \\
     \hline
@@
     $n$ is the number of individuals in the reference group,
-    and $m$ is the number of individuals in the  testing group.
+    and $m$ is the number of individuals in the  study group.
     For projection, $k_1$ is the number of PCs needed
-    For Online TRACE, $k_3$ is the number of PCs calculated by online SVD, which is usually much less than $n$.}
+    For OADT, $k_3$ is the number of PCs calculated by online SVD, which is usually much less than $n$.}
   \label{tbl:cplx}
@@
 
-In this study, we used both simulated data and real-world data.
+In this study, we used both simulated data and real-life data to conduct
+empirical tests of the computational methods for PCA.
 
@@
 $2m / K^2$)
-\item M: Migration rate. (Varies)
+\item M: Migration rate. (Fixed to $100$)
 \end{enumerate}
 
-\subsubsection{Real-World Datasets}
+For our study, we used different reference and study sample sizes.
+First, we fix
+the study size to be $m=200$ and increase $n$ from $1000$ to $3000$ with the
+increment equal to $500$.
+Next, we fix
+the reference size to be $n=600$ and increase $m$ from $1000$ to $3000$ with the increment equal to $500$.
 
-The real-world datasets used in our study are the HGDP and 1000 Genomes datasets.
+For comparing the speed of the algorithms,
+we only compare the runtimes for the study individuals.
+The time spent on calculations that are once-for-all (such as finding the SVD of
+the reference matrix) is ignored.
+
+For comparing the accuracy, we use two different measures.
+First, for a given cell of the grid, we compute the mean of the PC scores of the 
+reference individuals located at this cell.  
+This is the estimated coordinate of the cell.
+Next, we find the Euclidean distance between the estimated PC scores of the
+study individuals and the estimated coordinates of the cells they are located
+at.
+These are the geographic estimation errors for the study individuals.
+The average of these errors are used to measure the accuracy of the
+computational methods (that is, the square root of the sum of squares divided by the
+study size). 
+Second, since ADT is believed to be the most reliable, we compare the average
+Euclidean distance between the PC score estimates calculated by another method
+to those calculated by ADT. 
+This golden-standard error and the geographic error both range from zero to infinity.
+
+The comparison of the runtimes and accuracy are conducted by using both the
+standard SVD and randomized SVD on the training data matrix.
+
+\subsubsection{Real-Life Datasets}
+
+The real-life datasets used in our study are the UKBioBank and 1000 Genomes datasets.
 They contain the genotype readings of individuals from different ethnic groups over the world.
 For better comparison, we select only the SNPs that are included in both datasets,
-the number of which is approximately $6 \times 10^5$.
-Moreover, we removed individuals who have at least parent also in the dataset.
-The samples size is approximately $900$ for HGDP and $2300$ for 1000 Genomes.
+the number of which is approximately $1.5 \times 10^5$.
+We use the 1000 Genomes dataset as the reference and removed individuals who have at least one parent who is also in the dataset.
+The reference samples size is approximately $2300$.
+Moreover, we use the first 1000 individuals in the UKBioBank data set out of a
+total number of $5 \times 10^5$ individuals. 
+The measures we use for comparing speed and accuracy are the same as for
+simulated data,
+except that we do not use the geographic error here.
+In addition, both the standard and randomized SVD on the training data set are tested.
 
-\section{Results (Simulation Study)}
+\section{Results}
 
-We now present the results of our study.
-
-\subsection{Parameter Setting}
-
-\begin{enumerate}
-\item $n$ fixed, $m$ changing
-\begin{enumerate}
-\item $p = 100000$
-\item $n = 600$
-\item $m$: $\min = 1000$, $\max = 5000$, $\text{increment} = 500$ 
-\item $migration = 100$
-\end{enumerate}
-\item $m$ fixed, $n$ changing
-\begin{enumerate}
-\item $p = 100000$
-\item $n$: $\min = 1000$, $\max = 5000$, $\text{increment} = 500$ 
-\item $m = 200$
-\item $migration = 100$
-\end{enumerate}
-\end{enumerate}
-
-\subsection{Computation Time}
-
-\includegraphics[width=0.98\textwidth]{runtimes}
-
-\subsection{Accuracy}
-
-Next, we show the accuracy of each method.
-We have two ways to measure the accuracy.
-First, we can use the information from the simulation given by GGS to categorize the reference individuals into $K^2$ clusters and find the mean of each cluster,
-which will be treated as the center of that cluster.
-We then find the Euclidean distance between the study individuals and the centers,
-again using the true locations in the grid to categorize the study individuals.
-
-Second, we can use Online TRACE as the standard and find the Euclidean distance between the study individuals' PC scores calculated by Online TRACE and another method.
-
-The parameter setting is the same as in comparing the runtimes.
-
-\includegraphics[width=0.98\textwidth]{err_refcenter.pdf}
-
-\includegraphics[width=0.98\textwidth]{err_online.pdf}
+See plots.
 
 \section{Discussion}
+
+We find that AP improved the accuracy of SP without increasing the runtimes.
+Moreover, OADT improves the speed of ADT without much sacrifice of accuracy.
+Inaddition, randomized SVD is found to be as reliable as standard SVD in
+calculating the PC's of the reference group.
+
+\newpage
+
+\includegraphics[width=0.98\textwidth]{runtimes_rand}
+
+\includegraphics[width=0.98\textwidth]{err_refcenter_rand}
+
+\includegraphics[width=0.98\textwidth]{err_trace_rand}
+
+\includegraphics[width=0.98\textwidth]{err_refcenter_rand}
+
+\includegraphics[width=0.98\textwidth]{kgn_allChr_ukb_orphans_ukb_1k_comb}
+
+\newpage
+
+\includegraphics[width=0.98\textwidth]{kgn_allChr_ukb_orphans_ukb_1k_rand}
 
@@
 
+ \bibitem[Reich et al.(2008)]{reich} Reich, D., Price, A. L., \& Patterson, N.
+   (2008). Principal component analysis of genetic data. {\it Nature genetics}, 40(5), 491.
+   Reich, D., Price, A., Patterson, N (2008).
+   Principle component analysis of genetic data. {\it Nature Genetics}
+
+   \bibitem[Price et al.(2006)]{price} Price, A. L., Patterson, N. J., Plenge,
+     R. M., Weinblatt, M. E., Shadick, N. A., \& Reich, D. (2006). Principal
+     components analysis corrects for stratification in genome-wide association
+     studies. {\it Nature genetics}, 38(8), 904.
+
 \end{thebibliography}
 
-    % \begin{enumerate}
-    %     \item $X$ ($p$ by $n$): 
-    %         Genotype readings for the reference group.
-    %     \item $Y = (y_1 \cdots y_m)$ ($p$ by $m$): 
-    %         Genotype readings for the study group.
-    % \end{enumerate}
-
-% \begin{enumerate}
-% \item HGDP
-%   The HGDP dataset contains individuals from different ethnic groups in the world. It includes about $6 \times 10^5$ SNPs and 
-%   \begin{enumerate}
-%   \item Individuals from different countries in the world.
-%   \item We only use the SNPs that are also included in the 1000
-%     Genomes dataset.
-%   \item $p \approx 6 \times 10^5, n \approx 900$.
-%   \end{enumerate}
-% \item 1000 Genomes
-%   \begin{enumerate}
-%   \item Individuals from different countries in the world.
-%   \item We only use the SNPs that are also included in the HGDP
-%     dataset.
-%   \item $p \approx 6 \times 10^5, n \approx 2300$.
-%   \end{enumerate}
-% \item Data filtering
-%   \begin{enumerate}
-%   \item We only used the SNPs that are included in both the HGDP and
-%     1000 Genomes datasets.
-%   \item We have removed the individuals with at least one parent in
-%     the dataset.
-%   \end{enumerate}
-% \end{enumerate}
-
 \end{document}

### 2018-02-22 mergePlot.R

@@ -1,0 +1,92 @@
+library(RColorBrewer)
+library(scales)
+
+
+meth.n <- 7
+chg.var <- "n"
+chg.jobid <- "27483940"
+chg.range <- seq(1000, 3000, 500)
+chg.taskid <- c(0:4)
+chg.length <- length(chg.range)
+
+file.head <- "../flux/data/"
+file.neck <- paste0(
+  chg.var,
+  "Chg_rand_",
+  chg.jobid,
+  "_",
+  chg.taskid
+)
+file.upper <- paste0(
+  file.head,
+  file.neck,
+  "/",
+  file.neck,
+  "_100000_")
+if(chg.var == "n"){
+  file.ab <- paste(chg.range, "200", sep = "_")
+  xlab <- "Reference Size"
+} else if(chg.var == "m"){
+  file.ab <- paste("600", chg.range, sep = "_")
+  xlab <- "Study Size"
+} else{
+  stop("Invalid changing variable. Must be m or n. ")
+}
+file.leg <- "_2_1_100.rand"
+out.pref <- paste0(file.upper,
+                  file.ab,
+                  file.leg)
+
+runtimes.ref <- matrix(0, chg.length, meth.n + 1)
+runtimes.ref[,1] <- chg.range
+runtimes.stu <- matrix(0, chg.length, meth.n + 1)
+runtimes.stu[,1] <- chg.range
+accuracy.ctr <- matrix(0, chg.length, meth.n + 1)
+accuracy.ctr[,1] <- chg.range
+accuracy.gold <- matrix(0, chg.length, meth.n + 1)
+accuracy.gold[,1] <- chg.range
+for(i in 1:chg.length){
+  runtimes.file <- paste(out.pref[i], "runtimes", sep = ".")
+  accuracy.file <- paste(out.pref[i], "accuracy", sep = ".")
+  runtimes.df <- read.table(runtimes.file, header = TRUE)
+  accuracy.df <- read.table(accuracy.file, header = TRUE)
+  meth.name <- runtimes.df$method
+  runtimes.ref[i,(1:meth.n)+1] <- runtimes.df$reference
+  runtimes.stu[i,(1:meth.n)+1] <- runtimes.df$study
+  accuracy.ctr[i,(1:meth.n)+1] <- accuracy.df$err.center
+  accuracy.gold[i,(1:meth.n)+1] <- accuracy.df$err.golden
+}
+
+
[…]

### 2018-02-24 doc/paper.tex
Add readme
Add nChg and mChg plots to paper.

@@
 
-See plots.
+See \Cref{fig:nChg} and \Cref{fig:mChg} for the comparison of the runtimes and accuracy.
 
@@
 
-\includegraphics[width=0.98\textwidth]{runtimes_rand}
+\begin{figure}[p]
+  \label{fig:nChg}
+  \centering
+  \includegraphics[width=0.98\textwidth]{nChg}
+  \caption{
+    Runtimes and deviations for different methods as the reference sample size increases.
+    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
+    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+  }
+\end{figure}
 
-\includegraphics[width=0.98\textwidth]{err_refcenter_rand}
+\begin{figure}[p]
+  \label{fig:mChg}
+  \centering
+  \includegraphics[width=0.98\textwidth]{mChg}
+  \caption{
+    Runtimes and deviations for different methods as the study sample size increases.
+    The reference sample size is fixed to 600 and the number of SNPs is 100,000 (1000 per genealogy).
+    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+  }
+\end{figure}
 
-\includegraphics[width=0.98\textwidth]{err_trace_rand}
+% \includegraphics[width=0.98\textwidth]{runtimes_rand}
 
-\includegraphics[width=0.98\textwidth]{err_refcenter_rand}
+% \includegraphics[width=0.98\textwidth]{err_refcenter_rand}
 
-\includegraphics[width=0.98\textwidth]{kgn_allChr_ukb_orphans_ukb_1k_comb}
+% \includegraphics[width=0.98\textwidth]{err_trace_rand}
+
+% \includegraphics[width=0.98\textwidth]{err_refcenter_rand}
+
+% \includegraphics[width=0.98\textwidth]{kgn_allChr_ukb_orphans_ukb_1k_comb}
 
@@
 
-\includegraphics[width=0.98\textwidth]{kgn_allChr_ukb_orphans_ukb_1k_rand}
+% \includegraphics[width=0.98\textwidth]{kgn_allChr_ukb_orphans_ukb_1k_rand}
 

### 2018-02-24 mergePlot.R

@@ -56,37 +56,55 @@ for(i in 1:chg.length){
   accuracy.ctr[i,(1:meth.n)+1] <- accuracy.df$err.center
   accuracy.gold[i,(1:meth.n)+1] <- accuracy.df$err.golden
 }
-
-
-par(oma = c(4, 4, 4, 2), mar = c(5, 4, 4, 2), mfrow = c(2,2))
-lty = c(1, 1, 2, 1, 2, 1, 2)
-pch = c(15, 16, 1, 17, 2, 18, 5)
-col = alpha(brewer.pal(meth.n, "Paired"), 1)
-col = col[c(meth.n, 1:(meth.n-1))]
-cex = 2
-lwd = 2
-col.main = "gray25"
-matplot(runtimes.ref[,1], runtimes.ref[,(1:meth.n)+1],
-        main = "Runtimes (Reference)",
-        xlab = xlab,
-        ylab = "Seconds",
-        lty = lty, pch = pch, col = col, cex = cex, type = "b", lwd = lwd, col.main = col.main)
-matplot(runtimes.stu[,1], runtimes.stu[,(1:meth.n)+1],
-        main = "Runtimes (Study)",
-        xlab = xlab,
-        ylab = "Seconds",
-        lty = lty, pch = pch, col = col, cex = cex, type = "b", lwd = lwd, col.main = col.main)
-matplot(accuracy.ctr[,1], accuracy.ctr[,(1:meth.n)+1],
-        main = "Deviations (Distance from Centers)",
-        xlab = xlab,
-        ylab = "PC Score Units",
-        lty = lty, pch = pch, col = col, cex = cex, type = "b", lwd = lwd, col.main = col.main)
-matplot(accuracy.gold[,1], accuracy.gold[,(1:meth.n)+1],
-        main = "Deviations (Distance from TRACE)",
-        xlab = xlab,
-        ylab = "PC Score Units",
-        lty = lty, pch = pch, col = col, cex = cex, type = "b", lwd = lwd, col.main = col.main)
-par(fig = c(0, 1, 0, 1), oma = c(0, 0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE)
-plot(0, 0, type = "n", bty = "n", xaxt = "n", yaxt = "n")
-legend("bottom", legend = meth.name, xpd = TRUE, horiz = TRUE, inset = c(0, 0), bty = "n", pch = pch, col = col, cex = cex*0.8, pt.lwd = lwd)
[…]

### 2018-02-26 doc/paper.tex
Add PC score plots to paper.

@@
 
-\subsubsection{Simple Projection}
+\subsubsection{Simple Projection (SP)}
 The standard way for estimating PC scores is by projecting the study
@@
 
-\subsubsection{Augment, Decompose, and Transform}
+\subsubsection{Augment, Decompose, and Transform (ADT)}
 
@@
 
-\subsubsection{Adjusted Projection}
+\subsubsection{Adjusted Projection (AP)}
 
@@
 
-\subsubsection{Online ADT}
+\subsubsection{Online ADT (OADT)}
 
@@
 
-\begin{table} 
-  \centering
-  \begin{tabular}{|l|l|}
-    \hline
-    Method & Computational Complexity \\ 
-    \hline
-    Simple Projection (SP) & $\bO[n^2p + mk_1p]$ \\
-    \hline
-    Augment-Decompose-Transform (ADT) & $\bO[n^2 p + mn(p + n^2)]$ \\
-    \hline
-    Adjusted Projection (AP) & $\bO[n^2p + mk_1p]$ \\
-    \hline
-    Online Augment-Decompose-Transform (OADT) & $\bO[n^2 p + mk_3(p + k_3 n)]$ \\
-    \hline
-  \end{tabular}
-  \caption{
-    Comparison of computational complexity for the three methods.
-    Here $p$ is the number of SNPs,
-    $n$ is the number of individuals in the reference group,
-    and $m$ is the number of individuals in the  study group.
-    For projection, $k_1$ is the number of PCs needed
-    For OADT, $k_3$ is the number of PCs calculated by online SVD, which is usually much less than $n$.}
-  \label{tbl:cplx}
-\end{table}
-
 \subsection{Data}
@@
 
-See \Cref{fig:nChg} and \Cref{fig:mChg} for the comparison of the runtimes and accuracy.
+The predicted PC scores are shown in \Cref{fig:n1000} and \Cref{fig:ukb}.
+For the comparison of the runtimes and accuracy for different reference and sample sizes for the simulated data,
+see \Cref{fig:nChg} and \Cref{fig:mChg}.
+For comparing the methods for the real-life data, see
 
@@
 Moreover, OADT improves the speed of ADT without much sacrifice of accuracy.
-Inaddition, randomized SVD is found to be as reliable as standard SVD in
+In addition, randomized SVD is found to be as reliable as standard SVD in
 calculating the PC's of the reference group.
-
-\newpage
-
-\begin{figure}[p]
-  \label{fig:nChg}
-  \centering
-  \includegraphics[width=0.98\textwidth]{nChg}
-  \caption{
-    Runtimes and deviations for different methods as the reference sample size increases.
-    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
-    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
-  }
-\end{figure}
-
-\begin{figure}[p]
-  \label{fig:mChg}
-  \centering
-  \includegraphics[width=0.98\textwidth]{mChg}
-  \caption{
-    Runtimes and deviations for different methods as the study sample size increases.
-    The reference sample size is fixed to 600 and the number of SNPs is 100,000 (1000 per genealogy).
-    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
-  }
-\end{figure}
-
-% \includegraphics[width=0.98\textwidth]{runtimes_rand}
-
-% \includegraphics[width=0.98\textwidth]{err_refcenter_rand}
-
-% \includegraphics[width=0.98\textwidth]{err_trace_rand}
-
-% \includegraphics[width=0.98\textwidth]{err_refcenter_rand}
-
-% \includegraphics[width=0.98\textwidth]{kgn_allChr_ukb_orphans_ukb_1k_comb}
-
-\newpage
-
-% \includegraphics[width=0.98\textwidth]{kgn_allChr_ukb_orphans_ukb_1k_rand}
 
@@
 
+\newpage
+
+\begin{table} 
+  \centering
+  \begin{tabular}{|l|l|}
+    \hline
+    Method & Computational Complexity \\ 
+    \hline
+    Simple Projection (SP) & $\bO[n^2p + mk_1p]$ \\
+    \hline
+    Augment-Decompose-Transform (ADT) & $\bO[n^2 p + mn(p + n^2)]$ \\
+    \hline
+    Adjusted Projection (AP) & $\bO[n^2p + mk_1p]$ \\
+    \hline
+    Online Augment-Decompose-Transform (OADT) & $\bO[n^2 p + mk_3(p + k_3 n)]$ \\
+    \hline
+  \end{tabular}
+  \caption{
+    Comparison of computational complexity for the three methods.
+    Here $p$ is the number of SNPs,
+    $n$ is the number of individuals in the reference group,
+    and $m$ is the number of individuals in the  study group.
+    For projection, $k_1$ is the number of PCs needed
+    For OADT, $k_3$ is the number of PCs calculated by online SVD, which is usually much less than $n$.}
+  \label{tbl:cplx}
+\end{table}
+
+\begin{table} 
+  \centering
+  \begin{tabular}{|l|l|l|l|l|}
+    \hline
+    Method & Reference Runtimes & Study Runtimes & Reference Center Error & Golden Standard Error \\ 
+    \hline
+    Simple Projection (SP) &  \\
+    \hline
+    Augment-Decompose-Transform (ADT) & $\bO[n^2 p + mn(p + n^2)]$ \\
+    \hline
+    Adjusted Projection (AP) & $\bO[n^2p + mk_1p]$ \\
+    \hline
+    Online Augment-Decompose-Transform (OADT) & $\bO[n^2 p + mk_3(p + k_3 n)]$ \\
+    \hline
+  \end{tabular}
+  \caption{
+    Comparison of computational complexity for the three methods.
+    Here $p$ is the number of SNPs,
+    $n$ is the number of individuals in the reference group,
+    and $m$ is the number of individuals in the  study group.
+    For projection, $k_1$ is the number of PCs needed
+    For OADT, $k_3$ is the number of PCs calculated by online SVD, which is usually much less than $n$.}
+  \label{tbl:cplx}
+\end{table}
+
+\begin{figure}[p]
+  \centering
+  \includegraphics[width=0.98\textwidth]{n1000}
+  \caption{
+    Predicted PC scores on top of reference PC scores.
+    The four populations in the $2 \times 2$ grid simulated by the GGS software can be clearly identified.
+    Reference individuals are in color,
+    while predicted PC scores for the study individuals are in gray.
+    Two measurements of deviations (accuracy) are used.
+    The reference center error is the square root of the average squared distance between the study individuals and the population centers, which are the means of the PC scores of the reference individuals by the populations.
+    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADT.
+    The reference size is 1000 and the study size is 200.
+    There are 100,000 loci (1000 per genealogy) and the migration rate is 100.
+  }
+  \label{fig:n1000}
+\end{figure}
+
+\begin{figure}[p]
+  \centering
+  \includegraphics[width=0.98\textwidth]{ukb}
+  \caption{
+    The reference individuals are from the 1000 Genomes dataset with 2492 unrelated individuals from 5 superpopulations: African, American, East Asian, European, and South Asian. 
+    The study sample contain 1000 individuals from the UKBioBank dataset.
+    There are about 125,000 loci shared by the reference and study data.
+    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADT.
+  }
+  \label{fig:ukb}
+\end{figure}
+
+\begin{figure}[p]
+  \centering
+  \includegraphics[width=0.98\textwidth]{nChg}
+  \caption{
+    Runtimes and deviations for different methods as the reference sample size increases.
+    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
+    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+  }
+  \label{fig:nChg}
+\end{figure}
+
+\begin{figure}[p]
+  \centering
+  \includegraphics[width=0.98\textwidth]{mChg}
+  \caption{
+    Runtimes and deviations for different methods as the study sample size increases.
+    The reference sample size is fixed to 600 and the number of SNPs is 100,000 (1000 per genealogy).
+    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+  }
+  \label{fig:mChg}
+\end{figure}
+
+% \includegraphics[width=0.98\textwidth]{runtimes_rand}
+
+% \includegraphics[width=0.98\textwidth]{err_refcenter_rand}
+
+% \includegraphics[width=0.98\textwidth]{err_trace_rand}
+
+% \includegraphics[width=0.98\textwidth]{err_refcenter_rand}
+
+% \includegraphics[width=0.98\textwidth]{kgn_allChr_ukb_orphans_ukb_1k_comb}
+
+\newpage
+
+% \includegraphics[width=0.98\textwidth]{kgn_allChr_ukb_orphans_ukb_1k_rand}
+
+\newpage
+
 \end{document}

### 2018-02-26 trace.rand.cpp
Change test_hdpcaRand time to runtimes_sum[projRand] (rather than runtimes[projRand])

@@ -1506,7 +1506,7 @@ int main(int argc, char* argv[]){
 	runtimes_sum["test_hdpca"] = runtimes["proj"];
 	runtimes_sum["test_onlRand"] = runtimes["onlRandSVD"] + runtimes["onlRandProcrust"];
 	runtimes_sum["test_projRand"] = runtimes["projRand"];
-	runtimes_sum["test_hdpcaRand"] = runtimes["test_projRand"];
+	runtimes_sum["test_hdpcaRand"] = runtimes_sum["test_projRand"];
 	cout << "Runtime breakdown (sec): " << endl;
 	foutLog << "Runtime breakdown (sec): " << endl;
 	for(map<string, float>::iterator it = runtimes.begin(); it != runtimes.end(); it++){

### 2018-02-26 doc/paper.tex
Change test_hdpcaRand time to runtimes_sum[projRand] (rather than runtimes[projRand])
Create comparison tables and rsvd refs in paper.tex

@@
 computer.
-A randomized SVD (RSVD)algorithm has been developed to handle this issue.
+A randomized SVD (RSVD) algorithm has been developed to handle this issue.
 RSVD is slower than standard SVD,
 but it does not require loading the whole covariance matrix into the memory.
+For the details of the RSVD algorithm, see \cite{halko} and \cite{abraham}.
 
@@
 
+\bibitem[Abraham \& Inouye(2014)]{abraham} Abraham, G., \& Inouye, M. (2014). Fast principal component analysis of large-scale genome-wide data. {\it PloS one}, 9(4), e93766.
+
 \bibitem[Brand(2002)]{brand} Brand, M. (2002). Incremental singular value decomposition of uncertain data with missing values. {\it Computer Vision--ECCV 2002}, 707-720.
@@
 Chicago
+
+\bibitem[Halko et al.(2011)]{halko} Halko, N., Martinsson, P.
+G., \& Tropp, J.
+A.
+(2011).
+Finding structure with randomness: Probabilistic algorithms for constructing approximate matrix decompositions.
+{\it SIAM review}, 53(2), 217-288.
 
@@
   \centering
-  \begin{tabular}{|l|l|}
+  \begin{tabular}{|l|l|l|}
     \hline
-    Method & Computational Complexity \\ 
+    Method & Reference Complexity & Study Complexity \\ 
     \hline
-    Simple Projection (SP) & $\bO[n^2p + mk_1p]$ \\
+    Simple Projection (SP) & $\bO[n^2p]$ & $\bO[mkp]$ \\
     \hline
-    Augment-Decompose-Transform (ADT) & $\bO[n^2 p + mn(p + n^2)]$ \\
+    Randomized Simple Projection (RSP) & $\bO[k^2p]$ & $\bO[mkp]$ \\
     \hline
-    Adjusted Projection (AP) & $\bO[n^2p + mk_1p]$ \\
+    Augment-Decompose-Transform (ADT) & $\bO[n^2 p]$ & $\bO[mn(p + n^2)]$ \\
     \hline
-    Online Augment-Decompose-Transform (OADT) & $\bO[n^2 p + mk_3(p + k_3 n)]$ \\
+    Adjusted Projection (AP) & $\bO[n^2p]$ &  $\bO[mkp]$ \\
+    \hline
+    Randomized Adjusted Projection (RAP) & $\bO[k^2p]$ &  $\bO[mkp]$ \\
+    \hline
+    Online Augment-Decompose-Transform (OADT) & $\bO[n^2 p]$ & $\bO[mk(p + k n)]$ \\
+    \hline
+    Randomized Online Augment-Decompose-Transform (ROADT) & $\bO[k^2 p]$ & $\bO[mk(p + k n)]$ \\
     \hline
@@
   \caption{
-    Comparison of computational complexity for the three methods.
+    Comparison of computational complexity for different methods.
     Here $p$ is the number of SNPs,
     $n$ is the number of individuals in the reference group,
-    and $m$ is the number of individuals in the  study group.
-    For projection, $k_1$ is the number of PCs needed
-    For OADT, $k_3$ is the number of PCs calculated by online SVD, which is usually much less than $n$.}
+    $m$ is the number of individuals in the  study group,
+    and $k$ is the number of PCs calculated by the corresponding method.
+  }
   \label{tbl:cplx}
@@
     \hline
-    Method & Reference Runtimes & Study Runtimes & Reference Center Error & Golden Standard Error \\ 
+    Method & Reference Runtimes & Study Runtimes & Golden Standard Error \\ 
     \hline
-    Simple Projection (SP) &  \\
+    SP & 411 & 425 & 1.95 \\
     \hline
-    Augment-Decompose-Transform (ADT) & $\bO[n^2 p + mn(p + n^2)]$ \\
+    RSP & 2504 & 241 & 1.95 \\
     \hline
-    Adjusted Projection (AP) & $\bO[n^2p + mk_1p]$ \\
+    ADT & 411 & 29813 & 0.00 \\
     \hline
-    Online Augment-Decompose-Transform (OADT) & $\bO[n^2 p + mk_3(p + k_3 n)]$ \\
+    AP & 411 & 482 1 & 1.12 \\
+    \hline
+    RAP & 2505 & 250 & 1.83 \\ 
+    \hline
+    OADT & 2518 & 103 & 0.40 \\
+    \hline
+    ROADT & 2518 & 103 & 1.93 \\
     \hline
@@
   \caption{
-    Comparison of computational complexity for the three methods.
-    Here $p$ is the number of SNPs,
-    $n$ is the number of individuals in the reference group,
-    and $m$ is the number of individuals in the  study group.
-    For projection, $k_1$ is the number of PCs needed
-    For OADT, $k_3$ is the number of PCs calculated by online SVD, which is usually much less than $n$.}
+    Comparison of runtimes and accuracy for different methods applied to the real-life datasets.
+    The reference individuals are from the 1000 Genomes dataset with 2492 unrelated individuals.
+    The study sample contain 1000 individuals from the UKBioBank dataset.
+    There are about 125,000 loci shared by the reference and study data.
+    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADT.
+  }
   \label{tbl:cplx}

### 2018-02-26 chain.sh
Add id feature in chain.sh sim=0

@@ -39,7 +39,8 @@ elif [ ${sim} == "0" ]; then
     refn=$4
     stud=$5
     stun=$6
-    outd=${stud}
+    outid=$7
+    outd=${refd}_${stud}_${outid}
     outn=${refn}_${stun}
     study_file=../data/${stud}/${stun}.geno
     geno_file=../data/${refd}/${refn}.geno

### 2018-03-04 doc/paper.tex
Correct table label.
Add discussion and data tables in paper

@@
 \date{}
+\author{}
+% an example of a edit
 
@@
 \subsubsection{Randomized SVD}
+
 In addition to estimating the PC scores of the study individuals,
@@
 For the details of the RSVD algorithm, see \cite{halko} and \cite{abraham}.
+
+Notice that the randomized SVD method on the training data
+only calculates the first $k$ eigenvalues.
+However, in order to use adjusted projection,
+all the eigenvalues of the training data are required.
+Thus for the missing eigenvalues,
+we predict them by regressing their log-scale values on the ranks
+and using the first $k$ eigenvalues for training the linear regression.
 
@@
 see \Cref{fig:nChg} and \Cref{fig:mChg}.
-For comparing the methods for the real-life data, see
+For comparing the methods for the real-life data, see \Cref{tbl:ukb}.
 
@@
 
-We find that AP improved the accuracy of SP without increasing the runtimes.
-Moreover, OADT improves the speed of ADT without much sacrifice of accuracy.
-In addition, randomized SVD is found to be as reliable as standard SVD in
-calculating the PC's of the reference group.
+For the real-life datasets (\Cref{tbl:ukb} and \Cref{fig:ukb}),
+we see that ADT, which is the most reliable method,
+is the slowest in study runtimes.
+By using the online SVD algorithm,
+OADT achieved the highest accuracy (excluding ADT, which is the golden standard) and the fastest runtime.
+SP is fast, also,
+but its accuracy is the lowest.
+By using the HDPCA package,
+AP reduces the error of AP
+and achieved the second highest in accuracy.
+The randomized version of SP is as accurate as the standard SP,
+which shows that the randomized SVD algorithm gives reliable results,
+and that using linear regression to predict the uncalculated reference eigenvalues is satisfactory.
+It also reduces the runtime almost by half,
+due to not calculating all the eigenvalues.
+Similar reduction in study runtime is observed for the randomized version of AP,
+but in this case the accuracy is also decreased,
+thought it is still higher than that of SP and RSP.
+Finally, the randomized OADT is disappointing,
+as its error is almost as high as SP and RSP.
+
+The pattern for accuracy is very similar when we look at the simulated data.
+(\Cref{tbl:nChg-runtimes-ref}, \Cref{tbl:nChg-runtimes-study},
+\Cref{tbl:nChg-accuracy-gold}, \Cref{tbl:nChg-accuracy-ctr},
+\Cref{tbl:mChg-runtimes-study}).
+A noticeable difference in this case is the accuracy of ROADT.
+Here ROADT's accuracy is almost identical to OADT's,
+as measured by both the golden standard error and the geographic center err.
+However, as the reference size increases,
+the accuracies of different methods become closer and closer (\Cref{fig:nChg}).
+As for the speed,
+AWD's study runtime increases so fast with the reference size
+that the study runtimes for the other methods seem to be constant in comparison (\Cref{fig:nChg}).
+On the other hand,
+the study runtimes appear to increase linearly with respect to the study size (\Cref{fig:mChg}).
+This information can be used to predict the runtime for larger study samples.
 
@@
     \hline
-    Simple Projection (SP) & $\bO[n^2p]$ & $\bO[mkp]$ \\
+    ADT & $\bO[n^2 p]$ & $\bO[mn(p + n^2)]$ \\
     \hline
-    Randomized Simple Projection (RSP) & $\bO[k^2p]$ & $\bO[mkp]$ \\
+    OADT & $\bO[n^2 p]$ & $\bO[mk(p + k n)]$ \\
     \hline
-    Augment-Decompose-Transform (ADT) & $\bO[n^2 p]$ & $\bO[mn(p + n^2)]$ \\
+    ROADT & $\bO[k^2 p]$ & $\bO[mk(p + k n)]$ \\
     \hline
-    Adjusted Projection (AP) & $\bO[n^2p]$ &  $\bO[mkp]$ \\
+    SP & $\bO[n^2p]$ & $\bO[mkp]$ \\
     \hline
-    Randomized Adjusted Projection (RAP) & $\bO[k^2p]$ &  $\bO[mkp]$ \\
+    RSP & $\bO[k^2p]$ & $\bO[mkp]$ \\
     \hline
-    Online Augment-Decompose-Transform (OADT) & $\bO[n^2 p]$ & $\bO[mk(p + k n)]$ \\
+    AP & $\bO[n^2p]$ &  $\bO[mkp]$ \\
     \hline
-    Randomized Online Augment-Decompose-Transform (ROADT) & $\bO[k^2 p]$ & $\bO[mk(p + k n)]$ \\
+    RAP & $\bO[k^2p]$ &  $\bO[mkp]$ \\
     \hline
@@
     \hline
+    ADT & 411 & 29813 & 0.00 \\
+    \hline
+    OADT & 2518 & 103 & 0.40 \\
+    \hline
+    ROADT & 2518 & 103 & 1.93 \\
+    \hline
     SP & 411 & 425 & 1.95 \\
@@
     \hline
-    ADT & 411 & 29813 & 0.00 \\
-    \hline
-    AP & 411 & 482 1 & 1.12 \\
+    AP & 411 & 482 & 1.12 \\
     \hline
     RAP & 2505 & 250 & 1.83 \\ 
-    \hline
-    OADT & 2518 & 103 & 0.40 \\
-    \hline
-    ROADT & 2518 & 103 & 1.93 \\
     \hline
@@
   }
-  \label{tbl:cplx}
+  \label{tbl:ukb}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|l|l|l|l|l|l|}
+    \hline
+    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
+    \hline
+    ADT        &  68  & 1263 & 2871 & 5720 & 9737 \\
+    OADT       &  72  & 18   & 19   & 22   & 24   \\
+    ROADT      & 809  & 18   & 17   & 20   & 21   \\
+    SP         &  68  & 26   & 42   & 67   & 96   \\
+    RSP        & 805  & 23   & 31   & 39   & 46   \\
+    AP         &  68  & 69   & 107  & 91   & 119  \\
+    RAP        & 805  & 11   & 24   & 29   & 9    \\
+    \hline
+  \end{tabular}
+  \caption{
+    Reference runtimes for the simulated data as the reference size increases.
+    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
+    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+  }
+  \label{tbl:nChg-runtimes-ref}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|l|l|l|l|l|l|}
+    \hline
+    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
+    \hline
+    ADT        & 391  &  130 &  222 &  339 &  486 \\
+    OADT       & 15   &  136 &  230 &  350 &  499 \\
+    ROADT      & 15   & 1211 & 1633 & 2023 & 2422 \\
+    SP         & 16   &  130 &  222 &  339 &  486 \\
+    RSP        & 15   & 1204 & 1624 & 2012 & 2409 \\
+    AP         & 34   &  130 &  222 &  339 &  486 \\
+    RAP        & 18   & 1204 & 1624 & 2012 & 2409 \\
+    \hline
+  \end{tabular}
+  \caption{
+    Study runtimes for simulated data as the reference size increases.
+    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
+    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+  }
+  \label{tbl:nChg-runtimes-study}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|l|l|l|l|l|l|}
+    \hline
+    Study Size & 1000  & 1500 & 2000 & 2500 & 3000 \\
+    \hline
+    ADT        & 0.00  & 0.00 & 0.00 & 0.00 & 0.00 \\
+    OADT       & 0.23  & 0.15 & 0.11 & 0.08 & 0.07 \\
+    ROADT      & 0.23  & 0.15 & 0.12 & 0.08 & 0.08 \\
+    SP         & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
+    RSP        & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
+    AP         & 0.46  & 0.37 & 0.31 & 0.26 & 0.22 \\
+    RAP        & 3.02  & 2.93 & 2.54 & 2.15 & 1.91 \\
+    \hline
+  \end{tabular}
+  \caption{
+    Golden standard error for the simulated data as the reference size increases.
+    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
+    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+  }
+  \label{tbl:nChg-accuracy-gold}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|l|l|l|l|l|l|}
+    \hline
+    Reference Size & 1000  & 1500 & 2000 & 2500 & 3000 \\
+    \hline
+    ADT        & 4.53  & 4.13 & 3.81 & 3.59 & 3.23 \\
+    OADT       & 4.43  & 4.06 & 3.77 & 3.55 & 3.20 \\
+    ROADT      & 4.44  & 4.06 & 3.76 & 3.55 & 3.20 \\
+    SP         & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
+    RSP        & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
+    AP         & 4.59  & 4.10 & 3.76 & 3.57 & 3.24 \\
+    RAP        & 4.23  & 3.87 & 3.55 & 3.38 & 3.07 \\
+    \hline
+  \end{tabular}
+  \caption{
+    Geographic center error for the simulated data as the reference size increases.
+    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
+    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+  }
+  \label{tbl:nChg-accuracy-ctr}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|l|l|l|l|l|l|}
+    \hline
+    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
+    \hline
+    ADT        & 507  & 758 & 1011 & 1262 & 1549 \\
+    OADT       &  73  & 108 &  156 &  189 &  282 \\
+    ROADT      &  73  & 108 &  159 &  183 &  281 \\
+    SP         &  44  &  64 &   84 &  103 &  131 \\
+    RSP        &  43  &  63 &   83 &  101 &  129 \\
+    AP         &  66  &  81 &  100 &  117 &  146 \\
+    RAP        &  22  &  15 &   15 &   21 &   24 \\
+    \hline
+  \end{tabular}
+  \caption{
+    Study runtimes for the simulated data as the study size increases.
+    The reference size is fixed to 600 and the number of SNPs is 100,000 (1000 per genealogy). 
+    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+  }
+  \label{tbl:mChg-runtimes-study}
 \end{table}

### 2018-03-04 mergePlot.R

@@ -3,13 +3,13 @@ library(scales)
 
 
 meth.n <- 7
-chg.var <- "n"
-chg.jobid <- "27483940"
+chg.var <- "m"
+chg.jobid <- "27483942"
 chg.range <- seq(1000, 3000, 500)
 chg.taskid <- c(0:4)
 chg.length <- length(chg.range)
 
-file.head <- "../flux/data/"
+file.head <- "data/"
 file.neck <- paste0(
   chg.var,
   "Chg_rand_",
@@ -19,8 +19,6 @@ file.neck <- paste0(
 )
 file.upper <- paste0(
   file.head,
-  file.neck,
-  "/",
   file.neck,
   "_100000_")
 if(chg.var == "n"){
@@ -82,7 +80,7 @@ if(chg.var == "n"){
           ylab = "PC Score Units",
           lty = lty, pch = pch, col = col, cex = cex, type = "b", lwd = lwd, col.main = col.main)
   matplot(accuracy.gold[,1], accuracy.gold[,(1:meth.n)+1],
-          main = "Deviations (Distance from TRACE)",
+          main = "Deviations (Distance from ADT)",
           xlab = xlab,
           ylab = "PC Score Units",
           lty = lty, pch = pch, col = col, cex = cex, type = "b", lwd = lwd, col.main = col.main)

### 2018-04-08 frugalpca.py
Start procrustes_diffdim

@@ -1,22 +1,33 @@
 import numpy as np
+import pandas as pd
 from scipy.linalg import orthogonal_procrustes
 from pandas_plink import read_plink
 import dask.array as da
 from dask import compute
 from chest import Chest
 import time
-import pandas as pd
+from datetime import datetime
 from subprocess import call
 import os.path
 
 
 DIM_REF = 4
-DIM_STUDY = 8
+DIM_STUDY = 20
+DIM_STUDY_HIGH = DIM_STUDY * 2
 DIM_ONLINESVD = DIM_STUDY * 2
 DIM_RANDSVD = DIM_STUDY * 4
 NITER_RANDSVD = 2
 
 np.random.seed(21)
+
+def eig_sym(XTX):
+    ssq, V = np.linalg.eigh(XTX)
+    V = np.squeeze(V)
+    ssq = np.squeeze(ssq)
+    s = np.sqrt(abs(ssq))
+    V = V.T[::-1].T
+    s = s[::-1]
+    return s, V
 
 def svd_online(U1, d1, V1, b, l):
     n, k = V1.shape
@@ -153,90 +164,175 @@ def test_online_svd_procrust():
     assert np.allclose(rho_trace, rho)
     assert np.allclose(c_trace, c)
 
+    procrustes_diffdim(PC_ref, PC_new_head)
+
     print("Passed!")
+
+def procrustes_diffdim(Y, X, n_iter_max=int(1e4), epsilon_min=1e-6):
+    epsilon = 0.0
+    n_X, p_X = X.shape
+    n_Y, p_Y = Y.shape
+    assert n_X == n_Y
+    assert p_X >= p_Y
+    if p_X == p_Y:
+        return procrustes(Y, X)
+    else:
+        Z = np.zeros((n_X, p_X - p_Y))
+        for i in range(n_iter_max):
+            W = np.hstack((Y, Z))
+
+
 
 test_online_svd_procrust()
 
-# Read data
-X = read_plink('../data/kgn/kgn_chr_all_keep_orphans_snp_hgdp_train')[2]
-X_save = read_plink('../data/kgn/kgn_chr_all_keep_orphans_snp_hgdp_train')[2]
-W = read_plink('../data/kgn/kgn_chr_all_keep_orphans_snp_hgdp_test')[2]
-X = X.astype(np.float32)
-W = np.array(W)
+# # Read data
+# if 'X' not in locals():
+#     print("Reading reference data...")
[…]

### 2018-04-08 procrustes.cpp

@@ -20,8 +20,8 @@ using namespace std;
 extern "C" void openblas_set_num_threads(int num_threads);
 
 void pm(const mat &matrix);
-int procrustes(mat &X, mat &Y, mat &Xnew, double &t, double &rho, mat &A, rowvec &b, int ps);
-double pprocrustes(mat &X, mat &Y, mat &Xnew, double &t, double &rho, mat &A, rowvec &b, int iter, double eps, int ps);
+int procrustes(mat &X, mat &Y, mat &Xnew, double &t, double &rho, mat &A, rowvec &b, int ps=0);
+double pprocrustes(mat &X, mat &Y, mat &Xnew, double &t, double &rho, mat &A, rowvec &b, int iter=10000, double eps=0.000001, int ps=0);
 
 // Use "call pm(X)" in gdb to print matrix
 void pm(const mat &matrix) {
@@ -41,7 +41,6 @@ void pm(const mat &matrix) {
 }
 
 int main(){
-  cout << "Running TRACE's procrustes functions..." << endl;
   // mat X;
   // vec b;
   // X.load("test_X.dat");
@@ -55,6 +54,9 @@ int main(){
   double rho;
   mat A;
   rowvec c;
+
+  // Test procrustes
+  cout << "Testing TRACE's procrustes functions..." << endl;
   procrustes(PC_new_head, PC_ref_fat, PC_new_head_trsfed, t, rho, A, c, 0);
   A.save("procrustes_A.dat", raw_ascii);
   c.save("procrustes_c.dat", raw_ascii);
@@ -64,46 +66,16 @@ int main(){
   fout.close();
   cout << "Done." << endl;
 
+  // Test projection procrustes
+  cout << "Testing TRACE's projection procrustes functions..." << endl;
+  pprocrustes(PC_new_head, PC_ref_fat, PC_new_head_trsfed, t, rho, A, c, 10000, 0.000001, 0);
+  A.save("pprocrustes_A.dat", raw_ascii);
+  c.save("pprocrustes_c.dat", raw_ascii);
+  fout.open("pprocrustes_rho.dat");
+  fout << rho;
+  fout.close();
+  cout << "Done." << endl;
   return 0;
-}
-
-
-int procrustes(mat &X, mat &Y, mat &Xnew, double &t, double &rho, mat &A, rowvec &b, int ps){
-	int NUM = X.n_rows;
[…]

### 2018-04-09 frugalpca.py
Finish procrustes_diffdim

@@ -11,6 +11,41 @@ from subprocess import call
 import os.path
 
 
+# def procrustes_old(data1, data2):
+#     mtx1 = np.array(data1, dtype=np.double, copy=True)
+#     mtx2 = np.array(data2, dtype=np.double, copy=True)
+
+#     if mtx1.ndim != 2 or mtx2.ndim != 2:
+#         raise ValueError("Input matrices must be two-dimensional")
+#     if mtx1.shape != mtx2.shape:
+#         raise ValueError("Input matrices must be of same shape")
+#     if mtx1.size == 0:
+#         raise ValueError("Input matrices must be >0 rows and >0 cols")
+
+#     # translate all the data to the origin
+#     mtx1_mean = np.mean(mtx1, 0)
+#     mtx1 -= mtx1_mean
+#     mtx2_mean = np.mean(mtx2, 0)
+#     mtx2 -= mtx2_mean
+
+#     # change scaling of data (in rows) such that trace(mtx*mtx') = 1
+#     norm1 = np.linalg.norm(mtx1)
+#     norm2 = np.linalg.norm(mtx2)
+#     if norm1 == 0 or norm2 == 0:
+#         raise ValueError("Input matrices must contain >1 unique points")
+#     mtx1 /= norm1
+#     mtx2 /= norm2
+
+#     # transform mtx2 to minimize disparity
+#     R, s = orthogonal_procrustes(mtx2, mtx1)
+#     # orthogonal_procrustes can only find the best transformation between normalilzed matrices
+#     s *= norm1 / norm2
+#     b = mtx1_mean - mtx2_mean @ R * s
+
+#     return R, s, b
+
+
+
 DIM_REF = 4
 DIM_STUDY = 20
 DIM_STUDY_HIGH = DIM_STUDY * 2
@@ -46,48 +81,6 @@ def svd_online(U1, d1, V1, b, l):
     V_new[k, n] = 1
     V2 = (R_Vt @ V_new).transpose()[:,:l]
     return d2, V2
-
-
-def procrustes(data1, data2):
-    '''
-    Find the best transformation that maps data2 to data1
-    This is a simple modification of scipy.spatial.procrustes.
-    For our purpose, we need to get the rotation matrix and the scaling factor,
-    but these two values are not returned in the original function.
-    '''
[…]

### 2018-04-09 procrustes.cpp

@@ -19,25 +19,28 @@ using namespace arma;
 using namespace std;
 extern "C" void openblas_set_num_threads(int num_threads);
 
+void pm(const mat &matrix, int cut_p, int cut_q);
 void pm(const mat &matrix);
 int procrustes(mat &X, mat &Y, mat &Xnew, double &t, double &rho, mat &A, rowvec &b, int ps=0);
 double pprocrustes(mat &X, mat &Y, mat &Xnew, double &t, double &rho, mat &A, rowvec &b, int iter=10000, double eps=0.000001, int ps=0);
 
-// Use "call pm(X)" in gdb to print matrix
-void pm(const mat &matrix) {
+// Use "call pm(X)" in gdb to es.cpp
+void pm(const mat &matrix, int cut_p, int cut_q) {
   cout << endl << size(matrix) << endl;
   int p = matrix.n_rows - 1;
   int q = matrix.n_cols - 1;
-  int cut = 4;
-  if(p > cut){
-    p = cut;
+  if(p > cut_p){
+    p = cut_p-1;
   }
-  if (q > cut){
-    q = cut;
+  if (q > cut_q-1){
+    q = cut_q;
   }
   mat submat = matrix.submat(0,0,p,q);
   submat.print();
   cout << endl;
+}
+void pm(const mat &matrix){
+  pm(matrix, 5, 5);
 }
 
 int main(){
@@ -45,8 +48,10 @@ int main(){
   // vec b;
   // X.load("test_X.dat");
   // b.load("test_b.dat");
+  mat PC_ref;
   mat PC_ref_fat;
   mat PC_new_head;
+  PC_ref.load("test_PC_ref.dat");
   PC_ref_fat.load("test_PC_ref_fat.dat");
   PC_new_head.load("test_PC_new_head.dat");
   mat PC_new_head_trsfed;
@@ -68,7 +73,7 @@ int main(){
 
   // Test projection procrustes
   cout << "Testing TRACE's projection procrustes functions..." << endl;
-  pprocrustes(PC_new_head, PC_ref_fat, PC_new_head_trsfed, t, rho, A, c, 10000, 0.000001, 0);
+  pprocrustes(PC_new_head, PC_ref, PC_new_head_trsfed, t, rho, A, c, 10000, 0.000001, 0);
   A.save("pprocrustes_A.dat", raw_ascii);
   c.save("pprocrustes_c.dat", raw_ascii);
   fout.open("pprocrustes_rho.dat");

### 2018-04-09 frugalpca.py
Finish procrustes_diffdim

@@ -219,154 +219,152 @@ def procrustes_diffdim(Y_mat, X_mat, n_iter_max=int(1e4), epsilon_min=1e-6):
 
 test_online_svd_procrust()
 
-# # Read data
-# if 'X' not in locals():
-#     print("Reading reference data...")
-#     X = read_plink('../data/kgn/kgn_chr_all_keep_orphans_snp_hgdp_biallelic_a2allele_train')[2]
-#     X = X.astype(np.float32)
-#     X = X.compute()
-#     p_ref, n_ref = X.shape
-#     print("Done.")
+# Read data
+if 'X' not in locals():
+    print("Reading reference data...")
+    X = read_plink('../data/kgn/kgn_chr_all_keep_orphans_snp_hgdp_biallelic_a2allele_train')[2]
+    X = X.astype(np.float32)
+    X = X.compute()
+    p_ref, n_ref = X.shape
+    print("Done.")
 
-#     # Center and nomralize reference data
-#     print("Centering and normalizing reference data...")
-#     X_mean = np.nanmean(X, axis = 1).reshape((-1, 1))
-#     X_std = np.nanstd(X, axis = 1).reshape((-1,1))
-#     X_std[X_std == 0] = 1
-#     X -= X_mean
-#     X /= X_std
-#     print("Done")
+    # Center and nomralize reference data
+    print("Centering and normalizing reference data...")
+    X_mean = np.nanmean(X, axis = 1).reshape((-1, 1))
+    X_std = np.nanstd(X, axis = 1).reshape((-1,1))
+    X_std[X_std == 0] = 1
+    X -= X_mean
+    X /= X_std
+    print("Done")
 
-# if 'W' not in locals():
-#     print("Reading study data...")
-#     W = read_plink('../data/kgn/kgn_chr_all_keep_orphans_snp_hgdp_biallelic_a2allele_test')[2]
-#     W_save = read_plink('../data/kgn/kgn_chr_all_keep_orphans_snp_hgdp_biallelic_a2allele_test')[2]
-#     W = W.compute()
-#     print("Done.")
+if 'W' not in locals():
+    print("Reading study data...")
+    W = read_plink('../data/kgn/kgn_chr_all_keep_orphans_snp_hgdp_biallelic_a2allele_test')[2]
[…]

### 2018-04-09 frugalpca.py
Start using procrustes_diffdim to calculate pc scores

@@ -240,7 +240,6 @@ if 'X' not in locals():
 if 'W' not in locals():
     print("Reading study data...")
     W = read_plink('../data/kgn/kgn_chr_all_keep_orphans_snp_hgdp_biallelic_a2allele_test')[2]
-    W_save = read_plink('../data/kgn/kgn_chr_all_keep_orphans_snp_hgdp_biallelic_a2allele_test')[2]
     W = W.compute()
     print("Done.")
 
@@ -313,8 +312,8 @@ else:
         U = X @ (V / s)
     print("Done.")
 
-# if 'XTX' not in locals():
-#     XTX = np.loadtxt('XTX.dat')
+if 'XTX' not in locals():
+    XTX = np.loadtxt('XTX.dat')
 
 # Test result close to TRACE's
 print("Testing reference PC scores are the same as TRACE's...")
@@ -349,7 +348,7 @@ print(datetime.now())
 print("Procrustes analysis...")
 print(datetime.now())
 pcs_new_head, pcs_new_tail = pcs_new[:-1, :], pcs_new[-1, :].reshape((1,-1))
-R, rho, c = procrustes_diffdim(pcs_ref_fat, pcs_new_head)
+R, rho, c = procrustes_diffdim(pcs_ref, pcs_new_head)
 pcs_new_tail_trsfed = pcs_new_tail @ R * rho + c
 pcs_new_tail_trsfed = pcs_new_tail_trsfed.flatten()[:DIM_REF]
 print(datetime.now())

### 2018-04-09 frugalpca.py
Finish svd_online + procrustes and its testing.

@@ -11,23 +11,52 @@ from subprocess import call
 import os.path
 
 
+# print("Calculating pc scores with eigen decomposition...")
+# print(datetime.now())
+# if 'XTX' not in locals():
+#     XTX = np.loadtxt('XTX.dat')
+# pcs_stu_eig = np.zeros((4, DIM_REF))
+# XTX_new = np.zeros((n_ref + 1, n_ref + 1))
+# XTX_new[:-1, :-1] = XTX
+# for i in range(4):
+#     # print("Calculating XTX_new...")
+#     # print(datetime.now())
+#     b = W[:,i]
+#     bX = b @ X
+#     bb = np.sum(b**2)
+#     XTX_new[-1, :-1] = bX
+#     XTX_new[:-1, -1] = bX
+#     XTX_new[-1, -1] = bb
+#     # print("Calculating s_new and V_new...")
+#     # print(datetime.now())
+#     s_new, V_new = eig_sym(XTX_new)
+#     Vs_new = V_new * s_new
+#     pcs_new = Vs_new[:, :DIM_STUDY]
+#     # print("Done.")
+#     # print(datetime.now())
+#     # print("Procrustes analysis...")
+#     # print(datetime.now())
+#     pcs_new_head, pcs_new_tail = pcs_new[:-1, :], pcs_new[-1, :].reshape((1,-1))
+#     R, rho, c = procrustes_diffdim(pcs_ref, pcs_new_head)
+#     pcs_new_tail_trsfed = pcs_new_tail @ R * rho + c
+#     pcs_stu_eig[i, :] = pcs_new_tail_trsfed.flatten()[:DIM_REF]
+# print("Done.")
+# assert np.allclose(pcs_stu_trace[:4,:dim_stu_trace], pcs_stu_eig[:4,:dim_stu_trace], 0.01, 0.05)
+
 # def procrustes_old(data1, data2):
 #     mtx1 = np.array(data1, dtype=np.double, copy=True)
 #     mtx2 = np.array(data2, dtype=np.double, copy=True)
-
 #     if mtx1.ndim != 2 or mtx2.ndim != 2:
 #         raise ValueError("Input matrices must be two-dimensional")
 #     if mtx1.shape != mtx2.shape:
 #         raise ValueError("Input matrices must be of same shape")
 #     if mtx1.size == 0:
[…]

### 2018-04-09 procrustes.cpp

@@ -1,20 +1,7 @@
 #include <iostream>
-#include <iomanip>
 #include <fstream>
-#include <string>
-#include <map>
-#include <vector>
-#include <gsl/gsl_rng.h>
-#include <gsl/gsl_randist.h>
-#include <gsl/gsl_statistics_double.h>
-#include <cmath>
-#include <algorithm>
-#include <time.h>
-#include <stdio.h>
-#include <stdlib.h>
 #define  ARMA_DONT_USE_WRAPPER
 #include "armadillo"
-#include <list>
 using namespace arma;
 using namespace std;
 extern "C" void openblas_set_num_threads(int num_threads);

### 2018-04-09 frugalpca.py
Fix typos in printing

@@ -363,7 +363,13 @@ for i in range(n_stu):
     pcs_new_tail_trsfed = pcs_new_tail @ R * rho + c
     pcs_stu_onl[i, :] = pcs_new_tail_trsfed.flatten()[:DIM_REF]
     if (i + 1) % 100 == 0:
-        print("Finished analyzing " + i + " samples.")
+        print("Finished analyzing " + str(i) + " samples.")
+print("Done.")
+print(datetime.now())
+
+print("Calculating study pc scores with simple projection...")
+print(datetime.now())
+pcs_stu_proj = X.T @ U
 print("Done.")
 print(datetime.now())
 

### 2018-04-09 frugalpca.py
Finish simple projection

@@ -363,13 +363,13 @@ for i in range(n_stu):
     pcs_new_tail_trsfed = pcs_new_tail @ R * rho + c
     pcs_stu_onl[i, :] = pcs_new_tail_trsfed.flatten()[:DIM_REF]
     if (i + 1) % 100 == 0:
-        print("Finished analyzing " + str(i) + " samples.")
+        print("Finished analyzing " + str(i+1) + " samples.")
 print("Done.")
 print(datetime.now())
 
 print("Calculating study pc scores with simple projection...")
 print(datetime.now())
-pcs_stu_proj = X.T @ U
+pcs_stu_proj = W.T @ U[:, :DIM_REF]
 print("Done.")
 print(datetime.now())
 
@@ -379,5 +379,6 @@ pcs_stu_trace = pd.read_table(pcs_stu_trace_file)
 pcs_stu_trace = np.array(pcs_stu_trace.iloc[:, 6:])
 dim_stu_trace = pcs_stu_trace.shape[1]
 assert np.allclose(pcs_stu_trace[:,:dim_stu_trace], pcs_stu_onl[:, :dim_stu_trace], 0.01, 0.05)
+print("Passed.")
 
 

### 2018-04-10 frugalpca.py
Add hdpca

@@ -5,6 +5,9 @@ from pandas_plink import read_plink
 import dask.array as da
 from dask import compute
 from chest import Chest
+import rpy2.robjects as robjects
+from rpy2.robjects.packages import importr
+from rpy2.robjects import numpy2ri
 import time
 from datetime import datetime
 from subprocess import call
@@ -79,6 +82,7 @@ DIM_STUDY_HIGH = DIM_STUDY * 2
 DIM_SVDONLINE = DIM_STUDY * 2
 DIM_SVDRAND = DIM_STUDY * 4
 NITER_SVDRAND = 2
+NP_OUTPUT_FMT = '%.4f'
 
 np.random.seed(21)
 
@@ -243,6 +247,7 @@ def procrustes_diffdim(Y_mat, X_mat, n_iter_max=int(1e4), epsilon_min=1e-6):
                 Z = Z_new
     return R, rho, c
 
+print(datetime.now())
 test_online_svd_procrust()
 
 # Read data
@@ -277,48 +282,44 @@ if 'W' not in locals():
     print("Done.")
 
 # PCA on the reference data
-if not ('s' in locals() and 'V' in locals()):
-    sV_file_all_exists = os.path.isfile('s.dat') and os.path.isfile('V.dat')
-    if sV_file_all_exists:
-        print("Reading existing s.dat and V.dat...")
-        s = np.loadtxt('s.dat')
-        V = np.loadtxt('V.dat')
-        print("Done.")
-    else:
-        print(datetime.now())
-        start_time = time.time()
-        # X = da.rechunk(X, (X.chunks[0], (X.shape[1])))
-        cache = Chest(path='cache')
+sV_file_all_exists = os.path.isfile('s.dat') and os.path.isfile('V.dat')
+if sV_file_all_exists:
+    print("Reading existing s.dat and V.dat...")
+    s = np.loadtxt('s.dat')
+    V = np.loadtxt('V.dat')
+    print("Done.")
+else:
+    print(datetime.now())
+    start_time = time.time()
+    # X = da.rechunk(X, (X.chunks[0], (X.shape[1])))
+    cache = Chest(path='cache')
 
-        # Compressed (randomized) svd
-        # print("Doing randomized SVD on training data...")
[…]

### 2018-04-10 old/rpy2_hdpca.py
Update notes

@@ -1,0 +1,32 @@
+import rpy2.robjects as robjects
+from rpy2.robjects.packages import importr
+import numpy as np
+
+# This is the R instance
+r = robjects.r
+
+# Activate conversion from numpy to R
+rpy2.robjects.numpy2ri.activate()
+
+# Load hdpca package
+importr('hdpca')
+
+# Load the hdpca example dataset into R
+r('data(Example)')
+
+# Create rpy2 object (function) that correspond to the function in R
+pc_adjust = r['pc_adjust']
+
+# Convert vectors in R into numpy arrays
+train_eval = np.array(r['train.eval'])
+test_scores = np.array(r['testscore'])
+p = int(r['p'][0])
+n = int(r['n'][0])
+
+# Run the hdpca method
+# Remember to translate . into _ and FALSE into False
+out = pc_adjust(train_eval, p, n, test_scores, method="d.gsp", n_spikes_max=20)
+out_np = np.array(out)
+print(out_np[:4])
+
+print("Done")

### 2018-04-10 plot.py

@@ -1,0 +1,14 @@
+import matplotlib.pyplot as plt
+import numpy as np
+
+pcs_stu_proj = np.loadtxt('pcs_stu_proj.dat')
+pcs_stu_hdpca = np.loadtxt('pcs_stu_hdpca.dat')
+pcs_stu_onl = np.loadtxt('pcs_stu_onl.dat')
+pcs_stu_trace = np.loadtxt('pcs_stu_trace.dat')
+
+plt.plot(pcs_stu_proj[:, 0], pcs_stu_proj[:, 1], 'o', alpha=0.3, label='projection')
+plt.plot(pcs_stu_hdpca[:, 0], pcs_stu_hdpca[:, 1], 'o', alpha=0.3, label='hdpca')
+plt.plot(pcs_stu_onl[:, 0], pcs_stu_onl[:, 1], 'o', alpha=0.3, label='online')
+plt.plot(pcs_stu_trace[:, 0], pcs_stu_trace[:, 1], 'o', alpha=0.3, label='trace')
+plt.legend()
+plt.show()

### 2018-04-21 frugalpca.py
Refactor getting/predicting ref popu

@@ -1,6 +1,7 @@
 import numpy as np
 import pandas as pd
 from sklearn.neighbors import KNeighborsClassifier
+from sklearn.cluster import KMeans
 from pandas_plink import read_plink
 import dask.array as da
 # from dask import compute
@@ -542,7 +543,6 @@ def get_popu_ref_info(X_fam, popu_ref_filename, superpopu_ref_filename):
     indiv_ref_info = pd.merge(X_fam, popu_ref_df, on = 'iid')
     return indiv_ref_info[['Population', 'Superpopulation']]
 
-# Predict superpopulations for study individuals
 def pred_popu_stu(pcs_ref, popu_ref, pcs_stu):
     logging.info('Predicting populations for study individuals...')
     knn = KNeighborsClassifier(n_neighbors=N_NEIGHBORS)
@@ -595,7 +595,7 @@ def pca(X, W_dask, popu_ref, out_pref):
     return pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl
 
 
-def run_pca(ref_pref, stu_pref, popu_ref_filename, superpopu_ref_filename, popu_col_name):
+def run_pca(ref_pref, stu_pref, popu_ref_filename, use_memmap=True):
     logging.info('Reference data: ' + ref_pref)
     logging.info('Study data: ' + stu_pref)
     logging.info('Temp dir: ' + TMP_DIR)
@@ -603,8 +603,12 @@ def run_pca(ref_pref, stu_pref, popu_ref_filename, superpopu_ref_filename, popu_
     ref_pref_commsnpsrefal, stu_pref_commsnpsrefal = intersect_ref_stu_snps(ref_pref, stu_pref)
     # Load ref
     X_dask, X_bim, X_fam = bed2dask(ref_pref_commsnpsrefal)
-    X = dask2memmap(X_dask, 'X.memmap')
-    psp_ref = get_popu_ref_info(X_fam, popu_ref_filename, superpopu_ref_filename)
+    if use_memmap:
+        X = dask2memmap(X_dask, 'X.memmap')
+    else:
+        X = X_dask.compute()
+    # popu_ref = get_popu_ref_info(X_fam, popu_ref_filename, superpopu_ref_filename)[popu_col_name]
+    popu_ref = pd.read_table(popu_ref_filename).iloc[1]
     # Load stu
     W_dask, W_bim, W_fam = bed2dask(stu_pref_commsnpsrefal)
[…]

### 2018-04-21 frugalpca_tester.py

@@ -2,18 +2,16 @@ import pandas as pd
 import frugalpca as fp
 
 ref_pref = '../data/kgn/kgn_snps_ukb'
-stu_pref = '../data/ukb/ukb_snps_kgn_1k'
-popu_ref_filename = '../data/kgn/kgn_popu.table'
-superpopu_ref_filename = '../data/kgn/kgn_superpopu.table'
-popu_col_name = 'Superpopulation'
+stu_pref = '../data/ukb/ukb_snps_kgn'
+popu_ref_filename = '../data/kgn/kgn_orphans.popu'
 
 log = fp.create_logger(stu_pref)
 
 fp.test_online_svd_procrust()
 
-# fp.run_pca(ref_pref, stu_pref, popu_ref_filename, superpopu_ref_filename, popu_col_name)
+fp.run_pca(ref_pref, stu_pref, popu_ref_filename)
 
-pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl = fp.load_pcs('pcs')
-popu_ref = pd.read_table('../data/kgn/kgn_orphans.popu')[popu_col_name]
-popu_stu_pred = fp. pred_popu_stu(pcs_ref, popu_ref, pcs_stu_onl)
-fp.plot_pcs(pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl, popu_ref, popu_stu_pred, stu_pref, marker_stu='.', alpha_stu=0.3)
+# pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl = fp.load_pcs('pcs')
+# popu_ref = pd.read_table(popu_ref_filename)[popu_col_name]
+# popu_stu_pred = fp. pred_popu_stu(pcs_ref, popu_ref, pcs_stu_onl)
+# fp.plot_pcs(pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl, popu_ref, popu_stu_pred, stu_pref, marker_stu='.', alpha_stu=0.3)

### 2018-04-21 frugalpca_tester.py
Prepare for merging

@@ -11,9 +11,9 @@ log = fp.create_logger(stu_pref)
 
 fp.test_online_svd_procrust()
 
-# fp.run_pca(ref_pref, stu_pref, popu_ref_filename, superpopu_ref_filename, popu_col_name)
+fp.run_pca(ref_pref, stu_pref, popu_ref_filename, superpopu_ref_filename, popu_col_name)
 
-pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl = fp.load_pcs('pcs')
-popu_ref = pd.read_table('../data/kgn/kgn_orphans.popu')[popu_col_name]
-popu_stu_pred = fp. pred_popu_stu(pcs_ref, popu_ref, pcs_stu_onl)
-fp.plot_pcs(pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl, popu_ref, popu_stu_pred, stu_pref, marker_stu='.', alpha_stu=0.3)
+# pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl = fp.load_pcs('pcs')
+# popu_ref = pd.read_table('../data/kgn/kgn_orphans.popu')[popu_col_name]
+# popu_stu_pred = fp. pred_popu_stu(pcs_ref, popu_ref, pcs_stu_onl)
+# fp.plot_pcs(pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl, popu_ref, popu_stu_pred, stu_pref, marker_stu='.', alpha_stu=0.3)

### 2018-04-21 frugalpca.py
Add clustering function for reference samples

@@ -192,6 +192,15 @@ import logging
 #     print('Reading existing U.dat...')
 #     U = np.loadtxt('U.dat')
 
+# def get_popu_ref_info(X_fam, popu_ref_filename, superpopu_ref_filename):
+#     popu_ref_df = pd.read_table(popu_ref_filename)
+#     popu_ref_df = popu_ref_df.rename(index=str, columns={'Individual ID' : 'iid'})
+#     superpopu_ref_df = pd.read_table(superpopu_ref_filename)
+#     superpopu_ref_dict = superpopu_ref_df.set_index('Population Code')['Super Population Code'].to_dict()
+#     popu_ref_df['Superpopulation'] = [superpopu_ref_dict[popu_this] for popu_this in popu_ref_df['Population']]
+#     indiv_ref_info = pd.merge(X_fam, popu_ref_df, on = 'iid')
+#     return indiv_ref_info[['Population', 'Superpopulation']]
+
 DIM_REF = 4
 assert DIM_REF >= 4
 DIM_STUDY = 20
@@ -463,10 +472,10 @@ def hdpca_adjust(s, p_ref, n_ref, pcs_stu_proj, hdpca_n_spike_max=HDPCA_N_SPIKE_
         sys.stdout = old_stdout
     return pcs_stu_hdpca
 
-def pca_stu(W_dask, X_mean, X_std, U, s, V, pcs_ref, out_pref, dim_ref=DIM_REF, dim_study=DIM_STUDY, dim_study_high=DIM_STUDY_HIGH, chunk_size_study=CHUNK_SIZE_STUDY):
+def pca_stu(W_large, X_mean, X_std, U, s, V, pcs_ref, out_pref, dim_ref=DIM_REF, dim_study=DIM_STUDY, dim_study_high=DIM_STUDY_HIGH, chunk_size_study=CHUNK_SIZE_STUDY):
     p_ref = len(X_mean)
     n_ref = pcs_ref.shape[0]
-    p_stu, n_stu = W_dask.shape
+    p_stu, n_stu = W_large.shape
     pcs_stu_proj = np.zeros((n_stu, dim_ref), dtype=np.float32)
     pcs_stu_hdpca = np.zeros((n_stu, dim_ref), dtype=np.float32)
     pcs_stu_onl = np.zeros((n_stu, dim_ref), dtype=np.float32)
[…]

### 2018-04-21 frugalpca_tester.py

@@ -2,14 +2,16 @@ import pandas as pd
 import frugalpca as fp
 
 ref_pref = '../data/kgn/kgn_snps_ukb'
-stu_pref = '../data/ukb/ukb_snps_kgn'
-popu_ref_filename = '../data/kgn/kgn_orphans.popu'
-
+stu_pref = '../data/ukb/ukb_snps_kgn_1k'
+popu_ref_filename = '../data/kgn/kgn_orphans.superpopu'
 log = fp.create_logger(stu_pref)
 
 fp.test_online_svd_procrust()
 
-fp.run_pca(ref_pref, stu_pref, popu_ref_filename)
+# fp.run_pca(ref_pref, stu_pref, popu_ref_filename, use_memmap=True)
+
+fp.run_pca(ref_pref, stu_pref, popu_ref_k=5, use_memmap=False)
+
 # pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl = fp.load_pcs('pcs')
 # popu_ref = pd.read_table(popu_ref_filename)[popu_col_name]
 # popu_stu_pred = fp. pred_popu_stu(pcs_ref, popu_ref, pcs_stu_onl)

### 2018-04-22 frugalpca.py
Seperate the methods for predicting PC scores and let the user choose one

@@ -23,7 +23,7 @@ from tempfile import mkdtemp
 import subprocess
 import sys
 import logging
-
+import filecmp
 
 # print(datetime.now())
 # print('Loading reference dask array into memmap...')
@@ -103,7 +103,7 @@ import logging
 #     # print(datetime.now())
 #     s_new, V_new = svd_eigcov(XTX_new)
 #     Vs_new = V_new * s_new
-#     pcs_new = Vs_new[:, :DIM_STUDY]
+#     pcs_new = Vs_new[:, :DIM_STU]
 #     # print('Done.')
 #     # print(datetime.now())
 #     # print('Procrustes analysis...')
@@ -202,37 +202,42 @@ import logging
 #     return indiv_ref_info[['Population', 'Superpopulation']]
 
 DIM_REF = 4
-assert DIM_REF >= 4
-DIM_STUDY = 20
+DIM_STU = 20
+DIM_STU_HIGH = DIM_STU * 2
+N_NEIGHBORS=5
+HDPCA_N_SPIKE_MAX = 20
+CHUNK_SIZE_STUDY = 1000
+
 PLOT_ALPHA_REF=0.05
 PLOT_ALPHA_STU=0.7
-N_NEIGHBORS=5
 
-DIM_STUDY_HIGH = DIM_STUDY * 2
-DIM_SVDONLINE = DIM_STUDY * 2
-# DIM_SVDRAND = DIM_STUDY * 4
+# DIM_SVDRAND = DIM_STU * 4
 # NITER_SVDRAND = 2
-HDPCA_N_SPIKE_MAX = 20
 NP_OUTPUT_FMT = '%.4f'
 DELIMITER = '\t'
 TMP_DIR = mkdtemp()
-CHUNK_SIZE_STUDY = 1000
 
-def create_logger(prefix):
+def create_logger(prefix, level='info'):
     log = logging.getLogger()
+    if level == 'info':
+        log_level = logging.INFO
+    elif level == 'debug':
+        log_level = logging.DEBUG
+    else:
+        assert False
     log.handlers = [] # Avoid duplicated logs in interactive modes
-    log.setLevel(logging.INFO)
+    log.setLevel(log_level)
     # create formatter and add it to the handlers
     formatter = logging.Formatter('%(asctime)s - %(message)s')
     # create file handler which logs even debug messages
     filename = prefix + '.' + str(round(time.time())) + '.log'
     fh = logging.FileHandler(filename, 'w')
[…]

### 2018-04-22 frugalpca_tester.py

@@ -4,15 +4,16 @@ import frugalpca as fp
 ref_pref = '../data/kgn/kgn_snps_ukb'
 stu_pref = '../data/ukb/ukb_snps_kgn_1k'
 popu_ref_filename = '../data/kgn/kgn_orphans.superpopu'
-log = fp.create_logger(stu_pref)
+# log = fp.create_logger(stu_pref)
 
 fp.test_online_svd_procrust()
 
-# fp.run_pca(ref_pref, stu_pref, popu_ref_filename, use_memmap=True)
+fp.run_pca(ref_pref, stu_pref, popu_ref_filename, method='ap', use_memmap=False)
 
-fp.run_pca(ref_pref, stu_pref, popu_ref_k=5, use_memmap=False)
+# fp.run_pca(ref_pref, stu_pref, popu_ref_k=5, use_memmap=False)
 
 # pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl = fp.load_pcs('pcs')
-# popu_ref = pd.read_table(popu_ref_filename)[popu_col_name]
-# popu_stu_pred = fp. pred_popu_stu(pcs_ref, popu_ref, pcs_stu_onl)
+# pcs_stu = pcs_stu_onl
+# popu_ref = pd.read_table(popu_ref_filename)
+# popu_stu_pred = fp.pred_popu_stu(pcs_ref, popu_ref, pcs_stu)
 # fp.plot_pcs(pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl, popu_ref, popu_stu_pred, stu_pref, marker_stu='.', alpha_stu=0.3)

### 2018-04-22 frugalpca.py
Add adp

@@ -23,7 +23,7 @@ from tempfile import mkdtemp
 import subprocess
 import sys
 import logging
-
+import filecmp
 
 # print(datetime.now())
 # print('Loading reference dask array into memmap...')
@@ -103,7 +103,7 @@ import logging
 #     # print(datetime.now())
 #     s_new, V_new = svd_eigcov(XTX_new)
 #     Vs_new = V_new * s_new
-#     pcs_new = Vs_new[:, :DIM_STUDY]
+#     pcs_new = Vs_new[:, :DIM_STU]
 #     # print('Done.')
 #     # print(datetime.now())
 #     # print('Procrustes analysis...')
@@ -202,37 +202,42 @@ import logging
 #     return indiv_ref_info[['Population', 'Superpopulation']]
 
 DIM_REF = 4
-assert DIM_REF >= 4
-DIM_STUDY = 20
+DIM_STU = 20
+DIM_STU_HIGH = DIM_STU * 2
+N_NEIGHBORS=5
+HDPCA_N_SPIKE_MAX = 20
+CHUNK_SIZE_STUDY = 1000
+
 PLOT_ALPHA_REF=0.05
 PLOT_ALPHA_STU=0.7
-N_NEIGHBORS=5
 
-DIM_STUDY_HIGH = DIM_STUDY * 2
-DIM_SVDONLINE = DIM_STUDY * 2
-# DIM_SVDRAND = DIM_STUDY * 4
+# DIM_SVDRAND = DIM_STU * 4
 # NITER_SVDRAND = 2
-HDPCA_N_SPIKE_MAX = 20
 NP_OUTPUT_FMT = '%.4f'
 DELIMITER = '\t'
 TMP_DIR = mkdtemp()
-CHUNK_SIZE_STUDY = 1000
 
-def create_logger(prefix):
+def create_logger(prefix=None, level='info'):
     log = logging.getLogger()
+    if level == 'info':
+        log_level = logging.INFO
+    elif level == 'debug':
+        log_level = logging.DEBUG
+    else:
+        assert False
     log.handlers = [] # Avoid duplicated logs in interactive modes
-    log.setLevel(logging.INFO)
+    log.setLevel(log_level)
     # create formatter and add it to the handlers
     formatter = logging.Formatter('%(asctime)s - %(message)s')
     # create file handler which logs even debug messages
     filename = prefix + '.' + str(round(time.time())) + '.log'
[…]

### 2018-04-22 frugalpca_tester.py

@@ -1,18 +1,21 @@
 import pandas as pd
 import frugalpca as fp
 
-ref_pref = '../data/kgn/kgn_snps_ukb'
-stu_pref = '../data/ukb/ukb_snps_kgn_1k'
+pref_ref = '../data/kgn/kgn_snps_ukb'
+pref_stu = '../data/ukb/ukb_snps_kgn_1k'
 popu_ref_filename = '../data/kgn/kgn_orphans.superpopu'
-log = fp.create_logger(stu_pref)
+log = fp.create_logger(pref_stu, 'debug')
 
 fp.test_online_svd_procrust()
 
-# fp.run_pca(ref_pref, stu_pref, popu_ref_filename, use_memmap=True)
+# fp.run_pca(pref_ref, pref_stu, popu_ref_filename, method='ap', use_memmap=False)
 
-fp.run_pca(ref_pref, stu_pref, popu_ref_k=5, use_memmap=False)
+# fp.run_pca(pref_ref, pref_stu, popu_ref_k=5, use_memmap=False)
+
+fp.cmp_pcs(pref_stu, ['sp', 'ap', 'oadp'])
 
 # pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl = fp.load_pcs('pcs')
-# popu_ref = pd.read_table(popu_ref_filename)[popu_col_name]
-# popu_stu_pred = fp. pred_popu_stu(pcs_ref, popu_ref, pcs_stu_onl)
-# fp.plot_pcs(pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl, popu_ref, popu_stu_pred, stu_pref, marker_stu='.', alpha_stu=0.3)
+# pcs_stu = pcs_stu_onl
+# popu_ref = pd.read_table(popu_ref_filename)
+# popu_stu_pred = fp.pred_popu_stu(pcs_ref, popu_ref, pcs_stu)
+# fp.plot_pcs(pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl, popu_ref, popu_stu_pred, pref_stu, marker_stu='.', alpha_stu=0.3)

### 2018-04-22 frugalpca.py
Make study method easier to read in log

@@ -206,7 +206,7 @@ DIM_STU = 20
 DIM_STU_HIGH = DIM_STU * 2
 N_NEIGHBORS=5
 HDPCA_N_SPIKE_MAX = 20
-CHUNK_SIZE_STUDY = 1000
+CHUNK_SIZE_STUDY = 5000
 
 PLOT_ALPHA_REF=0.05
 PLOT_ALPHA_STU=0.7
@@ -454,7 +454,7 @@ def pca_ref(X, dim_ref=DIM_REF, save_XTXsV=False):
     #     np.savetxt(out_pref+'_XTX.dat', XTX, fmt=NP_OUTPUT_FMT)
     #     np.savetxt(out_pref+'_s.dat', s, fmt=NP_OUTPUT_FMT)
     #     np.savetxt(out_pref+'_V.dat', V, fmt=NP_OUTPUT_FMT)
-    return s, V
+    return s, V, XTX
 
 def get_pc_loading(X, s, V, dim_stu_high=DIM_STU_HIGH, out_pref=None):
     logging.info('Calculating PC loadings...')
@@ -463,16 +463,37 @@ def get_pc_loading(X, s, V, dim_stu_high=DIM_STU_HIGH, out_pref=None):
         np.savetxt(out_pref+'_U.dat', U, fmt=NP_OUTPUT_FMT)
     return U
 
-def online_procrustes(U, s, V, b, dim_ref=DIM_REF, dim_stu=DIM_STU, dim_stu_high=DIM_STU_HIGH):
+def ref_aug_procrustes(pcs_ref, pcs_aug):
+    n_ref, p_ref = pcs_ref.shape
+    n_aug, p_aug = pcs_aug.shape
+    assert n_aug == n_ref + 1
+    assert p_aug >= p_ref
+    pcs_aug_head = pcs_aug[:-1, :] 
+    pcs_aug_tail = pcs_aug[-1, :].reshape((1,-1))
+    R, rho, c = procrustes_diffdim(pcs_ref, pcs_aug_head)
+    pcs_aug_tail_trsfed = pcs_aug_tail @ R * rho + c
+    return pcs_aug_tail_trsfed.flatten()
+
+def oadp(U, s, V, b, dim_ref=DIM_REF, dim_stu=DIM_STU, dim_stu_high=DIM_STU_HIGH):
     pcs_ref = V[:, :dim_ref] * s[:dim_ref]
[…]

### 2018-04-22 frugalpca_tester.py

@@ -4,15 +4,15 @@ import frugalpca as fp
 pref_ref = '../data/kgn/kgn_snps_ukb'
 pref_stu = '../data/ukb/ukb_snps_kgn_1k'
 popu_ref_filename = '../data/kgn/kgn_orphans.superpopu'
-log = fp.create_logger(pref_stu, 'debug')
+# log = fp.create_logger(pref_stu, 'debug')
 
 fp.test_online_svd_procrust()
 
-# fp.run_pca(pref_ref, pref_stu, popu_ref_filename, method='ap', use_memmap=False)
+fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, method='oadp', use_memmap=False)
 
-# fp.run_pca(pref_ref, pref_stu, popu_ref_k=5, use_memmap=False)
+# fp.run_pca(pref_ref, pref_stu, popu_ref_k=5, method='oadp', use_memmap=False)
 
-fp.cmp_pcs(pref_stu, ['sp', 'ap', 'oadp'])
+# fp.cmp_pcs(pref_stu, ['sp', 'ap', 'oadp'])
 
 # pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl = fp.load_pcs('pcs')
 # pcs_stu = pcs_stu_onl

### 2018-04-22 frugalpca.py
Add method name to plot's filename

@@ -507,7 +507,7 @@ def hdpca_adjust(s, p_ref, n_ref, pcs_stu_proj, hdpca_n_spike_max=HDPCA_N_SPIKE_
 def pca_stu(W_large, X_mean, X_std, method='oadp',
             U=None, s=None, V=None, XTX=None, X=None, pcs_ref=None,
             dim_ref=DIM_REF, dim_stu=DIM_STU, dim_stu_high=DIM_STU_HIGH):
-    logging.info('Calculating study PC scores...')
+    logging.info('Calculating study PC scores with ' + method + '...')
     p_ref = len(X_mean)
     n_ref = len(s)
     p_stu, n_stu = W_large.shape
@@ -559,7 +559,7 @@ def pca_stu(W_large, X_mean, X_std, method='oadp',
             assert (U is not None) and (s is not None)
             pcs_stu_chunk_proj = W.T @ U[:,:dim_stu]
             pcs_stu_chunk = hdpca_adjust(s, p_ref, n_ref, pcs_stu_chunk_proj)[:, :dim_ref]
-            pcs_stu[sample_start:sample_end, :] 
+            pcs_stu[sample_start:sample_end, :] = pcs_stu_chunk
         else:
             logging.error(method + ' is not one of sp, ap, or oadp.')
         elapse_method += time.time() - t0
@@ -700,7 +700,7 @@ def run_pca(ref_pref, stu_pref, popu_ref_filename=None, popu_ref_k=None, method=
     logging.info('Study population prediction saved to ' + stu_pref+'_pred.popu')
 
     # Plot PC scores
-    plot_pcs(pcs_ref, pcs_stu, popu_ref, popu_stu_pred, out_pref=stu_pref)
+    plot_pcs(pcs_ref, pcs_stu, popu_ref, popu_stu_pred, out_pref=stu_pref+'_'+method)
 
     # Finer-level PCA on European individuals
     # ref_indiv_is_eur = popu_ref == 'EUR'

### 2018-04-22 frugalpca_tester.py

@@ -8,11 +8,13 @@ popu_ref_filename = '../data/kgn/kgn_orphans.superpopu'
 
 fp.test_online_svd_procrust()
 
-fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, method='oadp', use_memmap=False)
+# fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, method='sp', use_memmap=False)
+# fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, method='ap', use_memmap=False)
+# fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, method='oadp', use_memmap=False)
+fp.cmp_pcs(pref_stu, ['sp', 'ap', 'oadp'])
 
-# fp.run_pca(pref_ref, pref_stu, popu_ref_k=5, method='oadp', use_memmap=False)
+# fp.run_pca(pref_ref, pref_stu, popu_ref_k=5, method='sp', use_memmap=False)
 
-# fp.cmp_pcs(pref_stu, ['sp', 'ap', 'oadp'])
 
 # pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl = fp.load_pcs('pcs')
 # pcs_stu = pcs_stu_onl

### 2018-04-22 frugalpca_tester.py
Mix reading and predicting ref popu

@@ -8,12 +8,10 @@ popu_ref_filename = '../data/kgn/kgn_orphans.superpopu'
 
 fp.test_online_svd_procrust()
 
-# fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, method='sp', use_memmap=False)
-# fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, method='ap', use_memmap=False)
-# fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, method='oadp', use_memmap=False)
+fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, method='sp', use_memmap=False)
+fp.run_pca(pref_ref, pref_stu, popu_ref_k=5, method='ap', use_memmap=False)
+fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, method='oadp', use_memmap=False)
 fp.cmp_pcs(pref_stu, ['sp', 'ap', 'oadp'])
-
-# fp.run_pca(pref_ref, pref_stu, popu_ref_k=5, method='sp', use_memmap=False)
 
 # pcs_ref, pcs_stu_proj, pcs_stu_hdpca, pcs_stu_onl = fp.load_pcs('pcs')
 # pcs_stu = pcs_stu_onl

### 2018-04-23 frugalpca.py
Add option for plotting multiple methods in the same plot with ref PC scores

@@ -201,6 +201,23 @@ import filecmp
 #     indiv_ref_info = pd.merge(X_fam, popu_ref_df, on = 'iid')
 #     return indiv_ref_info[['Population', 'Superpopulation']]
 
+# def cmp_pcs(pref, methods):
+#     logging.info('Comparing PC scores predicted by different methods...')
+#     pcs_ref, pcs_stu_list = load_pcs(pref, methods)
+#     assert len(methods) <= 4
+#     for i, mth in enumerate(methods):
+#         pcs_stu_this = pcs_stu_list[i]
+#         plt.subplot(121)
+#         plt.plot(pcs_stu_this[:,0], pcs_stu_this[:,1], markers[i], label=mth, alpha=PLOT_ALPHA_STU)
+#         plt.subplot(122)
+#         plt.plot(pcs_stu_this[:,2], pcs_stu_this[:,3], markers[i], label=mth, alpha=PLOT_ALPHA_STU)
+#     out_filename = pref + '_cmp.png'
+#     plt.legend()
+#     plt.tight_layout()
+#     plt.savefig(out_filename, dpi=300)
+#     logging.info('PC comparison plot saved to ' + out_filename)
+#     plt.close()
+
 DIM_REF = 4
 DIM_STU = 20
 DIM_STU_HIGH = DIM_STU * 2
@@ -208,8 +225,9 @@ N_NEIGHBORS=5
 HDPCA_N_SPIKE_MAX = 20
 CHUNK_SIZE_STUDY = 5000
 
-PLOT_ALPHA_REF=0.05
-PLOT_ALPHA_STU=0.7
+PLOT_ALPHA_REF=0.1
+PLOT_ALPHA_STU=0.99
+PLOT_MARKERS = ['.', '+', 'x', '*', 'd', 's']
 
 # DIM_SVDRAND = DIM_STU * 4
 # NITER_SVDRAND = 2
@@ -591,24 +609,10 @@ def load_pcs(pref, methods):
         pcs_stu_list += [pcs_stu_this]
     return pcs_ref, pcs_stu_list
 
-def cmp_pcs(pref, methods):
-    logging.info('Comparing PC scores predicted by different methods...')
-    pcs_ref, pcs_stu_list = load_pcs(pref, methods)
-    markers = ['+', 'x', '.', '*', 's']
-    assert len(methods) <= 4
[…]

### 2018-04-23 doc/presentation_labmeeting_20180420/main.tex
Add lab meeting presentation 20180420

% Copyright 2004 by Till Tantau <tantau@users.sourceforge.net>.
%
% In principle, this file can be redistributed and/or modified under
% the terms of the GNU Public License, version 2.
%
% However, this file is supposed to be a template to be modified
% for your own needs. For this reason, if you use this file as a
% template and not specifically distribute it as part of a another
% package/program, I grant the extra permission to freely copy and
% modify this file as you see fit and even to delete this copyright
% notice.

\documentclass{beamer}

\newcommand{\bO}{\mathcal{O}}

% There are many different themes available for Beamer. A comprehensive
% list with examples is given here:
% http://deic.uab.es/~iblanes/beamer_gallery/index_by_theme.html
% You can uncomment the themes below if you would like to use a different
% one:
%\usetheme{AnnArbor}
%\usetheme{Antibes}
%\usetheme{Bergen}
%\usetheme{Berkeley}
%\usetheme{Berlin}
%\usetheme{Boadilla}
%\usetheme{boxes}
%\usetheme{CambridgeUS}
%\usetheme{Copenhagen}
%\usetheme{Darmstadt}
%\usetheme{default}
%\usetheme{Frankfurt}
%\usetheme{Goettingen}
%\usetheme{Hannover}
%\usetheme{Ilmenau}
%\usetheme{JuanLesPins}
%\usetheme{Luebeck}
\usetheme{Madrid}
%\usetheme{Malmoe}
%\usetheme{Marburg}
%\usetheme{Montpellier}
%\usetheme{PaloAlto}
%\usetheme{Pittsburgh}
%\usetheme{Rochester}
%\usetheme{Singapore}
%\usetheme{Szeged}
%\usetheme{Warsaw}

\title{Computational Methods for Principle Component Analysis}

% A subtitle is optional and this may be deleted
% \subtitle{Optional Subtitle}

\author{David Zhang}
% - Give the names in the same order as the appear in the paper.
% - Use the \inst{?} command only if the authors have different
%   affiliation.

\date{20 April 2018}
% - Either use conference name or its abbreviation.
% - Not really informative to the audience, more for people (including
%   yourself) who are reading the slides online

% \subject{Theoretical Computer Science}
% This is only inserted into the PDF information catalog. Can be left
% out.

% If you have a file called "university-logo-filename.xxx", where xxx
% is a graphic format that can be processed by latex or pdflatex,
% resp., then you can add a logo as follows:

% \pgfdeclareimage[height=0.5cm]{university-logo}{university-logo-filename}
% \logo{\pgfuseimage{university-logo}}

% Let's get started
\begin{document}

\begin{frame}
  \titlepage
\end{frame}

% Section and subsections will appear in the presentation overview
% and table of contents.
\section{First Main Section}

\subsection{First Subsection}

\begin{frame}{Recap of Principle Component Analysis (PCA)}
\begin{itemize}
\item PC scores capture ancestral history well
\item PCA by singular value decomposition (SVD)
    \[
        \underset{p \times n}{X} = \underset{p \times n}{U} \; \underset{n \times n}{D} \; \underset{n \times n}{V^T}
    \]
    \begin{itemize}
        \item $p$: \# Features (SNPs)
        \item $n$: \# Individuals
        \item $U$: PC loadings (direction vectors)
        \item $D$: Diagonal matrix of PC standard deviations (descending)
        \item $V$: PC scores (what we are the most interested in)
    \end{itemize}
\item Can also use eigen decomposition
    \[
        X^T X = VD^2V^T, \quad X X^T = U D^2 U^T
    \]
\item When applied to all the study samples,
    the PC scores capture the population stratification (PS) well
\end{itemize}
\end{frame}

\begin{frame}{Predicting PC scores}
\begin{itemize}
\item Instead of applying PCA to all the study samples
    \begin{itemize}
    \item apply PCA to reference samples
    \item use reference PC scores to predict each study sample's PC scores
    \end{itemize}
\item Methods
    \begin{itemize}
    \item Simple Projection (SP)
    \item Adjusted Projection (AP)
        \cite{dey2016asymptotic}
    \item Augment, Decompose, Procrustes (ADP) 
        \cite{wang2015improved}
    \item Online ADP (OADP)
        \cite{brand2002incremental}
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Simple Projection (SP)}
\begin{itemize}
\item Algorithm
    \begin{enumerate}
    \item SVD on reference samples
        \[
            X_{ref} = U_{ref} D_{ref} V_{ref}^T
        \]
    \item Project the new study sample to the reference PC loadings
        \[
           v_{new} = U_{ref}^T \; x_{new}
        \]
    \end{enumerate}
\item Pro: Fast and simple to implement $\bO[mkp]$
    \begin{itemize}
    \item $m$: \# study samples
    \item $k$: \# PCs needed
    \item $p$: \# SNPs 
    \end{itemize}
\item Con: PC scores shrink toward zero when $p >> n$
\end{itemize}
\end{frame}

\begin{frame}{Adjusted Projection (AP)}
\begin{itemize}
\item Algorithm
    \begin{enumerate}
    \item SVD on reference samples
        \[
            X_{ref} = U_{ref} D_{ref} V_{ref}^T
        \]
    \item Project the new study sample to the reference PC loadings
        \[
           v_{new} = U_{ref}^T \; x_{new}
        \]
    \item \textcolor{red}{Estimate the magnitude and angle of the shrinkage by using $D_{ref}^2$}
    \item \textcolor{red}{Adjust $v_{new}$ for this shrinkage}
    \end{enumerate}
\item Computational complexity is the same as SP $\bO[mkp]$
\item Empirical runtime slightly higher than SP
\item Software: HDPCA (R package)
\end{itemize}
\end{frame}

\begin{frame}{Augment, Decompose, Procrustes (ADP)}
\begin{enumerate}
\item Calculate the reference covariance matrix $X_{ref}^T X_{ref}$
\item Eigendecomposition on the reference covariance matrix:
\[
    X_{ref}^T X_{ref} = V_{ref} D_{ref}^2 V_{ref}^T
\]
\item Augment the reference samples by appending the new study sample: $X_{aug} = (X_{ref}, x_{new})$
\item Find the augmented covariance matrix $X_{aug}^T X_{aug}$
\item Eigendecomposition on the augmented covariance matrix
\[
    X_{aug}^T X_{aug} = V_{aug} D_{aug}^2 V_{aug}^T
\]
\item Divide the augmented PC scores into the reference part and the new part: $V_{aug}^T = (V_{aug, ref}^T, v_{aug, new}^T)$
\item Find a linear transformation $f(X) = \rho XA + b$ that minimizes the difference between $f(V_{aug,ref})$ and $V_{ref}$ (Procrustes analysis)
\item Apply the same transformation to $v_{aug,new}$ to get $f(v_{aug,new})$
\end{enumerate}
\end{frame}

\begin{frame}{Augment, Decompose, Procrustes (ADP)}
    \begin{itemize}
    \item Pro: Very accurate (The golden standard)
    \item Con: Very slow when reference size is large $\bO[m(np+n^3)]$
        \begin{itemize}
        \item $m$: \# study samples
        \item $n$: \# reference samples
        \item $p$: \# features
        \end{itemize}
    \item Software: TRACE (written in C++) 
    \end{itemize}
\end{frame}

\begin{frame}{Online Augment, Decompose, Procrustes (OADP)}
\begin{itemize}
\item Algorithm
    \begin{enumerate}
    \item \textcolor{red}{SVD on the reference samples}
        \[
            X_{ref} = U_{ref} D_{ref} V_{ref}^T
        \]
    \item \textcolor{red}{Update $V_{ref}$ into $V_{aug}$ by using $U_{ref}$, $D_{ref}$, and $x_{new}$}
    \item Divide the augmented PC scores
        into the reference part and the new part:
        $V_{aug}^T = (V_{aug, ref}^T, v_{aug, new}^T)$
    \item Find a linear transformation
        $f(X) = \rho XA + b$ 
        that minimizes the difference between
        $f(V_{aug,ref})$ and $V_{ref}$
        (Procrustes analysis)
    \item Apply the same transformation to $v_{aug,new}$
        to get $f(v_{aug,new})$
    \end{enumerate}
\item Speed up ADP by avoiding repetitive calculations
    ($\bO[mk(p+kn)])$
    \begin{itemize}
    \item $m$: \# study samples
    \item $k$: \# PCs needed
    \item $p$: \# features
    \end{itemize}
\item PC scores are almost identical to ADP's
\end{itemize}
\end{frame}

\begin{frame}{Application to Real Data}
\begin{itemize}
    \item Reference samples: 1000 Genomes Project
        \begin{itemize}
        \item 2492 individuals from 26 populations in Africa, America, East Asia, South Asia, and Europe
        \item 84M SNPs
        \item 627K SNPS left after intersecting with the Human Genome Diversity Project (HGDP)
        \end{itemize}
    \item Study samples: UK Biobank
        \begin{itemize}
        \item 488K individuals studied at 23 centers in UK
        \item 148K SNPs
        \end{itemize}
    \item \# Shared SNPs between reference and study samples: 18k
\end{itemize}
\end{frame}

\begin{frame}{1000 Genomes and UK Biobank (1000 Global Individuals)}
\includegraphics[height=0.9\textheight]{pcs_save.png}
\end{frame}

\begin{frame}{1000 Genomes and UK Biobank (Global)}
\includegraphics[height=0.9\textheight]{ukb_snps_kgn_1k.png}
\end{frame}

\begin{frame}{1000 Genomes and UK Biobank (Predicted Europeans)}
\includegraphics[height=0.9\textheight]{ukb_snps_kgn_1k_eur.png}
\end{frame}

\begin{frame}{1000 Genomes and UK Biobank (488K Global Individuals)}
\includegraphics[height=0.9\textheight]{ukb_snps_kgn.png}
\end{frame}

\begin{frame}{Comparison of Efficiency}
\begin{table} 
  \centering
  \begin{tabular}{|l|l|l|l|}
    \hline
    & Reference & Study & UK Biobank \\
    Method & Complexity & Complexity & Study Runtime \\ 
    \hline
    SP & $\bO[n^2p]$ & $\bO[mkp]$ & 22 seconds \\
    \hline
    AP & $\bO[n^2p]$ &  $\bO[mkp]$ & 2.4 hours \\
    \hline
    ADP & $\bO[n^2 p]$ & $\bO[mn(p + n^2)]$ & 76 days* \\
    \hline
    OADP & $\bO[n^2 p]$ & $\bO[mk(p + k n)]$ & 4.8 hours \\
    \hline
  \end{tabular}
\end{table}

*Estimated by using the record that 1000 study samples took 3.75 hours.
\end{frame}

\begin{frame}{References}
    \bibliographystyle{apalike}
    \bibliography{main}
\end{frame}

\end{document}

### 2018-04-23 frugalpca.py
Delete commented old codes

@@ -25,199 +25,6 @@ import sys
 import logging
 import filecmp
 
-# print(datetime.now())
-# print('Loading reference dask array into memmap...')
-# W_memmap_filename = os.path.join(TMP_DIR, 'W_memmap.dat')
-# W = np.memmap(W_memmap_filename, dtype=np.float32, mode='w+', shape=W_dask.shape)
-# # W[:] = W_dask
-# W = np.zeros(dtype=np.float32, shape=W_dask.shape)
-# for i in range(len(W_dask.chunks[0])):
-#     start = sum(W_dask.chunks[0][:i])
-#     end = sum(W_dask.chunks[0][:i+1])
-#     W[start:end, :] = W_dask[start:end, :]
-# TODO: Add checking for ind-major vs snp-major
-
-
-# print('Sorting ref snps by chrom and pos...')
-# print(datetime.now())
-# X_bim_full['chrom'] = X_bim_full['chrom'].astype(np.int64)
-# W_bim_full['chrom'] = W_bim_full['chrom'].astype(np.int64)
-# # X_bim_full = X_bim_full.sort_values(by=['chrom', 'pos'])
-# print('Sorting stu snps by chrom and pos...')
-# print(datetime.now())
-# # W_bim_full = W_bim_full.sort_values(by=['chrom', 'pos'])
-# print('Intersecting snps...')
-# print(datetime.now())
-# snp_intersect = np.intersect1d(X_bim_full['snp'], W_bim_full['snp'], assume_unique=True)
-# print('Filtering reference snps...')
-# print(datetime.now())
-# X_snp_isshared = np.isin(X_bim_full['snp'], snp_intersect, assume_unique=True)
-# print('Filtering study snps...')
-# print(datetime.now())
-# W_snp_isshared = np.isin(W_bim_full['snp'], snp_intersect, assume_unique=True)
-# print('Creating filtered reference set...')
-# print(datetime.now())
-# X = X_full[X_snp_isshared].compute()
-# X_bim = X_bim_full[X_snp_isshared]
-# print('Creating filtered study set...')
-# print(datetime.now())
-# W_dask = W_full[W_snp_isshared]
-# W_memmap_filename = os.path.join(mkdtemp(), 'W_memmap.dat')
[…]

### 2018-04-29 frugalpca.py
Add procrustes similarity score and trace result loader

@@ -30,8 +30,10 @@ DIM_REF = 4
 DIM_STU = 20
 DIM_STU_HIGH = DIM_STU * 2
 N_NEIGHBORS=5
-HDPCA_N_SPIKES_MAX = 20
+HDPCA_N_SPIKES_MAX = 18
 CHUNK_SIZE_STUDY = 5000
+PROCRUSTES_NITER_MAX = 10000
+PROCRUSTES_EPSILON_MIN = 1e-6
 
 PLOT_ALPHA_REF=0.1
 PLOT_ALPHA_STU=0.99
@@ -42,15 +44,6 @@ PLOT_MARKERS = ['.', '+', 'x', '*', 'd', 's']
 NP_OUTPUT_FMT = '%.4f'
 DELIMITER = '\t'
 TMP_DIR = mkdtemp()
-
-# def hdpca_adjust(s, p_ref, n_ref, pcs_stu_proj, hdpca_n_spikes_max=HDPCA_N_SPIKES_MAX):
-#     # Run hdpca but suppress output to stdout
-#     with open(os.devnull, 'w') as devnull:
-#         old_stdout = sys.stdout
-#         sys.stdout = devnull
-#         pcs_stu_hdpca = np.array(pc_adjust(s**2, p_ref, n_ref, pcs_stu_proj, n_spikes_max=hdpca_n_spikes_max))
-#         sys.stdout = old_stdout
-#     return pcs_stu_hdpca
 
 def create_logger(prefix=None, level='info'):
     log = logging.getLogger()
@@ -113,92 +106,7 @@ def svd_online(U1, d1, V1, b, l=None):
     V2 = (R_Vt @ V_new).transpose()[:,:l]
     return d2, V2
 
-def test_online_svd_procrust():
-    np.random.seed(21)
-    # def test_svd_online():
-    logging.debug('Testing test_svd_online...')
-
-    # # For debugging only
-    # # For comparing with the R script written by Shawn
-    # X = np.loadtxt('test_X.dat')
-    # b = np.loadtxt('test_b.dat').reshape((-1,1))
-    # p, n = X.shape
-
-    # Generate testing matrices
-    p = 1000
-    n = 200
-    X = np.random.normal(size = p * n).reshape((p,n))
-    b = np.random.normal(size = p).reshape((p,1))
-    np.savetxt('test_X.dat', X)
-    np.savetxt('test_b.dat', b)
-
-    # Center reference data
[…]

### 2018-04-29 frugalpca_tester.py

@@ -1,17 +1,136 @@
 import pandas as pd
 import frugalpca as fp
 
-pref_ref = '../data/kgn/kgn_snps_ukb'
-pref_stu = '../data/ukb/ukb_snps_kgn_1k'
-popu_ref_filename = '../data/kgn/kgn_orphans.superpopu'
-# log = fp.create_logger(pref_stu, 'debug')
-dim_ref = 4
 
-fp.test_online_svd_procrust()
-pcs_ref, pcs_stu_ap, popu_ref, popu_stu_pred_ap =  fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, dim_ref=dim_ref, method='ap', use_memmap=False)
-pcs_ref, pcs_stu_sp, popu_ref, popu_stu_pred_sp =  fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, dim_ref=dim_ref, method='sp', use_memmap=False)
-pcs_ref, pcs_stu_oadp, popu_ref, popu_stu_pred_oadp =  fp.run_pca(pref_ref, pref_stu, popu_ref_filename = popu_ref_filename, dim_ref=dim_ref, method='oadp', use_memmap=False)
-pcs_stu_list = [pcs_stu_sp, pcs_stu_ap, pcs_stu_oadp]
-popu_stu_list = [popu_stu_pred_oadp]*3
-method_list = ['sp', 'adp', 'oadp']
-fp.plot_pcs(pcs_ref, pcs_stu_list, popu_ref, popu_stu_list, method_list, out_pref=pref_stu)
+def test_online_svd_procrust():
+    np.random.seed(21)
+    # def test_svd_online():
+    print('Testing test_svd_online...')
+
+    # # For debugging only
+    # # For comparing with the R script written by Shawn
+    # X = np.loadtxt('test_X.dat')
+    # b = np.loadtxt('test_b.dat').reshape((-1,1))
+    # p, n = X.shape
+
+    # Generate testing matrices
+    p = 1000
+    n = 200
+    X = np.random.normal(size = p * n).reshape((p,n))
+    b = np.random.normal(size = p).reshape((p,1))
+    np.savetxt('test_X.dat', X)
+    np.savetxt('test_b.dat', b)
+
+    # Center reference data
+    X_mean = np.mean(X, axis = 1).reshape((p,1))
+    X -= X_mean
[…]

### 2018-04-29 frugalpca.py
Save commsnps and commrefal files

@@ -187,7 +187,7 @@ def intersect_ref_stu_snps(ref_pref, stu_pref):
         return ref_pref, stu_pref
     else:
         logging.info('Intersecting SNPs in reference and study samples...')
-        bashout = subprocess.run(['bash', 'intersect_bed.sh', ref_pref, stu_pref, TMP_DIR], stdout=subprocess.PIPE)
+        bashout = subprocess.run(['bash', 'intersect_bed.sh', ref_pref, stu_pref], stdout=subprocess.PIPE)
         ref_pref_commsnpsrefal, stu_pref_commsnpsrefal = bashout.stdout.decode('utf-8').split('\n')[-3:-1]
         assert len(ref_pref_commsnpsrefal) > 0
         assert len(stu_pref_commsnpsrefal) > 0

### 2018-04-29 frugalpca_tester.py

@@ -1,5 +1,7 @@
 import pandas as pd
+import numpy as np
 import frugalpca as fp
+import subprocess
 
 
 def test_online_svd_procrust():
@@ -48,7 +50,7 @@ def test_online_svd_procrust():
     U1 = U[:, :svd_online_dim]
     d1 = d[:svd_online_dim]
     V1 = V[:, :svd_online_dim]
-    d2, PC_new = svd_online(U1, d1, V1, b, PC_new_dim)
+    d2, PC_new = fp.svd_online(U1, d1, V1, b, PC_new_dim)
 
     # Test if the result is close enough
     trueAns = np.linalg.svd(np.concatenate((X,b),axis=1))[2].transpose()[:,:PC_new_dim]
@@ -66,7 +68,7 @@ def test_online_svd_procrust():
     np.savetxt('test_PC_ref_fat.dat', PC_ref_fat)
     np.savetxt('test_PC_new_head.dat', PC_new_head)
     # Test procrustes with the same dimension
-    R, rho, c = procrustes(PC_ref_fat, PC_new_head)
+    R, rho, c = fp.procrustes(PC_ref_fat, PC_new_head)
     # PC_new_tail_trsfed = PC_new_tail @ R * rho + c
     # PC_new_tail_trsfed = PC_new_tail_trsfed.flatten()[:PC_ref_dim]
     subprocess.run(['make', 'procrustes.o'], stdout=subprocess.PIPE)
@@ -78,7 +80,7 @@ def test_online_svd_procrust():
     assert np.allclose(rho_trace, rho)
     assert np.allclose(c_trace, c)
     # Test procrustes with different dimensions
-    R_diffdim, rho_diffdim, c_diffdim = procrustes_diffdim(PC_ref, PC_new_head)
+    R_diffdim, rho_diffdim, c_diffdim = fp.procrustes_diffdim(PC_ref, PC_new_head)
     R_diffdim_trace = np.loadtxt('pprocrustes_A.dat')
     rho_diffdim_trace = np.loadtxt('pprocrustes_rho.dat')
     c_diffdim_trace = np.loadtxt('pprocrustes_c.dat')
@@ -89,11 +91,13 @@ def test_online_svd_procrust():
 
 
 def test_pca():
+    # pref_ref = '../data/kgn/kgn_bial'
[…]

### 2018-04-29 intersect_bed.sh

@@ -2,15 +2,18 @@
 
 ref=$1
 stu=$2
-tmpdir=$3
+
+pref_ref_commsnps=${ref}_snps_comm
+pref_stu_commsnps=${stu}_snps_comm
+pref_stu_commsnps_commrefal=${pref_stu_commsnps}_refal_comm
 
 cut -f2 ${ref}.bim > ${ref}.rs
 cut -f2 ${stu}.bim > ${stu}.rs
-plink --noweb --bfile ${ref} --extract ${stu}.rs --snps-only --out ${tmpdir}/ref_commsnpsrefal --make-bed
-plink --noweb --bfile ${stu} --extract ${ref}.rs --snps-only --out ${tmpdir}/stu_commsnps --make-bed
-cut -f2,5 ${tmpdir}/ref_commsnpsrefal.bim > ${tmpdir}/ref_commsnpsrefal.refal
-plink --noweb --bfile ${tmpdir}/stu_commsnps --reference-allele ${tmpdir}/ref_commsnpsrefal.refal --out ${tmpdir}/stu_commsnpsrefal --make-bed
-cmp --silent ${tmpdir}/ref_commsnpsrefal.bim ${tmpdir}/stu_commsnpsrefal.bim || {>&2 echo "Error: Output bim files not identical"; echo; echo; exit 1; } # The last two echos are placeholders for output file names
+plink --noweb --bfile ${ref} --extract ${stu}.rs --snps-only --out ${pref_ref_commsnps} --make-bed
+plink --noweb --bfile ${stu} --extract ${ref}.rs --snps-only --out ${pref_stu_commsnps} --make-bed
+cut -f2,5 ${pref_ref_commsnps}.bim > ${pref_ref_commsnps}.refal
+plink --noweb --bfile ${pref_stu_commsnps} --reference-allele ${pref_ref_commsnps}.refal --out ${pref_stu_commsnps_commrefal} --make-bed
+cmp --silent ${pref_ref_commsnps}.bim ${pref_stu_commsnps_commrefal}.bim || {>&2 echo "Error: Output bim files not identical"; echo; echo; exit 1; } # The last two echos are placeholders for output file names
 echo "Intersection finished."
-echo ${tmpdir}/ref_commsnpsrefal
-echo ${tmpdir}/stu_commsnpsrefal
+echo ${pref_ref_commsnps}
+echo ${pref_stu_commsnps_commrefal}

### 2018-04-30 doc/paper.tex
Addres some of Shawn's comments in paper.tex
Updates from ShareLaTeX

@@
 \date{}
-\author{}
-% an example of a edit
 
@@
 
-Ancestral history can be a influential confounder for the association between
+Population stratification (PS) is an important confounder for the association  between
 genotypes and phenotypes (\cite{price}).
-Thus stratification is needed to eliminate this bias.
-Principle component analysis (PCA) has been proven to be useful in estimating
-ancestral history (\cite{reich}).
-The standard way of doing PCA is to find the principle components (PC) of the
-reference data and project the data points of the study data onto them.
+To adjust for PS, we need to accurately estimate the ancestry structure in the study samples.
+Principle component analysis (PCA) is a multivariate statistical method which finds the direction of the maximum variability.
+By aggregating information across all genetic markers, PCA can accurate infer the ancestry structure  (\cite{reich}).
+To adjust for PS, PCA can be applied to study data to calculate PC scores, which are regarded as variables of ancestry direction and can be used as covariates to adjust for.
+An alternative approach is predicting the PC scores of the study samples using reference genotyped samples with detailed ancestry information.
+This prediction-based approach not only allows to adjust for PS but also to infer ancestry memberships of study samples.
+
+The standard way of predicting PC scores is to find the PC loadings of the reference data and project the data points of the study samples onto them.
 We call this method simple projection.
-However, when the number of features are greatly larger than the size of the
-reference sample,
-the estimated PC scores by this method are known to be biased and less
-than the populational PC scores.
-One way of solving this issue is presented by \cite{wang}.
+However, when the number of features are greatly larger than the size of the reference sample,
+the predicted PC scores by this method are known to be biased toward null.
+One way of addressing this shrinkage bias is presented by \cite{wang}.
 Their solution is to combine one study individual with the reference set and
 find the PC scores of this augmented data set by singular value decomposition (SVD).
-The study individual's PC scores is then retrieved from the PC score matrix and
-mapped to the reference PC's by a Procrustes transformation.
-We call this method ``augment, decompose, and transform (ADT).
-This method has been shown to be useful in eliminating the shrinking bias of
-study PC scores, but its disadvantage is computational inefficiency.
+The PC scores of the study individuals are then retrieved from the singular vector matrices and mapped to the reference PC space by a Procrustes transformation.
+We call this method ``augment, decompose, and transform'' (ADT).
+This method has been shown to be effective in eliminating the shrinking bias of
+study PC scores.
+However since ADT needs to run SVD for each study samples, ADT is computationally expensive.
 
-Two methods have been developed to solve this dillema of accuracy verses efficiency.
-The magnitude and angle of shrinkage in simple projection has been calculated by
-\cite{dey} and the result is implemented into the High Dimensional Principle
-Component Analysis (HDPCA) package in R.
-We call this method ``adjusted projection'' (AP).
-The computational complexity of AP is significantly lower than that of ADT.
-Moreover, an online SVD algorithm has been developed to speed up SVD by calculating an approximation of the top few PC scores only.
-This is useful for our purpose since in genetic studies usually only the first few PC scores are needed. 
-We call this method ``online augment-decompose-transform'' (OADT).
+To address the limitations of simple projection approach and ADT, we propose and evaluate two alternative PC score prediction methods.
+The first approach removes the bias in the simple projection method by estimating the asymptotic bias factor, which is calculated based on Random Matrix Theory \cite{dey}.
+The second approach improves the computation efficiency of ADT by using online SVD algorithm. 
+We call the first approach ``adjusted projection'' (AP), and the second approach ``online augment-decompose-transform'' (OADT).
 
-In addition, we aim at solving a problem with finding the PC's of the reference set.
+In addition, we also address a computation problem with finding the PCs of the reference set when the reference sample sizes are large. 
 The previously mentioned four methods all need to find the SVD of the reference data first.
@@
 
-In this paper, we compare the theoretical and empirical properties of simple
-projection, augment-decompose-transform, adjusted projection, and online augment-decompose-transform. In particular, we are
-interested in their computational complexity as well as empirical speed and
-accuracy.
-Secondly, we compare using standard SVD and randomized SVD for the reference set
-and examine the difference in accuracy and runtimes.
+In this paper, we compare the properties of
+simple projection, augment-decompose-transform, adjusted projection, and online augment-decompose-transform
+in terms of computation time and accuracy.
+We also compare using standard SVD and randomized SVD for the reference set
+and examine the difference in accuracy and computation time.
 We found that HDPCA projection and OADT have both achieved the accuracy of ADT
-and the efficiency of simple projection.
-In addition, randomized SVD gives results similar to standard SVD and only
-increase the computational time for the reference data.
+and the computational efficiency of simple projection.
+In addition, randomized SVD gives results similar to standard SVD and only increase the computational time for the reference data.
 
@@
 \subsubsection{Data Format}
-As mentioned previously, there are two datasets: a reference set and a study
-set.
-They are both represented as matrices.
-A data matrix contains the genotypes for the individuals.
-Each row represents a single-nucleotide polymorphism (SNP),
-while each column represents an individual.
-Each entry of the matrix is $0$, $1$, or $2$,
-representing the number of mutations observed at the SNP.
-In our study, we only use the SNPs that are observed in both the reference and
-the study group and have the same alleles.
-Thus we have the same number of SNPs for both groups.
-The genotypes of the reference group are stored in the $p \times n$ matrix $X$,
-and those of the study group are stored in the $p \times m$ matrix $Y$.
-For genotype data, it is common for $p$ to be significantly larger than $n$.
+There are two genotype datasets, reference and study datasets, and both can be represented by matrices.
+Suppose $X$ is a $p \times n$ matrix of reference genotypes and 
+$Y$ is a $p \times m$ matrix of study sample genotypes, where 
+$p$ is the number of genetic markers, $n$ is the number of reference samples and $m$ is the number of study samples.
+Each entry of $X$ and $Y$ is the number of minor alleles (0,1,2) of single nucleotide polymorphism (SNP).
+For genotype data, it is common for $p$ to be significantly larger than either $n$ or $m$.
 Thus we assume this to be true except for the part where we discuss the randomized SVD algorithm,
@@
 \subsubsection{Principle Component Analysis}\label{pca-intro}
-  Given a multi-dimensional data, principle component analysis (PCA) looks for the directions in which the data has the greatest variance.
-These unit vectors are called the principle components (PC's).
+Given a multi-dimensional data, PCA searches for the directions in which the data has the greatest variance.
+These unit vectors are called the principle components (PCs).
 The PC's form an orthonormal basis for the sample space,
@@
 The standard way for estimating PC scores is by projecting the study
-individuals' data points to the PCs of the reference individuals.
+individuals' data points to the PC space of the reference individuals by using reference sample PC loadings.
 We call this method ``simple projection'' (SP).
@@
   ($\bO[n^3]$.)
-\item Calculate $W = Y^T (X V D_1^{-1})$, where $Y$ is the data matrix for all the study individuals, $D_1$ the first $k_1$ columns of $D$. ($\bO[mpk_1 + pnk_1]$.)
+\item Calculate PC loading matrix $L = X V_1 D_1^{-1}$,
+where $V_1$ and $D_1$ are the the first $k_1$ columns of $V$ and $D$, respectively ($\bO[k_1^2 n + knp]$)
+\item Calculate predicted PC scores $W = Y^T L$, where $Y$ is the data matrix for all the study individuals ($\bO[mpk_1]$.)
 \end{enumerate}
 
-Total computational complexity: $\bO[n^2p + n^3 + mpk_1 + pnk_1)] \approx \bO[n^2p + mk_1p]$,
+Total computational complexity: 
+\[
+    \bO[n^2p + n^3 + k_1^2 n + pnk_1 + pmk_1)] \approx \bO[n^2p + mk_1p]
+\]
 given that $k_1 \ll n \ll p$.
@@
 
-One way to solve the shrinking bias of SP is developed by \cite{wang}.
+One way to solve the shrinking bias of SP is the ADT approach developed by \cite{wang}.
 In this method, each study individual's data point is appended to the reference
@@
 which is calculated in the beginning,
-and a linear (Procrustes) map is found to transform the latter into the former
+and a linear (Procrustes) map is constructed to transform the latter into the former
 with the smallest error.
@@
 
-Although accurate, ADT is the slowest among the other three computational
-methods.
-Its computational complexity increases cubically as we increase the reference sample size to increase the accuracy of the PC scores.
+Although accurate, ADT is the slowest among the other three computational methods.
+Its computational complexity increases cubically as the reference sample size and linearly as the study sampl sizes.
 The ADT method is has been implemented in the TRACE software by \cite{wang}.
@@
 
-A remedy for ADT is to reduce the computational complexity withou much sacrifice of the accuracy.
+A remedy for ADT is to reduce the computational complexity without much sacrifice of the accuracy.
 This is achieved by the online ADT (OADT) method.
@@
 
-Here we see that the computational complexity of OADT for the study individuals increases linearly with respect to the reference sample size.
-This makes the runtimes of OADT close to those of SP.
+The computational complexity of OADT for the study individuals increases linearly with respect to the reference sample size, which makes the computation time of OADT close to those of SP.
 The closeness between the results given by OADT and ADT will be empirically shown later.
 
-A comparison of computatinal complexities of the four PCA methods are shown in \Cref{tbl:cplx}.
+A comparison of computational complexities of the four PCA methods are shown in \Cref{tbl:cplx}.
 
@@
 However, when $n$ is exceedingly large,
-the covariance matrix cannot be loaded into the memory of an average personal
-computer.
+computation and memory cost of PCA on the reference samples can be enormous.
 A randomized SVD (RSVD) algorithm has been developed to handle this issue.
-RSVD is slower than standard SVD,
-but it does not require loading the whole covariance matrix into the memory.
+When the reference sample size is small, RSVD can be slower than standard SVD,
+but it is scalable to large sample size.
 For the details of the RSVD algorithm, see \cite{halko} and \cite{abraham}.
@@
 study size). 
-Second, since ADT is believed to be the most reliable, we compare the average
+Second, since ADT is a model free approach and believed to be the most reliable, we compare the average
 Euclidean distance between the PC score estimates calculated by another method
 to those calculated by ADT. 
-This golden-standard error and the geographic error both range from zero to infinity.
-
-The comparison of the runtimes and accuracy are conducted by using both the
+The comparison of the runtimes and accuracy are also conducted by using both the
 standard SVD and randomized SVD on the training data matrix.
 
-\subsubsection{Real-Life Datasets}
+\subsubsection{UK Biobank data analysis with 1000 Genomes as reference}
 
@@
     \hline
-    Method & Reference Runtimes & Study Runtimes & Golden Standard Error \\ 
+    Method & Reference Runtimes (n=?) & Study Runtimes (m=?)& Golden Standard Error \\ 
     \hline

### 2018-04-30 frugalpca.py
Change tester settings

@@ -45,7 +45,7 @@ NP_OUTPUT_FMT = '%.4f'
 DELIMITER = '\t'
 TMP_DIR = mkdtemp()
 
-def create_logger(prefix=None, level='info'):
+def create_logger(prefix='frugalpca', level='info'):
     log = logging.getLogger()
     if level == 'info':
         log_level = logging.INFO

### 2018-04-30 frugalpca_tester.py

@@ -91,10 +91,10 @@ def test_online_svd_procrust():
 
 
 def test_pca():
-    # pref_ref = '../data/kgn/kgn_bial'
-    pref_ref = '../data/kgn/kgn_snps_ukb'
-    # pref_stu = '../data/ukb/ukb_2c'
-    pref_stu = '../data/ukb/ukb_snps_kgn_1k'
+    pref_ref = '../data/kgn/kgn_bial_orphans_snps_ukb'
+    # pref_ref = '../data/kgn/kgn_snps_ukb'
+    pref_stu = '../data/ukb/ukb_snps_kgn_2c'
+    # pref_stu = '../data/ukb/ukb_snps_kgn_1k'
     popu_ref_filename = '../data/kgn/kgn_orphans.superpopu'
     pcs_trace_ref_filename = '../data/kgn_ukb_2c/kgn_ukb_2c.RefPC.coord'
     pcs_trace_stu_filename = '../data/kgn_ukb_2c/kgn_ukb_2c.ProPC.coord'
@@ -107,33 +107,33 @@ def test_pca():
 
     assert np.allclose(pcs_stu_ap, pcs_stu_oadp, 1e-1, 5e-2)
 
-    # pcs_ref_trace = fp.load_trace(pcs_trace_ref_filename, isref=True)
-    # pcs_stu_trace = fp.load_trace(pcs_trace_stu_filename)
-    # for i in range(dim_ref):
-    #     corr = np.correlate(pcs_ref[:,i], pcs_ref_trace[:,i])
-    #     sign = np.sign(corr)
-    #     pcs_ref_trace[:,i] *= sign
-    #     pcs_stu_trace[:,i] *= sign
-    # assert np.allclose(pcs_ref, pcs_ref_trace, 1e-3, 1e-3)
-    # assert np.allclose(pcs_stu_oadp, pcs_stu_trace, 1e-1, 5e-2)
-    # assert np.allclose(pcs_stu_ap, pcs_stu_trace, 1e-1, 5e-2)
+    pcs_ref_trace = fp.load_trace(pcs_trace_ref_filename, isref=True)
+    pcs_stu_trace = fp.load_trace(pcs_trace_stu_filename)
+    for i in range(dim_ref):
+        corr = np.correlate(pcs_ref[:,i], pcs_ref_trace[:,i])
+        sign = np.sign(corr)
+        pcs_ref_trace[:,i] *= sign
+        pcs_stu_trace[:,i] *= sign
[…]

### 2018-04-30 doc/paper.tex

@@
 \usepackage{verbatim}
-\usepackage{graphicx}
+\usepackage{float}
 
@@
 The standard way of predicting PC scores is to find the PC loadings of the reference data and project the data points of the study samples onto them.
-We call this method simple projection.
+We call this method simple projection (SP).
 However, when the number of features are greatly larger than the size of the reference sample,
@@
 The PC scores of the study individuals are then retrieved from the singular vector matrices and mapped to the reference PC space by a Procrustes transformation.
-We call this method ``augment, decompose, and transform'' (ADT).
+We call this method ``augment, decompose, and Procrustes transform'' (ADP).
 This method has been shown to be effective in eliminating the shrinking bias of
 study PC scores.
-However since ADT needs to run SVD for each study samples, ADT is computationally expensive.
+However since ADP needs to run SVD for each study samples, ADP is computationally expensive.
 
-To address the limitations of simple projection approach and ADT, we propose and evaluate two alternative PC score prediction methods.
-The first approach removes the bias in the simple projection method by estimating the asymptotic bias factor, which is calculated based on Random Matrix Theory \cite{dey}.
-The second approach improves the computation efficiency of ADT by using online SVD algorithm. 
-We call the first approach ``adjusted projection'' (AP), and the second approach ``online augment-decompose-transform'' (OADT).
+To address the limitations of SP and ADP,
+we propose and evaluate two alternative PC score prediction methods.
+The first approach removes the bias in SP by estimating the asymptotic bias factor,
+which is calculated based on Random Matrix Theory \cite{dey}.
+The second approach improves the computation efficiency of ADP by using online SVD algorithm. 
+We call the first approach ``adjusted projection'' (AP), and the second approach ``online augment, decompose, and and Procrustes transform'' (OADP).
 
@@
 In this paper, we compare the properties of
-simple projection, augment-decompose-transform, adjusted projection, and online augment-decompose-transform
-in terms of computation time and accuracy.
+SP, ADP, AP, and OADP in terms of computation time and accuracy.
 We also compare using standard SVD and randomized SVD for the reference set
 and examine the difference in accuracy and computation time.
-We found that HDPCA projection and OADT have both achieved the accuracy of ADT
-and the computational efficiency of simple projection.
+We found that AP and OADP have both achieved the accuracy of ADP
+and the computational efficiency of SP.
 In addition, randomized SVD gives results similar to standard SVD and only increase the computational time for the reference data.
@@
 
-\subsubsection{Augment, Decompose, and Transform (ADT)}
+\subsubsection{Augment, Decompose, and Transform (ADP)}
 
-One way to solve the shrinking bias of SP is the ADT approach developed by \cite{wang}.
+One way to solve the shrinking bias of SP is the ADP approach developed by \cite{wang}.
 In this method, each study individual's data point is appended to the reference
@@
 
-Although accurate, ADT is the slowest among the other three computational methods.
+Although accurate, ADP is the slowest among the other three computational methods.
 Its computational complexity increases cubically as the reference sample size and linearly as the study sampl sizes.
-The ADT method is has been implemented in the TRACE software by \cite{wang}.
+The ADP method is has been implemented in the TRACE software by \cite{wang}.
 
@@
 
-\subsubsection{Online ADT (OADT)}
+\subsubsection{Online ADP (OADP)}
 
-A remedy for ADT is to reduce the computational complexity without much sacrifice of the accuracy.
-This is achieved by the online ADT (OADT) method.
-Notice that in OADT, only the top $k_2$ PC's are used.
-OADT uses the online SVD method, which calculates a close approximation of the first $k_3$ columns of $\tilde{V}$ for some $k_2 \leq k_3 \ll n$ and keeps everything else the same as in ADT, including Procrustes analysis.
+A remedy for ADP is to reduce the computational complexity without much sacrifice of the accuracy.
+This is achieved by the online ADP (OADP) method.
+Notice that in OADP, only the top $k_2$ PC's are used.
+OADP uses the online SVD method, which calculates a close approximation of the first $k_3$ columns of $\tilde{V}$ for some $k_2 \leq k_3 \ll n$ and keeps everything else the same as in ADP, including Procrustes analysis.
 The algorithm is
@@
 
-The computational complexity of OADT for the study individuals increases linearly with respect to the reference sample size, which makes the computation time of OADT close to those of SP.
-The closeness between the results given by OADT and ADT will be empirically shown later.
+The computational complexity of OADP for the study individuals increases linearly with respect to the reference sample size, which makes the computation time of OADP close to those of SP.
+The closeness between the results given by OADP and ADP will be empirically shown later.
 
@@
 training data.
-For SP, AP, ADT, and OADT, we need to find the covariance matrix of the training
+For SP, AP, ADP, and OADP, we need to find the covariance matrix of the training
 data, which has dimension $n \times n$, and conduct eigendecomosition on it.
@@
 study size). 
-Second, since ADT is a model free approach and believed to be the most reliable, we compare the average
+Second, since ADP is a model free approach and believed to be the most reliable, we compare the average
 Euclidean distance between the PC score estimates calculated by another method
-to those calculated by ADT. 
+to those calculated by ADP. 
 The comparison of the runtimes and accuracy are also conducted by using both the
@@
 For the real-life datasets (\Cref{tbl:ukb} and \Cref{fig:ukb}),
-we see that ADT, which is the most reliable method,
+we see that ADP, which is the most reliable method,
 is the slowest in study runtimes.
 By using the online SVD algorithm,
-OADT achieved the highest accuracy (excluding ADT, which is the golden standard) and the fastest runtime.
+OADP achieved the highest accuracy (excluding ADP, which is the golden standard) and the fastest runtime.
 SP is fast, also,
@@
 thought it is still higher than that of SP and RSP.
-Finally, the randomized OADT is disappointing,
+Finally, the randomized OADP is disappointing,
 as its error is almost as high as SP and RSP.
@@
 \Cref{tbl:mChg-runtimes-study}).
-A noticeable difference in this case is the accuracy of ROADT.
-Here ROADT's accuracy is almost identical to OADT's,
+A noticeable difference in this case is the accuracy of ROADP.
+Here ROADP's accuracy is almost identical to OADP's,
 as measured by both the golden standard error and the geographic center err.
@@
    \bibitem[Price et al.(2006)]{price} Price, A. L., Patterson, N. J., Plenge,
-     R. M., Weinblatt, M. E., Shadick, N. A., \& Reich, D. (2006). Principal
-     components analysis corrects for stratification in genome-wide association
-     studies. {\it Nature genetics}, 38(8), 904.
+     R. M., Weinblatt, M. E., Shadick, N. A., \& Reich, D. (2006).
+     Principal components analysis corrects for stratification in genome-wide association studies.
+     {\it Nature genetics}, 38(8), 904.
 
@@
     \hline
-    ADT & $\bO[n^2 p]$ & $\bO[mn(p + n^2)]$ \\
+    ADP & $\bO[n^2 p]$ & $\bO[mn(p + n^2)]$ \\
     \hline
-    OADT & $\bO[n^2 p]$ & $\bO[mk(p + k n)]$ \\
+    OADP & $\bO[n^2 p]$ & $\bO[mk(p + k n)]$ \\
     \hline
-    ROADT & $\bO[k^2 p]$ & $\bO[mk(p + k n)]$ \\
+    ROADP & $\bO[k^2 p]$ & $\bO[mk(p + k n)]$ \\
     \hline
@@
     \hline
-    ADT & 411 & 29813 & 0.00 \\
+    ADP & 411 & 29813 & 0.00 \\
     \hline
-    OADT & 2518 & 103 & 0.40 \\
+    OADP & 2518 & 103 & 0.40 \\
     \hline
-    ROADT & 2518 & 103 & 1.93 \\
+    ROADP & 2518 & 103 & 1.93 \\
     \hline
@@
     There are about 125,000 loci shared by the reference and study data.
-    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADT.
+    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADP.
   }
@@
     \hline
-    ADT        &  68  & 1263 & 2871 & 5720 & 9737 \\
-    OADT       &  72  & 18   & 19   & 22   & 24   \\
-    ROADT      & 809  & 18   & 17   & 20   & 21   \\
+    ADP        &  68  & 1263 & 2871 & 5720 & 9737 \\
+    OADP       &  72  & 18   & 19   & 22   & 24   \\
+    ROADP      & 809  & 18   & 17   & 20   & 21   \\
     SP         &  68  & 26   & 42   & 67   & 96   \\
@@
     \hline
-    ADT        & 391  &  130 &  222 &  339 &  486 \\
-    OADT       & 15   &  136 &  230 &  350 &  499 \\
-    ROADT      & 15   & 1211 & 1633 & 2023 & 2422 \\
+    ADP        & 391  &  130 &  222 &  339 &  486 \\
+    OADP       & 15   &  136 &  230 &  350 &  499 \\
+    ROADP      & 15   & 1211 & 1633 & 2023 & 2422 \\
     SP         & 16   &  130 &  222 &  339 &  486 \\
@@
     \hline
-    ADT        & 0.00  & 0.00 & 0.00 & 0.00 & 0.00 \\
-    OADT       & 0.23  & 0.15 & 0.11 & 0.08 & 0.07 \\
-    ROADT      & 0.23  & 0.15 & 0.12 & 0.08 & 0.08 \\
+    ADP        & 0.00  & 0.00 & 0.00 & 0.00 & 0.00 \\
+    OADP       & 0.23  & 0.15 & 0.11 & 0.08 & 0.07 \\
+    ROADP      & 0.23  & 0.15 & 0.12 & 0.08 & 0.08 \\
     SP         & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
@@
     \hline
-    ADT        & 4.53  & 4.13 & 3.81 & 3.59 & 3.23 \\
-    OADT       & 4.43  & 4.06 & 3.77 & 3.55 & 3.20 \\
-    ROADT      & 4.44  & 4.06 & 3.76 & 3.55 & 3.20 \\
+    ADP        & 4.53  & 4.13 & 3.81 & 3.59 & 3.23 \\
+    OADP       & 4.43  & 4.06 & 3.77 & 3.55 & 3.20 \\
+    ROADP      & 4.44  & 4.06 & 3.76 & 3.55 & 3.20 \\
     SP         & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
@@
     \hline
-    ADT        & 507  & 758 & 1011 & 1262 & 1549 \\
-    OADT       &  73  & 108 &  156 &  189 &  282 \\
-    ROADT      &  73  & 108 &  159 &  183 &  281 \\
+    ADP        & 507  & 758 & 1011 & 1262 & 1549 \\
+    OADP       &  73  & 108 &  156 &  189 &  282 \\
+    ROADP      &  73  & 108 &  159 &  183 &  281 \\
     SP         &  44  &  64 &   84 &  103 &  131 \\
@@
     The reference center error is the square root of the average squared distance between the study individuals and the population centers, which are the means of the PC scores of the reference individuals by the populations.
-    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADT.
+    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADP.
     The reference size is 1000 and the study size is 200.

### 2018-04-30 doc/paper.tex

@@
 Population stratification (PS) is an important confounder for the association  between
-genotypes and phenotypes (\cite{price}).
+genotypes and phenotypes (\cite{price2006principal}).
 To adjust for PS, we need to accurately estimate the ancestry structure in the study samples.
@@
 
-\begin{thebibliography}{}
-
-\bibitem[Abraham \& Inouye(2014)]{abraham} Abraham, G., \& Inouye, M. (2014). Fast principal component analysis of large-scale genome-wide data. {\it PloS one}, 9(4), e93766.
-
-\bibitem[Brand(2002)]{brand} Brand, M. (2002). Incremental singular value decomposition of uncertain data with missing values. {\it Computer Vision--ECCV 2002}, 707-720.
-
-\bibitem[Dey \& Lee(2016)]{dey} Dey, R., \& Lee, S.
-(2016).
-Asymptotic properties of Principal Component Analysis and shrinkage-bias adjustment under the Generalized Spiked Population model.
-{\it arXiv preprint arXiv:1607.08647}.
-
-\bibitem[Jolliffe(2002)]{jolliffe} Jolliffe, I. (2002). {\it Principal component analysis}. John Wiley \& Sons, Ltd.
-Chicago
-
-\bibitem[Halko et al.(2011)]{halko} Halko, N., Martinsson, P.
-G., \& Tropp, J.
-A.
-(2011).
-Finding structure with randomness: Probabilistic algorithms for constructing approximate matrix decompositions.
-{\it SIAM review}, 53(2), 217-288.
-
- \bibitem[Wang et al.(2015)]{wang} Wang, C., Zhan, X., Liang, L., Abecasis, G.
-R., \& Lin, X.
-(2015).
-Improved ancestry estimation for both genotyping and sequencing data using projection procrustes analysis and genotype imputation.
-{\it The American Journal of Human Genetics, 96}(6), 926-937.
-
- \bibitem[Reich et al.(2008)]{reich} Reich, D., Price, A. L., \& Patterson, N.
-   (2008). Principal component analysis of genetic data. {\it Nature genetics}, 40(5), 491.
-   Reich, D., Price, A., Patterson, N (2008).
-   Principle component analysis of genetic data. {\it Nature Genetics}
-
-   \bibitem[Price et al.(2006)]{price} Price, A. L., Patterson, N. J., Plenge,
-     R. M., Weinblatt, M. E., Shadick, N. A., \& Reich, D. (2006).
-     Principal components analysis corrects for stratification in genome-wide association studies.
-     {\it Nature genetics}, 38(8), 904.
-
-\end{thebibliography}
+\bibliographystyle{plain}
+\bibliography{main}
 

### 2018-05-01 doc/paper.tex
Finish addressing all Shawn's editings

@@
 \usepackage{epsfig} \usepackage[margin=1.5in]{geometry}
-\usepackage[semicolon,authoryear]{natbib} \bibliographystyle{natbib}
+\usepackage[round]{natbib}
 \usepackage{amsmath}
@@
 Population stratification (PS) is an important confounder for the association  between
-genotypes and phenotypes (\cite{price2006principal}).
+genotypes and phenotypes \citep{price2006principal}.
 To adjust for PS, we need to accurately estimate the ancestry structure in the study samples.
 Principle component analysis (PCA) is a multivariate statistical method which finds the direction of the maximum variability.
-By aggregating information across all genetic markers, PCA can accurate infer the ancestry structure  (\cite{reich}).
+By aggregating information across all genetic markers, PCA can accurate infer the ancestry structure \citep{reich2008principal}.
 To adjust for PS, PCA can be applied to study data to calculate PC scores, which are regarded as variables of ancestry direction and can be used as covariates to adjust for.
@@
 However, when the number of features are greatly larger than the size of the reference sample,
-the predicted PC scores by this method are known to be biased toward null.
-One way of addressing this shrinkage bias is presented by \cite{wang}.
+the predicted PC scores by this method are known to be biased toward null \citep{dey2016asymptotic}.
+One way of addressing this shrinkage bias is presented by \citet{wang2014ancestry, wang2015improved}.
 Their solution is to combine one study individual with the reference set and
@@
 The first approach removes the bias in SP by estimating the asymptotic bias factor,
-which is calculated based on Random Matrix Theory \cite{dey}.
-The second approach improves the computation efficiency of ADP by using online SVD algorithm. 
-We call the first approach ``adjusted projection'' (AP), and the second approach ``online augment, decompose, and and Procrustes transform'' (OADP).
+which is calculated based on Random Matrix Theory \citep{dey2016asymptotic}.
+The second approach improves the computation efficiency of ADP by using online SVD algorithm \citep{halko2011finding}. 
+We call the first approach ``adjusted projection'' (AP), and the second approach ``online augment, decompose, and Procrustes transform'' (OADP).
 
@@
 and we can ignore all but the first few PC's to reduce the dimension of the data.
-In particular, let $X=UDV^T$ be the SVD of $X$,
+In particular, let $X$ be standardized so that each row (feature) has mean equal to zero and standard deviation equal to one.
+Moreover, Let $X=UDV^T$ be the SVD of $X$,
 where $U$ is a $p \times n$ matrix whose columns are the PC's,
@@
 
-We assume that the reference data and the study data are generated from the same
 population.
-When the genotypes of all individuals of the population are given (which is very
-unlikely in global genetic studies),
+In the ideal case in which the genotypes of all individuals of the population are given
 we use the method in \Cref{pca-intro} to find the PC scores of the individuals
 we are interested in.
-In practice, we have a reference dataset and a study dataset assumed to be generated from the
+In practice, however, we have a reference dataset and a study dataset assumed to be generated from the
 same population and use the former to estimate the populational PC scores of the latter.
@@
 The standard way for estimating PC scores is by projecting the study
-individuals' data points to the PC space of the reference individuals by using reference sample PC loadings.
+individuals' data points to the PC space of the reference individuals by using reference sample PC loadings,
+which is the matrix $U$ in the eigendecomposition of the (standardized) reference sample matrix $X$.
+The PC loadings are unit vectors pointing toward the directions of the greatest variances in the $p$-dimension sample space.
 We call this method ``simple projection'' (SP).
@@
 except for decomposing the reference matrix, which is done once for all study individuals.
-However, it is known that the projection method tends to underestimates the magnitude of the PC scores when $n \ll p$,
-which we call the ``shrinking bias''. (See \cite{dey}.)
+However, it is known that the projection method tends to underestimates the magnitude of the PC scores when $n \ll p$ \citep{dey2016asymptotic},
+which we call the ``shrinking bias''.
 This limits the accuracy of SP for high dimensional data.
@@
 
-One way to solve the shrinking bias of SP is the ADP approach developed by \cite{wang}.
+One way to solve the shrinking bias of SP is the ADP approach developed by \citet{wang2014ancestry, wang2015improved}.
 In this method, each study individual's data point is appended to the reference
@@
 Although accurate, ADP is the slowest among the other three computational methods.
-Its computational complexity increases cubically as the reference sample size and linearly as the study sampl sizes.
-The ADP method is has been implemented in the TRACE software by \cite{wang}.
+Its computational complexity increases cubically with respect to the reference sample size and linearly with respect to the study sample size.
+The ADP method is has been implemented in the TRACE software \citep{wang2014ancestry, wang2015improved}.
 
@@
 magnitude and angle of the shrinkage and adjust the study individuals' PC scores accordingly.
-This method has been developed by \cite{dey} and implemented in the High
-Dimension Principal Component Analysis (HDPCA) in the R language.
+This method has been developed by \citet{dey2016asymptotic} and implemented in the High
+Dimension Principal Component Analysis (HDPCA) in the R language \citep{hdpca}.
 HDPCA reads the PC scores and variances calculated by projection and estimate the shrinkage factor to restore the unshrinked scores. 
@@
 but it is scalable to large sample size.
-For the details of the RSVD algorithm, see \cite{halko} and \cite{abraham}.
+For the details of the RSVD algorithm, see \citet{halko2011finding}.
 
@@
 
-\subsection{Data}
+\section{Data}
 
@@
 
-\bibliographystyle{plain}
-\bibliography{main}
+\bibliographystyle{plainnat}
+\bibliography{frugalpca}
 

### 2018-06-14 frugalpca.py
Parallelize pca by chunks

@@ -25,20 +25,25 @@ import subprocess
 import sys
 import logging
 import filecmp
+import multiprocessing as mp
+from joblib import Parallel, delayed
 
 DIM_REF = 4
 DIM_STU = 20
 DIM_STU_HIGH = DIM_STU * 2
 N_NEIGHBORS=5
 HDPCA_N_SPIKES_MAX = 18
-SAMPLE_CHUNK_SIZE_STU = 1000
+SAMPLE_CHUNK_SIZE_STU = 50000
 SAMPLE_SPLIT_PREF_LEN = 4
 PROCRUSTES_NITER_MAX = 10000
 PROCRUSTES_EPSILON_MIN = 1e-6
+NUM_CORES = mp.cpu_count()
 
-PLOT_ALPHA_REF=0.1
-PLOT_ALPHA_STU=0.2
+
+PLOT_ALPHA_REF=0.05
+PLOT_ALPHA_STU=0.40
 PLOT_MARKERS = ['.', '+', 'x', '*', 'd', 's']
+LOG_LEVEL = 'info'
 
 # DIM_SVDRAND = DIM_STU * 4
 # NITER_SVDRAND = 2
@@ -46,6 +51,20 @@ NP_OUTPUT_FMT = '%.4f'
 DELIMITER = '\t'
 # DIR_TMP = mkdtemp()
 DIR_TMP = 'tmp'
+
+
+        # pcs_stu[sample_start:sample_end, :] = Parallel(n_jobs=NUM_CORES)(delayed(get_pcs_stu_this)(i, W, X_mean, X_std, U, s, V, method, dim_ref, dim_stu, dim_stu_high) for i in range(W.shape[1]))
+
+        # output = mp.Queue()
+        # processes = [mp.Process(target=get_pcs_stu_this, args=(output, i, W, X_mean, X_std, U, s, V, method, dim_ref, dim_stu, dim_stu_high)) for i in range(W.shape[1])]
+        # for p in processes:
+        #     p.start()
+        # for p in processes:
+        #     p.join()
+        # results = [output.get() for p in processes]
+        # for (i, pcs_stu_this) in results:
+        #     pcs_stu[sample_start + i, :] = pcs_stu_this
+
 
 def create_logger(prefix='frugalpca', level='info'):
     log = logging.getLogger()
@@ -223,9 +242,50 @@ def read_bed(bed_filepref, bed_store='memory', dtype=np.int8):
         bed = 2 - bed
     return bed, bim, fam
 
[…]

### 2018-06-14 frugalpca_test_runner.py

@@ -1,0 +1,8 @@
+import frugalpca_tester as fpt
+
+# fpt.convert()
+# fpt.test_pca_subpopu(pref_ref, pref_stu, 'EUR', cmp_trace=True)
+
+fpt.test_pca_ggsim()
+# fpt.test_pca_5c()
+# fpt.test_pca_5c_EUR()

### 2018-06-14 frugalpca_tests.py

@@ -1,0 +1,275 @@
+import pandas as pd
+import numpy as np
+import frugalpca as fp
+import subprocess
+import os.path
+
+# def run_tests():
+#     # pref_ref = '../data/kgn/kgn_bial_orphans_snps_ukb_snpscap_ukb'
+#     # pref_stu = '../data/ukb/ukb_snpscap_kgn_bial_orphans'
+#     # pref_stu = '../data/ukb/ukb_snpscap_kgn_bial_orphans_5c'
+#     # pref_stu = '../data/ukb/old/ukb_snps_kgn' # Old, don't use except for testing plotting
+#     # cmp_trace = False
+#     # test_standardize()
+#     # test_online_svd_procrust()
+#     # test_pca(pref_ref, pref_stu, cmp_trace=cmp_trace)
+#     # test_pca_subpopu(pref_ref, pref_stu, 'EUR')
+#     # test_pca_subpopu(pref_ref, pref_stu, 'EAS')
+#     # test_pca_subpopu(pref_ref, pref_stu, 'SAS')
+#     # test_pca_subpopu(pref_ref, pref_stu, 'AMR')
+#     # test_pca_subpopu(pref_ref, pref_stu, 'AFR')
+#     # plot_results(pref_ref, pref_stu)
+
+
+def test_standardize():
+    print('Testing standardize...')
+    x = np.array([[0.0,1.0,2.0],
+                  [2,2,2],
+                  [2,3,2],
+                  [0,1,3]])
+    x_standardized_crct = np.array([[-1.22474487,0,1.22474487],
+                                    [0,0,0],
+                                    [0,0,0],
+                                    [-1,1,0]])
+    x_mean_crct = np.array([1,2,2,0.5]).reshape((-1,1))
+    x_std_crct = np.array([0.81649658,1.,1.,0.5]).reshape((-1,1))
+    x_mean, x_std = fp.standardize(x, miss=3)
+    assert np.allclose(x, x_standardized_crct)
+    assert np.allclose(x_mean, x_mean_crct)
+    assert np.allclose(x_std, x_std_crct)
+    print('Passed!')
+
+    y = np.array([[ 1.,  0.,  1.],
+                  [ 0., 3.,  0.],
+                  [ 2.,  2.,  0.],
+                  [3.,  0.,  0.]])
[…]

### 2018-06-14 runTrace_ori.sh

@@ -1,0 +1,25 @@
+#!/bin/bash
+
+ver=$1
+geno_file=$2
+study_file=$3
+out_pref=$4
+
+DIM=4
+DIM_HIGH=20
+
+echo "Running TRACE..."
+
+echo -e "hostname\t`hostname`" >> ${out_pref}.info
+
+# For running trace in this directory
+# Better than putting all the info as arguments. This way running gdb will be easier
+echo "\
+GENO_FILE ${geno_file}
+STUDY_FILE ${study_file}
+OUT_PREFIX ${out_pref}
+DIM        	${DIM}
+DIM_HIGH	${DIM_HIGH}
+" > trace.conf
+
+trace -p trace.conf

### 2018-06-14 simGeno.sh

@@ -9,6 +9,7 @@ m=$5
 k=$6
 s=$7
 mig=$8
+pref=${pref}${n}
 let "nplusm = n + m"
 let "nplusmHigh = nplusm"
 # Not using unbalanced for now. Kind of buggy right now.
@@ -36,7 +37,7 @@ else
     date
     echo "Done."
     echo "Removing intermediate files..."
-    rm ${ggs_file} ${dup_file} ${geno_both_file} 
+    rm ${ggs_file} ${dup_file} ${geno_both_file}
     echo "Done."
 fi
 

### 2018-06-17 frugalpca.py
Start parallel submission

@@ -33,7 +33,7 @@ DIM_STU = 20
 DIM_STU_HIGH = DIM_STU * 2
 N_NEIGHBORS=5
 HDPCA_N_SPIKES_MAX = 18
-SAMPLE_CHUNK_SIZE_STU = 50000
+SAMPLE_CHUNK_SIZE_STU = 5000
 SAMPLE_SPLIT_PREF_LEN = 4
 PROCRUSTES_NITER_MAX = 10000
 PROCRUSTES_EPSILON_MIN = 1e-6
@@ -79,7 +79,7 @@ def create_logger(prefix='frugalpca', level='info'):
     # create formatter and add it to the handlers
     formatter = logging.Formatter('%(asctime)s - %(message)s')
     # create file handler which logs even debug messages
-    filename = prefix + '.' + str(round(time.time())) + '.log'
+    filename = os.path.dirname(prefix) + '/logs/' + os.path.basename(prefix) + '.' + str(round(time.time())) + '.log'
     fh = logging.FileHandler(filename, 'w')
     fh.setLevel(log_level)
     fh.setFormatter(formatter)
@@ -387,62 +387,6 @@ def get_pcs_stu_this(output, i, W, X_mean, X_std, U, s, V, method, dim_ref=DIM_R
     output.put((i, pcs_stu_this))
 
 
-def pca_stu(W, X_mean, X_std, method, path_tmp,
-            U=None, s=None, V=None, XTX=None, X=None, pcs_ref=None,
-            dim_ref=DIM_REF, dim_stu=DIM_STU, dim_stu_high=DIM_STU_HIGH):
-    p_ref = len(X_mean)
-    n_ref = len(s)
-    p_stu, n_stu = W.shape
-    pcs_stu = np.zeros((n_stu, dim_ref))
-
-    elapse_load = 0.0
-    elapse_standardize = 0.0
-    elapse_method = 0.0
-
-    if method == 'oadp':
-        assert (U is not None) & (s is not None) and (V is not None)
-    elif method == 'adp':
-        assert (XTX is not None) and (X is not None)
-    elif method == 'ap':
-        assert U is not None
-    elif method == 'sp':
-        assert U is not None
-    else:
-        logging.error(Method + ' is not one of sp, ap, adp, or oadp.')
-        assert False
-
-    for i in range(W.shape[1]):
-        t0 = time.time()
-        logging.debug('Extracting one row...')
[…]

### 2018-06-17 frugalpca_runner.py

@@ -1,0 +1,17 @@
+import frugalpca_tests as fpt
+import multiprocessing as mp
+from joblib import Parallel, delayed
+NUM_CORES = mp.cpu_count()
+
+print('Number of cores: ' + str(NUM_CORES))
+
+# fpt.convert()
+# fpt.test_pca_subpopu(pref_ref, pref_stu, 'EUR', cmp_trace=True)
+
+# fpt.test_pca_ggsim()
+fpt.test_pca_5c()
+# fpt.test_pca_5c_EUR()
+# Parallel(n_jobs=NUM_CORES)(delayed(fpt.convert_ggsim)(i) for i in range(5))
+# fpt.convert_ggsim(4)
+# fpt.test_pca_ggsim()
+

### 2018-06-17 frugalpca_test_runner.py

@@ -1,8 +1,0 @@
-import frugalpca_tester as fpt
-
-# fpt.convert()
-# fpt.test_pca_subpopu(pref_ref, pref_stu, 'EUR', cmp_trace=True)
-
-fpt.test_pca_ggsim()
-# fpt.test_pca_5c()
-# fpt.test_pca_5c_EUR()

### 2018-06-17 frugalpca_tests.py

@@ -164,9 +164,15 @@ def test_pca(pref_ref, pref_stu, cmp_trace=True, load_results=False, assert_resu
         popu_stu_pred_ap = np.loadtxt(pref_out + '_pred_ap.popu', dtype=np.object)[:,2]
         popu_stu_pred_oadp = np.loadtxt(pref_out + '_pred_ap.popu', dtype=np.object)[:,2]
     else:
-        pcs_ref, pcs_stu_sp, popu_ref, popu_stu_pred_sp =  fp.run_pca(pref_ref, pref_stu, popu_filename_ref = popu_filename_ref, dim_ref=dim_ref, method='sp', use_memmap=use_memmap, load_saved_ref_decomp=load_saved_ref_decomp, log_level=log_level)
-        pcs_ref, pcs_stu_ap, popu_ref, popu_stu_pred_ap =  fp.run_pca(pref_ref, pref_stu, popu_filename_ref = popu_filename_ref, dim_ref=dim_ref, method='ap', use_memmap=use_memmap, load_saved_ref_decomp=load_saved_ref_decomp, log_level=log_level)
-        pcs_ref, pcs_stu_oadp, popu_ref, popu_stu_pred_oadp =  fp.run_pca(pref_ref, pref_stu, popu_filename_ref = popu_filename_ref, dim_ref=dim_ref, method='oadp', use_memmap=use_memmap, load_saved_ref_decomp=load_saved_ref_decomp, log_level=log_level)
+        pcs_ref, pcs_stu_sp, popu_ref, popu_stu_pred_sp =  fp.run_pca(
+            pref_ref, pref_stu, popu_filename_ref = popu_filename_ref,
+            method='sp', use_memmap=use_memmap, load_saved_ref_decomp=load_saved_ref_decomp, log_level=log_level)[:4]
+        pcs_ref, pcs_stu_ap, popu_ref, popu_stu_pred_ap =  fp.run_pca(
+            pref_ref, pref_stu, popu_filename_ref = popu_filename_ref,
+            method='ap', use_memmap=use_memmap, load_saved_ref_decomp=load_saved_ref_decomp, log_level=log_level)[:4]
+        pcs_ref, pcs_stu_oadp, popu_ref, popu_stu_pred_oadp =  fp.run_pca(
[…]

### 2018-06-17 split_fam.sh

@@ -8,7 +8,8 @@ chunk_n_lines=$2
 
 chunk_len_suff=4
 chunk_midf=_chunksize${chunk_n_lines}_
-chunk_filepref=${bed_filepref}${chunk_midf}
+dir=${bed_filepref}_chunksize${chunk_n_lines}
+chunk_filepref=${dir}/${bed_filepref}${chunk_midf}
 
 # Split .fam file
 split -l ${chunk_n_lines} -a ${chunk_len_suff} -d ${bed_filepref}.fam ${chunk_filepref}

### 2018-06-17 doc/paper.tex

@@
 \usepackage{amsfonts}
-\usepackage{cleveref}
+\usepackage[capitalise]{cleveref}
 \usepackage{verbatim}
@@
 \newcommand{\bO}{\mathcal{O}}
+\newcommand{\bR}{\mathbb{R}}
+\newcommand{\ostu}{\operatorname{stu}}
+\newcommand{\oref}{\operatorname{ref}}
 \newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg\;min}}\;}
@@
 
+\begin{abstract}
+    Population stratification (PS) is an important confounder
+    for the association between genotypes and phenotypes.
+    To adjust for PS,
+    we need to accurately estimate the ancestral structure
+    in the study samples.
+    This can be done effectively
+    by using principle component (PC) analysis
+    to predict the PC scores of the study genotypes
+    from a set of reference genotypes.
+    There are two major existing methods for predicting PC scores:
+    simple projection (SP) and augmentation-decomposition-transformation (ADP).
+    However, when the reference size is large,
+    the former is computationally efficient but inaccurate
+    (PC scores shrinking toward zero),
+    while the latter is accurate but computationally costly ($\bO[n^3]$).
+    To solve these problems,
+    we created the genoPCA software
+    that implemented 
+    the adjusted projection (AP) method to remove the bias in SP, and
+    the online ADP (OADP) method to speed up ADP.
+    We compared the accuracy and computational efficiency of these four methods
+    by conducting a simulation study
+    with the GGS software.
+    We also used them to study the UKbiobank data by using the 1000Genomes data as the reference.
+    We found that AP achieved the accuracy of ADP without adding much runtime,
+    while OADP reduced the runtime of ADP by $x$-fold without losing accuracy.
+    Furturmore, when studying a large number of samples
+    from a narrower geographic area
+    where the signal-to-noise ratio is low,
+    AP and OADP performed reasonably well
+    while SP frequently misclassified the populations
+    and the runtime of ADP became impractical.
+
+\end{abstract}
+
 \section{Introduction}
@@
 To adjust for PS, we need to accurately estimate the ancestry structure in the study samples.
-Principle component analysis (PCA) is a multivariate statistical method which finds the direction of the maximum variability.
-By aggregating information across all genetic markers, PCA can accurate infer the ancestry structure \citep{reich2008principal}.
-To adjust for PS, PCA can be applied to study data to calculate PC scores, which are regarded as variables of ancestry direction and can be used as covariates to adjust for.
-An alternative approach is predicting the PC scores of the study samples using reference genotyped samples with detailed ancestry information.
-This prediction-based approach not only allows to adjust for PS but also to infer ancestry memberships of study samples.
+Principle component analysis (PCA) is a multivariate statistical method
+which finds the direction of the maximum variability among the samples.
+By aggregating information across all genetic markers,
+PCA can accurately infer the ancestry structure \citep{reich2008principal}.
+To adjust for PS,
+PCA can be applied to study data to calculate the principle component (PC) scores,
+which are regarded as variables of ancestry direction
+and can be used as covariates to adjust for.
+An alternative approach for doing PCA is
+predicting the PC scores of the study samples
+by using some reference genotyped samples with detailed ancestry information.
+This prediction-based approach allows not only adjustment for PS
+but also infererence of the ancestry memberships of the study samples.
 
-The standard way of predicting PC scores is to find the PC loadings of the reference data and project the data points of the study samples onto them.
+The standard way of predicting PC scores is to find the PC loadings of the reference samples and project the study samples onto them.
 We call this method simple projection (SP).
-However, when the number of features are greatly larger than the size of the reference sample,
-the predicted PC scores by this method are known to be biased toward null \citep{dey2016asymptotic}.
+However, when the number of features greatly exceeds the size of the reference samples,
+which is common for genotype data,
+the PC scores predicted by SP are known to be biased toward null \citep{dey2016asymptotic}.
 One way of addressing this shrinkage bias is presented by \citet{wang2014ancestry, wang2015improved}.
-Their solution is to combine one study individual with the reference set and
-find the PC scores of this augmented data set by singular value decomposition (SVD).
+Their solution is to combine one study sample with all the reference samples and
+find the PC scores of this augmented data set by using singular value decomposition (SVD).
 The PC scores of the study individuals are then retrieved from the singular vector matrices and mapped to the reference PC space by a Procrustes transformation.
-We call this method ``augment, decompose, and Procrustes transform'' (ADP).
-This method has been shown to be effective in eliminating the shrinking bias of
-study PC scores.
-However since ADP needs to run SVD for each study samples, ADP is computationally expensive.
+We call this method ``augmentation, decomposition, and Procrustes transformation'' (ADP).
+This method has been shown to be effective in eliminating the shrinkage bias of study PC scores.
+However since ADP needs to run SVD for each study sample,
+ADP is computationally expensive.
 
@@
 The first approach removes the bias in SP by estimating the asymptotic bias factor,
-which is calculated based on Random Matrix Theory \citep{dey2016asymptotic}.
-The second approach improves the computation efficiency of ADP by using online SVD algorithm \citep{halko2011finding}. 
-We call the first approach ``adjusted projection'' (AP), and the second approach ``online augment, decompose, and Procrustes transform'' (OADP).
-
-In addition, we also address a computation problem with finding the PCs of the reference set when the reference sample sizes are large. 
-The previously mentioned four methods all need to find the SVD of the reference data first.
-This requires loading the covariance matrix of the reference data into the
-computer's memory. 
-However, when the size of the reference set is large,
-the resulted covariance matrix will far exceed the size of a normal personal
-computer's memory.
-To solve this issue, a randomized SVD method has been developed that does not
-require finding the covariance matrix.
-
-In this paper, we compare the properties of
-SP, ADP, AP, and OADP in terms of computation time and accuracy.
-We also compare using standard SVD and randomized SVD for the reference set
-and examine the difference in accuracy and computation time.
-We found that AP and OADP have both achieved the accuracy of ADP
-and the computational efficiency of SP.
-In addition, randomized SVD gives results similar to standard SVD and only increase the computational time for the reference data.
+which is calculated based on random matrix theory \citep{dey2016asymptotic}.
+The second approach improves the computational efficiency of ADP
+by using an online SVD algorithm \citep{halko2011finding}, 
+which obtains the SVD results of the augmented matrix
+by updating the SVD results of the reference matrix,
+since the latter only differs slightly from the former
+and the overlapping calculations can be avoided
+(at least approximately).
+We call the first approach ``adjusted projection'' (AP)
+and the second approach ``online augmentation, decomposition, and Procrustes transformation'' (OADP). 
+We found that AP and OADP have both achieved accuracy close to that of ADP
+and computational efficiency close to that of SP.
 
@@
 \subsubsection{Data Format}
-There are two genotype datasets, reference and study datasets, and both can be represented by matrices.
-Suppose $X$ is a $p \times n$ matrix of reference genotypes and 
-$Y$ is a $p \times m$ matrix of study sample genotypes, where 
-$p$ is the number of genetic markers, $n$ is the number of reference samples and $m$ is the number of study samples.
-Each entry of $X$ and $Y$ is the number of minor alleles (0,1,2) of single nucleotide polymorphism (SNP).
-For genotype data, it is common for $p$ to be significantly larger than either $n$ or $m$.
-Thus we assume this to be true except for the part where we discuss the randomized SVD algorithm,
-which is developed to handle datasets with exceedingly large $n$.
+For PC score prediction,
+we have a reference group and a study group.
+The reference and study data sets are represented by matrices.
+Let $X$ be a $p \times n$ matrix of reference genotypes
+and  let $Y$ be a $p \times m$ matrix of study sample genotypes,
+where $p$ is the number of genetic markers,
+$n$ is the number of reference samples
+and $m$ is the number of study samples.
+In our study we only consider genotypes composed of biallelic single nucleotide polymorphisms (SNP),
+so each entry of $X$ and $Y$ is a minor allele count of 0, 1, or 2.
 
-\subsubsection{Principle Component Analysis}\label{pca-intro}
-Given a multi-dimensional data, PCA searches for the directions in which the data has the greatest variance.
-These unit vectors are called the principle components (PCs).
-The PC's form an orthonormal basis for the sample space,
-and we can ignore all but the first few PC's to reduce the dimension of the data.
-In particular, let $X$ be standardized so that each row (feature) has mean equal to zero and standard deviation equal to one.
-Moreover, Let $X=UDV^T$ be the SVD of $X$,
-where $U$ is a $p \times n$ matrix whose columns are the PC's,
-$D$ is an $n \times n$ diagonal matrix storing the standard deviations of the PC's,
-and $V$ an $n \times n$ matrix such that each row contains the standardized PC scores of an individual.
-(Alternatively, we can use eigendecomposition to find the matrices since $X^T X = VD^2V^T$ and $X X^T = UD^2U^T$.)
-We sort the diagonal elements of $D$ in descending order and rearrange the
-columns of $U$ and $V$ accordingly.
-We are interested in the top $k$ PC's (where $k \ll n$) and look at the top $k$
-columns of $V$ for their PC scores.
+\subsubsection{Principle Component Analysis on the Reference Data}
+First, we conduct a PCA on the reference data $X$.
+The reference data matrix is standardized
+so that the mean and standard deviation of each feature are $0$ and $1$, respectively.
+With $n$ data points residing in a $p$-dimensional vector space and $n < p$,
+there is a unique $n$-dimensional hyperplane that covers all the data points.
+This hyperplane can be represented as the span of an orthonormal basis composed of $n$ unit vectors.
+Of course, the choice for such orthonormal basis is not unique,
+but what what we are interested in PCA is 
+to maximize the variance of the samples on the direction of the first basis element,
+maximize the variance on the samples on the second basis element after they are collapsed in the direction of the first basis element,
+and so forth,
+until we have $n$ basis elements.
+At the end, PCA gives us three variables:
+\begin{enumerate}
+    \item $n$ pairwise-orthogonal unit vectors of $p$ dimensions
+    representing the orthonormal basis.
+    These are called principle components (PC) or PC loadings,
+    and we organized them into a $p \times n$ matrix $U$.
+    \item $n$ non-negative real numbers
+    each of which represents the standard deviations of the samples
+    on the direction of the corresponding PC loading.
+    They are organized into an $n \times n$ diagonal matrix $D$.
+    \item $n$ tuples of $n$ dimensions
+    representing the normalized coordinates of the data points
+    in the $n$-dimensional space spanned by the PC loadings.
+    These are called the normalized PC scores,
+    and they are organized into an $n \times n$ matrix $V$,
+    where each column corresponds to the coordinate of a PC loading
+    and each row corresponds to a sample.
+\end{enumerate}
+If we multiply each normalized PC score by the corresponding standard deviation
+(i.e. calculate $VD$),
+we get the unnormalized PC scores.
+In our study, we use unnormalized PC scores and simply call them PC scores.
+The study can be carried out by using normalized PC scores, too,
+and the result would be equivalent.
+Moreover, the standard deviation matrix $D$ is in descending order,
+so that the first column of $U$ is the direction of the greatest sample variance,
+the first column of $V$ represents the coordinates of the $n$ samples in this direction.
+The same correspondence applies to $U$'s and $V$'s second column, third column, etc.
+Thus PCA reduces the dimension of the sample space from $p$ to $n$.
+We can further reduces the sample space dimension
+by using only the first $k$ PCs ($k \ll n$) and ignore the rest $n-k$ PCs.
+In genetic studies, choosing $k \leq 10$ is often sufficient.
 
-\subsubsection{Estimating PC scores}
+The (standardized) reference data matrix can be restored by $X = U D V^T$.
+In fact, this is exactly the SVD of $X$ \citep{jolliffe2002principal}.
+Thus PCA on the reference data can be done easily on any language
+in which a SVD algorithm has been implemented.
+If all that we are interested in is the PC scores,
+we can reduce the computational complexity by applying eigen decomposition (EGD) on $X^TX$
+to obtain $X^T X = V D^2 V^T$.
+Afterwards, $U$ can be retrieved by $U = X V D^{-1}$ if needed.
 
-population.
-In the ideal case in which the genotypes of all individuals of the population are given
-we use the method in \Cref{pca-intro} to find the PC scores of the individuals
-we are interested in.
-In practice, however, we have a reference dataset and a study dataset assumed to be generated from the
-same population and use the former to estimate the populational PC scores of the latter.
-There are several computational methods one can use to do this estimation.
-In \Cref{sec:compu}, we compare the performance of four of them.
+\subsubsection{Predicting the PC Scores of the Study Samples}
 
-\subsection{Computational Methods}\label{sec:compu}
+Once PCA is finished on the reference data,
+we predict the PC scores of the study data.
+There are more than one ways to accomplish this,
+and two common methods are SP and ADP.
+The performance of these two methods can be improved by using AP and OADP.
+We now discuss the algorithm, computational efficiency, and accuracy of these four methods.
 
-\subsubsection{Simple Projection (SP)}
-The standard way for estimating PC scores is by projecting the study
-individuals' data points to the PC space of the reference individuals by using reference sample PC loadings,
-which is the matrix $U$ in the eigendecomposition of the (standardized) reference sample matrix $X$.
-The PC loadings are unit vectors pointing toward the directions of the greatest variances in the $p$-dimension sample space.
-We call this method ``simple projection'' (SP).
-The algorithm goes as follows.
+Before we using any of these methods,
+we need to standardize the study samples
+by subtracting the reference means from them
+and then dividing them by the reference standard deviations.
+For simplicity,
+we overload the symbol $Y$
+and let it represent the standardized study data for the rest of the paper.
+
+\paragraph{Simple Projection (SP)}
+This method is simple and intuitive.
+Since the PCA on the reference data has already given us the PC loadings,
+we can simply project a (standardized) study sample onto them by calculating their dot products.
+The resulted projection lengths are the predicted PC scores for this study sample.
+The algorithm is
 
 \begin{enumerate}
-\item Compute $X^T X$.
-  (Computational complexity: $\bO[n^2p]$.)  
-\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
-  ($\bO[n^3]$.)
-\item Calculate PC loading matrix $L = X V_1 D_1^{-1}$,
-where $V_1$ and $D_1$ are the the first $k_1$ columns of $V$ and $D$, respectively ($\bO[k_1^2 n + knp]$)
-\item Calculate predicted PC scores $W = Y^T L$, where $Y$ is the data matrix for all the study individuals ($\bO[mpk_1]$.)
+\item Compute the reference covariance matrix $X^T X$.
+  (Computational complexity (CC): $\bO[pn^2]$.)  
+\item EGD on the reference covariance matrix: $X^T X = V D^2 V^T$.
+  (CC: $\bO[n^3]$.)
+\item Compute the PC loading matrix for the top $k$ PCs only: $U_k= X V_k D_k^{-1}$.
+Here $V_k$ and $D_k$
+are the the first $k$ columns of $V$
+and the upper-left $k \times k$ submatrix of $D$, respectively.
+(CC: $\bO[pnk]$.)
+\item Compute the predicted study PC scores for the top $k$ PCs: $W_k = Y^T U_k$.
+(CC: $\bO[mpk]$.)
 \end{enumerate}
+The total computational complexity is  $\bO[pn^2 + mpk]$
+given that $k \ll n \ll p$.
+The computational complexity for each study sample is $\bO[pk]$,
+which is the lowest among all the methods discussed in this paper.
+However, though computationally efficient,
+a major disadvantae of SP is the loss of accuracy when the number of features $p$ greatly exceeds the reference sample size $n$,
+a situation that is not uncommon for genotype data.
+\citet{dey2016asymptotic} have shown that when $n \ll p$,
+especially when the signals in the reference samples are not strong,
+the predicted PC scores will shrink toward the origin.
+This shrinkage bias limits the accuracy of SP for high dimensional data.
+Fortunately, \citet{dey2016asymptotic} have also provided a way
+to estimate the asymptotic shrinkage magnitude and angle
+and implemented it in the \verb|hdpca| package in the R language \citep{hdpca}.
+We use this estimation to eliminate the shrinkage bias,
+as in the AP method.
 
-Total computational complexity: 
+\paragraph{Adjusted Projection (AP)}
+
+This method is almost identical to SP,
+with the only difference being the addition of the adjustment for the shrinkage bias
+after the projections are finished to restore the unshrinked PC scores.
+For estimating the asymptotic shrinkage factor,
+we need all the eigenvalues
+of the the reference data matrix,
+as well as the number of features
+and the reference sample size.
+The details for estimating the shrinkage factor can be found in \citet{dey2016asymptotic}.
+The algorithm is summarized below.
+
+\begin{enumerate}
+\item Compute the reference covariance matrix $X^T X$.
+  (CC: $\bO[pn^2]$.)  
+\item EGD on the reference covariance matrix: $X^T X = V D^2 V^T$.
+  (CC: $\bO[n^3]$.)
+\item Estimate the shrinkage magnitudes and organize them in a $n \times n$ diagonal matrix $S$.
+  (CC: $\bO[p]$.)
+\item Compute the PC loading matrix for the top $k$ PCs with the adjustment for the shrinkage bias: $U_k= X V_k D_k^{-1} S_k^{-1}$.
+Here $V_k$, $D_k$, and $S_k$
+are the the first $k$ columns of $V$
+and the upper-left $k \times k$ submatrices $D$ and $S$, respectively.
+(CC: $\bO[pnk]$.)
+\item Compute the predicted study PC scores for the top $k$ PCs: $W_k = Y^T U_k$.
+(CC: $\bO[mpk]$.)
+\end{enumerate}
+Given that $k \ll n \ll p$,
+the total computational complexity is $\bO[pn^2 + mpk]$,
+which is the same as that of SP.
+Thus the adjustment for the shrinkage bias in AP
+does not increase the computational complexity of SP
+(though it does slightly increase the empirical runtimes).
+A minor disadvantage of AP is that
+while the computatinal complexity and memory usage of SP
+can be further reduced
+by using some truncated SVD algorithm
+(such as the randomized SVD algorithm
+by \citet{halko2011finding}
+as discussed in \cref{sec:randsvd})
+to compute the SVD for only the top $k$ PCs of the reference matrix,
+AP requires all the eigenvalues
+and thus a full SVD or EGD of the reference matrix.
+
+\paragraph{Augmentation, Decomposition, and Procrustes Transformation (ADP)}
+
+ADP predicts the study PC scores by using a completely different approach
+compared to SP and AP.
+The details of this method
+and an implementation in C++ called TRACE
+are described in \cite{wang2014ancestry,wang2015improved}.
+In ADP,
+we augment the (standardized) reference matrix by one column
+by appending a (standardized) study sample to it.
+Then we apply SVD or EGD to this $p \times (n+1)$ matrix $\tilde{X}$.
+The resulted $(n+1) \times (n+1)$ normalized PC score matrix $\tilde{V}$
+can be divided into two submatrices:
+the first $n$ rows $\tilde{V}_{\oref}$,
+which correspond to the reference samples,
+and the last row $\tilde{v}_{\ostu}$,
+which corresponds to the study sample.
+Notice that $\tilde{V}_{\oref}$ is now different from $V$,
+the $n \times n$ normalized PC score matrix of the reference data,
+so we look for a linear transformation
 \[
-    \bO[n^2p + n^3 + k_1^2 n + pnk_1 + pmk_1)] \approx \bO[n^2p + mk_1p]
+    f(\tilde{V}_{\oref}) = \rho A \tilde{V}_{\oref} + c
 \]
-given that $k_1 \ll n \ll p$.
+where $\rho$ is a non-negative real number
+(for uniform scaling),
+$A$ is a $n \times (n+1)$ orthonormal matrix
+(for projection, rotation and reflection),
+and $c$ is a column vector
+(for translation),
+so that the distance
+between $V$ and the transformed $f(\tilde{V}_{\oref})$
+(i.e. the sum of squares of the error matrix $V - f(\tilde{V}_{\oref})$)
+is minimized.
+This is called the Procrustes transformation
+from $f(\tilde{V}_{\oref})$ to $V$.
+We then apply this transformation to $\tilde{v}_{\ostu}$
+to obtain $f(\tilde{v}_{\ostu})$.
+Finally,
+by multiplying these Procrustes-transformed normalized PC scores
+by the reference PC standard deviations,
+we obtain the predicted PC score for the study sample $f(\tilde{v}_{\ostu})D$.
 
-As the reference sample size increases,
-the computational complexity of SP stays as a constant,
-except for decomposing the reference matrix, which is done once for all study individuals.
-However, it is known that the projection method tends to underestimates the magnitude of the PC scores when $n \ll p$ \citep{dey2016asymptotic},
-which we call the ``shrinking bias''.
-This limits the accuracy of SP for high dimensional data.
+Notice that so far we have been using all the PCs for ADP.
+However, this is not necessary
+if all that we are interested in
+is the top few PCs.
+Thus we can use only
+the top $k$ PC scores of the reference matrix
+and the top $lk$ PC scores of the augmented matrix
+(where $l$ is a constant greater than or equal to $1$ and $lk \ll n$)
+and find the Procrustes transformation from $\bR^{lk}$ to $\bR^k$.
+The algorithm is summarized as follows.
 
-\subsubsection{Augment, Decompose, and Transform (ADP)}
+\begin{enumerate}
+\item Compute the reference covariance matrix $X^T X$.
+  (CC: $\bO[pn^2]$.)  
+\item Apply EGD to the reference corariance matrix
+  to obtain $X^T X = V D^2 V^T$.
+  (CC: $\bO[n^3]$.)
+\item For a study sample $y$,
+  find the augmented covariance matrix
+  $\tilde{X}^T \tilde{X}$
+  by computing $X^T y$, $(X^T y)^T$, and $y^T y$
+  and appending them to the
+  bottom edge,
+  right edge,
+  and bottom-right corner of $X^TX$.
+  (CC: $\bO[pn]$.)
+\item Apply EGD to $\tilde{X}^T \tilde{X}$ to get $\tilde{X}^T \tilde{X} = \tilde{V} \tilde{D}^2 \tilde{V}^T$.
+  (CC: $\bO[n^3]$.)
+\item Find the Procrutes transformation
+from the first $lk$ columns of $\tilde{V}$
+to the first $k$ columns of $V$.
+  (CC: $\bO[nk^2]$)
+  \item Go to step 3 for the next study sample
+  until all the study samples are analyzed.
+\end{enumerate}
+The total computational complexity is
+$\bO[pn^2 + m(np + n^3)]$
+given that $lk \ll n \ll p$.
 
-One way to solve the shrinking bias of SP is the ADP approach developed by \citet{wang2014ancestry, wang2015improved}.
-In this method, each study individual's data point is appended to the reference
-data matrix.
-Then SVD is applied to the augmented matrix,
-and PC scores in this augmented sample are found for both the reference
-individuals and the study individual.
-The the PC scores of the reference individuals in this augmented sample is
-compared with the PC scores of the reference individuals in the reference
-sample,
-which is calculated in the beginning,
-and a linear (Procrustes) map is constructed to transform the latter into the former
-with the smallest error.
-This transformation is then applied to the PC scores of the study individual in
-the augmented sample.
-The resulted PC scores are used as an estimate for the study individual's
-population PC scores.
-Specifically, let
-\[
-  V =
-  \begin{bmatrix}
-    V_{\text{ref}} & W
-  \end{bmatrix},
-  \quad
-  \tilde{V} =
-  \begin{bmatrix}
-    \tilde{V}_{\text{ref}} & \tilde{W}_\text{ref} \\
-    \tilde{V}_\text{stu} & \tilde{W}_\text{stu}
-  \end{bmatrix},
-\]
-where the dimensions of $V_\text{ref}$, $\tilde{V}_\text{ref}$, and $\tilde{V}_\text{stu}$ are $n \times k_1$, $n \times k_2$, and $1 \times k_2$, respectively, with $ 1 \leq k_1 \leq k_2 \leq n+1$. 
-Next, we use Procrustes analysis to find
-\[
-  A, b = \argmin{A, b} d(V_\text{ref}, \tilde{V}_{\text{ref}}A+b)
-\]
-for some metric $d$.
-Finally, we get the Procrustes-adjusted top $k_1$ PC scores for the study individual $V_\text{pro} = \tilde{V}_\text{stu} A + b$.
-The algorithm is summarized as follows:
-\begin{enumerate}
-\item Compute $X^T X$.
-  (Computational complexity: $\bO[n^2p]$.)  
-\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$ (columns of $D$ and $V$ are sorted by PC variance with the first column corresponding to the greatest PC variance).
-  ($\bO[n^3]$.)
-\item Append $X^T y$ and its transpose to the last column and row of $X^T X$, respectively.
-  The entry of the bottom-right corner is $y^T y$.
-  This is exactly the same as computing $\tilde{X}^T \tilde{X}$,
-  where $\tilde{X} = (X, y)$ is the reference group with the study individual added.
-  ($\bO[np]$.)
-\item Apply eigendecomposition to $\tilde{X}^T \tilde{X}$ to get $\tilde{X}^T \tilde{X} = \tilde{V} \tilde{D}^2 \tilde{V}^T$.
-  ($\bO[n^3]$.)
-\item Project the first $k_2$ columns of $\tilde{V}$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual (the last row of $V$),
-  where $k_1 < k_2 \ll n$.
-  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
-  \item Go to step 3 for the next study individual.
-\end{enumerate}
+ADP is regarded as the most accurate method for predicting study PC scores.
+It is does not have SP's shrinkage bias.
+However, the major disadvantage of ADP
+is its high computational complexity.
+In particular,
+as the reference size increases,
+the computational cost for a study sample
+increases cubically.
+Fortunately, this disadvantage can be mitigated by taking some computational shortcuts,
+as in the OADP method.
 
-Total Computational complexity: $\bO[n^2p + m(np + n^3)]$.
+\paragraph{Online Augmentation, Decomposition, and Procrustes Transformation (OADP)}
 
-Although accurate, ADP is the slowest among the other three computational methods.
-Its computational complexity increases cubically with respect to the reference sample size and linearly with respect to the study sample size.
-The ADP method is has been implemented in the TRACE software \citep{wang2014ancestry, wang2015improved}.
+Notice that the augmented data matrix $\tilde{X}$
+only differs slightly from the reference matrix $X$
+when the reference sample size is large.
+Thus the computational process for the SVD of $\tilde{X}$
+is numerically close to that for the SVD of $X$.
+If we can avoid the repeated computation in this process
+and obtain the SVD of $\tilde{X}$
+by updating the SVD of $X$
+instead of recalculating everything,
+we could greatly reduce the computational cost.
+One of such ``online'' algorithms for computing the SVD of the augmented matrix is discussed in \citet{brand2002incremental}.
+This algorithm calculates SVD in the incremental manner described above
+and has the ability to find the approximate SVD for the top few singular values only,
+which exactly fits our need for obtaining the top few PCs.
+Thus we use the algorithm by \citet{brand2002incremental}
+to replace the standard algorithm for decomposing the augmented matrices
+and coin the name ``online augmentation, decomposition, and Procrustes transformation'' (OADP).
+The algorithm for this method is as follows:
 
-\subsubsection{Adjusted Projection (AP)}
-
-A more efficient way to solve the shrinking bias of SP is to estimate the
-magnitude and angle of the shrinkage and adjust the study individuals' PC scores accordingly.
-This method has been developed by \citet{dey2016asymptotic} and implemented in the High
-Dimension Principal Component Analysis (HDPCA) in the R language \citep{hdpca}.
-HDPCA reads the PC scores and variances calculated by projection and estimate the shrinkage factor to restore the unshrinked scores. 
-The algorithm is
-\begin{enumerate}
-\item Compute $X^T X$.
-  (Computational complexity: $\bO[n^2p]$.)  
-\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
-  ($\bO[n^3]$.)
-\item Calculate $W = Y^T (X V D_1^{-1})$, where $Y$ is the data matrix for all the study individuals, $D_1$ the first $k_1$ columns of $D$. ($\bO[mpk_1 + pnk_1]$.)
-  \item Input $D$ and $W$ to the pc\_adjust function in the HDPCA function to obtain the PC scores adjusted for the shrinking bias.
-($\bO[p]$ for finding the number of distant spikes (function select.nspike), $\bO[n]$ for adjusting for the shrinking bias (function hdpc\_est))
-\end{enumerate}
-
-Total computational complexity: $\bO[n^2p + mk_1(p + n) + p + n] \approx \bO[n^2p + mk_1p]$,
-given that $m$ or $n$ is large. The computational complexities between simple projection and adjusted projection are the same.
-The accuracy of adjusted projection will be empirically shown later.
-
-\subsubsection{Online ADP (OADP)}
-
-A remedy for ADP is to reduce the computational complexity without much sacrifice of the accuracy.
-This is achieved by the online ADP (OADP) method.
-Notice that in OADP, only the top $k_2$ PC's are used.
-OADP uses the online SVD method, which calculates a close approximation of the first $k_3$ columns of $\tilde{V}$ for some $k_2 \leq k_3 \ll n$ and keeps everything else the same as in ADP, including Procrustes analysis.
-The algorithm is
 \begin{enumerate}
@@
 
-\subsubsection{Randomized SVD}
+\subsection{Simulation Studies}
 
-In addition to estimating the PC scores of the study individuals,
-we have another algorithm that aims at improving the decomposition of the
-training data.
-For SP, AP, ADP, and OADP, we need to find the covariance matrix of the training
-data, which has dimension $n \times n$, and conduct eigendecomosition on it.
-However, when $n$ is exceedingly large,
-computation and memory cost of PCA on the reference samples can be enormous.
-A randomized SVD (RSVD) algorithm has been developed to handle this issue.
-When the reference sample size is small, RSVD can be slower than standard SVD,
-but it is scalable to large sample size.
-For the details of the RSVD algorithm, see \citet{halko2011finding}.
+We used the Grid Genotype Simulator (GGS)
+by \citet{ mathieson2012differential}
+to simulate population migration.
+The grid's dimension is 2 by 2
+with the same number of individuals
+in each of the 4 populations.
+Moreover, we let the number of reference samples
+range from 1000 to 3000
+and fix the number of study samples to 200.
+Each genotype sample has 100 genealogies
+with 1000 loci in each genealogy.
+In addition, the migration rate was set to 100.
+After the individuals were simulated,
+we applied SP, ADP, AP, and OADP to the data
+to predict the PC scores for the study samples.
+Then we used the k-nearest neighbors method
+to predict which ethnic group a study sample comes from.
 
-Notice that the randomized SVD method on the training data
-only calculates the first $k$ eigenvalues.
-However, in order to use adjusted projection,
-all the eigenvalues of the training data are required.
-Thus for the missing eigenvalues,
-we predict them by regressing their log-scale values on the ranks
-and using the first $k$ eigenvalues for training the linear regression.
+\subsection{UK Biobank data analysis}
 
-\section{Data}
-
-In this study, we used both simulated data and real-life data to conduct
-empirical tests of the computational methods for PCA.
-
-\subsubsection{Simulation Studies}
-
-We used the GGS software to simulate individuals migrating on a grid.
-The parameters we chose to use are as follows:
-\begin{enumerate}
-\item L: Number of loci in each genealogy. (Fixed to $1000$.)
-\item G: Number of genealogies. (Equal to $p / L$)
-\item K: Side length of the grid. (Fixed to $2$.)
-\item c: Number of haploids at each cell of the grid. (Equal to $2n / K^2$ and
-$2m / K^2$)
-\item M: Migration rate. (Fixed to $100$)
-\end{enumerate}
-
-For our study, we used different reference and study sample sizes.
-First, we fix
-the study size to be $m=200$ and increase $n$ from $1000$ to $3000$ with the
-increment equal to $500$.
-Next, we fix
-the reference size to be $n=600$ and increase $m$ from $1000$ to $3000$ with the increment equal to $500$.
-
-For comparing the speed of the algorithms,
-we only compare the runtimes for the study individuals.
-The time spent on calculations that are once-for-all (such as finding the SVD of
-the reference matrix) is ignored.
-
-For comparing the accuracy, we use two different measures.
-First, for a given cell of the grid, we compute the mean of the PC scores of the 
-reference individuals located at this cell.  
-This is the estimated coordinate of the cell.
-Next, we find the Euclidean distance between the estimated PC scores of the
-study individuals and the estimated coordinates of the cells they are located
-at.
-These are the geographic estimation errors for the study individuals.
-The average of these errors are used to measure the accuracy of the
-computational methods (that is, the square root of the sum of squares divided by the
-study size). 
-Second, since ADP is a model free approach and believed to be the most reliable, we compare the average
-Euclidean distance between the PC score estimates calculated by another method
-to those calculated by ADP. 
-The comparison of the runtimes and accuracy are also conducted by using both the
-standard SVD and randomized SVD on the training data matrix.
-
-\subsubsection{UK Biobank data analysis with 1000 Genomes as reference}
-
-The real-life datasets used in our study are the UKBioBank and 1000 Genomes datasets.
-They contain the genotype readings of individuals from different ethnic groups over the world.
-For better comparison, we select only the SNPs that are included in both datasets,
-the number of which is approximately $1.5 \times 10^5$.
-We use the 1000 Genomes dataset as the reference and removed individuals who have at least one parent who is also in the dataset.
-The reference samples size is approximately $2300$.
-Moreover, we use the first 1000 individuals in the UKBioBank data set out of a
-total number of $5 \times 10^5$ individuals. 
-The measures we use for comparing speed and accuracy are the same as for
-simulated data,
-except that we do not use the geographic error here.
-In addition, both the standard and randomized SVD on the training data set are tested.
+We also applied the four methods
+to study the UK Biobank data \citep{sudlow2015uk},
+which consists of 490,000 middle- or elderly-aged individuals
+in the United Kingdom.
+Our reference samples
+are from the 1000 Genomes Project \citep{10002015global},
+which collected 2500 samples from Europe, Africa, East Asia, South Asia, and America.
+(Individuals whose children are also in the data were removed.)
+The reference and the study samples
+have 145,000 SNPs in common.
+After we predicted the PC scores and ethnic groups
+of the study samples,
+we selected the study samples predicted to by Europeans
+and applied PCA to them
+by using the European reference samples only.
 
 \section{Results}
+
+Simulation study
+
+Runtimes vs ref size
+Distance to ADP vs ref size
+Distance to geo center vs ref size
+PC1 vs PC2, 5 x 3 grid (uncolored study samples)
+
+UK Biobank
+
+500 samples, 4 methods
+All continents
+European only
+misclassification rate compared to ADP
+distance to ADP
+
+all samples, 3 methods
+same as above
 
@@
 
+\appendix
+
+\section{Randomized SVD}\label{sec:randsvd}
+
+In addition to estimating the PC scores of the study individuals,
+we have another algorithm that aims at improving the decomposition of the
+training data.
+For SP, AP, ADP, and OADP, we need to find the covariance matrix of the training
+data, which has dimension $n \times n$, and conduct eigendecomosition on it.
+However, when $n$ is exceedingly large,
+computation and memory cost of PCA on the reference samples can be enormous.
+A randomized SVD (RSVD) algorithm has been developed to handle this issue.
+When the reference sample size is small, RSVD can be slower than standard SVD,
+but it is scalable to large sample size.
+For the details of the RSVD algorithm, see \citet{halko2011finding}.
+
+Notice that the randomized SVD method on the training data
+only calculates the first $k$ eigenvalues.
+However, in order to use adjusted projection,
+all the eigenvalues of the training data are required.
+Thus for the missing eigenvalues,
+we predict them by regressing their log-scale values on the ranks
+and using the first $k$ eigenvalues for training the linear regression.
+
+\newpage
+
 \bibliographystyle{plainnat}
@@
 \newpage
+
+\section{Tables and Figures}
 

### 2018-06-17 mergePlot.R
Add ukbiobank result in paper

@@ -1,108 +1,0 @@
-library(RColorBrewer)
-library(scales)
-
-
-meth.n <- 7
-chg.var <- "m"
-chg.jobid <- "27483942"
-chg.range <- seq(1000, 3000, 500)
-chg.taskid <- c(0:4)
-chg.length <- length(chg.range)
-
-file.head <- "data/"
-file.neck <- paste0(
-  chg.var,
-  "Chg_rand_",
-  chg.jobid,
-  "_",
-  chg.taskid
-)
-file.upper <- paste0(
-  file.head,
-  file.neck,
-  "_100000_")
-if(chg.var == "n"){
-  file.ab <- paste(chg.range, "200", sep = "_")
-  xlab <- "Reference Size"
-} else if(chg.var == "m"){
-  file.ab <- paste("600", chg.range, sep = "_")
-  xlab <- "Study Size"
-} else{
-  stop("Invalid changing variable. Must be m or n. ")
-}
-file.leg <- "_2_1_100.rand"
-out.pref <- paste0(file.upper,
-                  file.ab,
-                  file.leg)
-
-runtimes.ref <- matrix(0, chg.length, meth.n + 1)
-runtimes.ref[,1] <- chg.range
-runtimes.stu <- matrix(0, chg.length, meth.n + 1)
-runtimes.stu[,1] <- chg.range
-accuracy.ctr <- matrix(0, chg.length, meth.n + 1)
-accuracy.ctr[,1] <- chg.range
-accuracy.gold <- matrix(0, chg.length, meth.n + 1)
-accuracy.gold[,1] <- chg.range
-for(i in 1:chg.length){
-  runtimes.file <- paste(out.pref[i], "runtimes", sep = ".")
-  accuracy.file <- paste(out.pref[i], "accuracy", sep = ".")
-  runtimes.df <- read.table(runtimes.file, header = TRUE)
-  accuracy.df <- read.table(accuracy.file, header = TRUE)
-  meth.name <- runtimes.df$method
-  runtimes.ref[i,(1:meth.n)+1] <- runtimes.df$reference
-  runtimes.stu[i,(1:meth.n)+1] <- runtimes.df$study
-  accuracy.ctr[i,(1:meth.n)+1] <- accuracy.df$err.center
-  accuracy.gold[i,(1:meth.n)+1] <- accuracy.df$err.golden
-}
-if(chg.var == "n"){
-  png("doc/nChg.png", width = 4000, height = 4000, res = 300)
[…]

### 2018-06-17 doc/paper/paper.tex

--- 2018-06-17 doc/paper.tex
@@
 \usepackage{float}
+\usepackage{graphicx}
 
@@
 by using the European reference samples only.
+Since ADP is very slow compared to the other three methods,
+we randomly selected 500 study samples
+and used them to compare the four methods.
+After that,
+we applied SP, AP, and OADP to all the study samples.
 
@@
 
-Simulation study
+\subsection{Simulation study}
 
-Runtimes vs ref size
-Distance to ADP vs ref size
-Distance to geo center vs ref size
-PC1 vs PC2, 5 x 3 grid (uncolored study samples)
+% Runtimes vs ref size
+% Distance to ADP vs ref size
+% Distance to geo center vs ref size
+% PC1 vs PC2, 5 x 3 grid (uncolored study samples)
+% The predicted PC scores are shown in \Cref{fig:n1000} and \Cref{fig:ukb}.
 
-UK Biobank
-
-500 samples, 4 methods
-All continents
-European only
-misclassification rate compared to ADP
-distance to ADP
-
-all samples, 3 methods
-same as above
-
-The predicted PC scores are shown in \Cref{fig:n1000} and \Cref{fig:ukb}.
-For the comparison of the runtimes and accuracy for different reference and sample sizes for the simulated data,
-see \Cref{fig:nChg} and \Cref{fig:mChg}.
-For comparing the methods for the real-life data, see \Cref{tbl:ukb}.
-
-\section{Discussion}
-
-For the real-life datasets (\Cref{tbl:ukb} and \Cref{fig:ukb}),
+For the simulation study
+(\Cref{tbl:nChg-runtimes-ref}, \Cref{tbl:nChg-runtimes-study},
+\Cref{tbl:nChg-accuracy-gold}, \Cref{tbl:nChg-accuracy-ctr},
+\Cref{tbl:mChg-runtimes-study}),
 we see that ADP, which is the most reliable method,
@@
 By using the HDPCA package,
-AP reduces the error of AP
+AP reduces the error of SP
 and achieved the second highest in accuracy.
-The randomized version of SP is as accurate as the standard SP,
-which shows that the randomized SVD algorithm gives reliable results,
-and that using linear regression to predict the uncalculated reference eigenvalues is satisfactory.
-It also reduces the runtime almost by half,
-due to not calculating all the eigenvalues.
-Similar reduction in study runtime is observed for the randomized version of AP,
-but in this case the accuracy is also decreased,
-thought it is still higher than that of SP and RSP.
-Finally, the randomized OADP is disappointing,
-as its error is almost as high as SP and RSP.
-
-The pattern for accuracy is very similar when we look at the simulated data.
-(\Cref{tbl:nChg-runtimes-ref}, \Cref{tbl:nChg-runtimes-study},
-\Cref{tbl:nChg-accuracy-gold}, \Cref{tbl:nChg-accuracy-ctr},
-\Cref{tbl:mChg-runtimes-study}).
-A noticeable difference in this case is the accuracy of ROADP.
-Here ROADP's accuracy is almost identical to OADP's,
-as measured by both the golden standard error and the geographic center err.
 However, as the reference size increases,
@@
 As for the speed,
-AWD's study runtime increases so fast with the reference size
+ADP's study runtime increases so fast with the reference size
 that the study runtimes for the other methods seem to be constant in comparison (\Cref{fig:nChg}).
@@
 This information can be used to predict the runtime for larger study samples.
+
+\subsection{UK Biobank data}
+
+When we used the global samples,
+we found that the PC scores predicted by the four methods' are almost identical.
+The separation of the ethnic groups was clear,
+and the shrinkage bias of SP was not obvious,
+which is consistent with the theoretical understanding
+that the shrinkage bias decreases
+as the signal-to-noise ratio increases.
+
+However, when we focus on the European (EUR) samples only,
+the results given by SP were clearly unreliable.
+The samples from British in England and Scotland (GBR)
+and those from Utah residents with northern and western European ancestry	(CEU)
+drifted very close to NULL.
+This causes SP to incorrectly classified almost all of them as CEU,
+since most GBR reference samples are far away from NULL.
+A few of the study samples predicted by ADP to be GBR or CEU
+were even classified as the Iberian population in Spain (IBS),
+since the IBS reference samples are close to NULL.
+Moreover, the shrinkage for the IBS
+and for the Toscani in Italia (TSI) populations
+caused significant errors within these two groups.
+The IBS reference samples
+are approximately in the middle
+between the TSI reference samples and NULL.
+Thus most of the the study samples predicted by ADP to be TSI
+were classified by SP as IBS.
+
+For OADP, its predicted PC1 and PC2 scores
+for the European study samples
+are almost identical
+to those done by ADP.
+OADP's results for PC3 and PC4
+are also very close to ADP's results,
+though we observed a very slight decrease in the magnitude of the PC scores.
+This minor shrinkage did not affect
+OADP's classification for the IBS and TSI samples,
+since their PC1 and PC2 scores were accurately predicted by OADP.
+Neither was any CEU/GBR sample classified as IBS/TSI. 
+For the CEU and GBR samples,
+their predicted PC scores are tightly mixed together
+for all the 4 PCs,
+which were even difficult for ADP to separate them.
+Thus the slight difference in PC3 and PC4 scores
+between OADP and ADP
+caused the classification of many CEU and GBR samples
+to be different on the two methods.
+
+Furthermore, AP successfully adjusted for the shrinkage
+in PC1 and PC2 for the European samples,
+so its predicted PC scores are almost identical to ADP's.
+However, it did not detect the shrinkage in PC 3 and PC4,
+so its predicted PC scores are the same as SP's for these two PCs.
+This still allowed AP to correctly separate
+the CEU/GBR samples from the IBS/TSI samples.
+However, its separation of the IBS and TSI samples
+did not agree completely with that done by ADP,
+most likely caused by the shrinkage on PC3.
+For the classification between CEU and GBR,
+the AP's results differ far from ADP's,
+as AP classified many more CEU/GBR samples as CEU.
+
+The patterns for the accuracy of SP, AP, and OADP
+discussed above
+were also observed
+when we used all the 490,000 study samples in the UK Biobank dataset
+and those predicted to be Europeans among them.
+See \Cref{tbl:ukb} for details.
+
+\section{Discussion}
+
+For the future, we would like
+to improve the parallelization of our software.
+Currently it uses the poor-man’s parallelism,
+that is,
+dividing the study dataset into multiple smaller files on the hard drive,
+applying the analysis to each of them,
+and merge the results.
+This parallelism saves both the memory usage and the runtime.
+However, we would like to utilize
+more sophisticated parallel computation techniques
+to further improve the performance of our software.
 
@@
 
-In addition to estimating the PC scores of the study individuals,
-we have another algorithm that aims at improving the decomposition of the
-training data.
-For SP, AP, ADP, and OADP, we need to find the covariance matrix of the training
-data, which has dimension $n \times n$, and conduct eigendecomosition on it.
-However, when $n$ is exceedingly large,
-computation and memory cost of PCA on the reference samples can be enormous.
-A randomized SVD (RSVD) algorithm has been developed to handle this issue.
-When the reference sample size is small, RSVD can be slower than standard SVD,
-but it is scalable to large sample size.
-For the details of the RSVD algorithm, see \citet{halko2011finding}.
+% In addition to estimating the PC scores of the study individuals,
+% we have another algorithm that aims at improving the decomposition of the
+% training data.
+% For SP, AP, ADP, and OADP, we need to find the covariance matrix of the training
+% data, which has dimension $n \times n$, and conduct eigendecomosition on it.
+% However, when $n$ is exceedingly large,
+% computation and memory cost of PCA on the reference samples can be enormous.
+% A randomized SVD (RSVD) algorithm has been developed to handle this issue.
+% When the reference sample size is small, RSVD can be slower than standard SVD,
+% but it is scalable to large sample size.
+% For the details of the RSVD algorithm, see \citet{halko2011finding}.
 
-Notice that the randomized SVD method on the training data
-only calculates the first $k$ eigenvalues.
-However, in order to use adjusted projection,
-all the eigenvalues of the training data are required.
-Thus for the missing eigenvalues,
-we predict them by regressing their log-scale values on the ranks
-and using the first $k$ eigenvalues for training the linear regression.
+% Notice that the randomized SVD method on the training data
+% only calculates the first $k$ eigenvalues.
+% However, in order to use adjusted projection,
+% all the eigenvalues of the training data are required.
+% Thus for the missing eigenvalues,
+% we predict them by regressing their log-scale values on the ranks
+% and using the first $k$ eigenvalues for training the linear regression.
 
@@
 \bibliographystyle{plainnat}
-\bibliography{frugalpca}
+\bibliography{paper}
 
@@
     \hline
-    ROADP & $\bO[k^2 p]$ & $\bO[mk(p + k n)]$ \\
-    \hline
     SP & $\bO[n^2p]$ & $\bO[mkp]$ \\
     \hline
-    RSP & $\bO[k^2p]$ & $\bO[mkp]$ \\
-    \hline
     AP & $\bO[n^2p]$ &  $\bO[mkp]$ \\
-    \hline
-    RAP & $\bO[k^2p]$ &  $\bO[mkp]$ \\
     \hline
@@
     \hline
-    Method & Reference Runtimes (n=?) & Study Runtimes (m=?)& Golden Standard Error \\ 
+    Method & Reference Runtimes (n=2492) & Study Runtimes (m=1000)& Golden Standard Error \\ 
     \hline
@@
     \hline
-    ROADP & 2518 & 103 & 1.93 \\
-    \hline
     SP & 411 & 425 & 1.95 \\
     \hline
-    RSP & 2504 & 241 & 1.95 \\
-    \hline
     AP & 411 & 482 & 1.12 \\
-    \hline
-    RAP & 2505 & 250 & 1.83 \\ 
     \hline
@@
     OADP       &  72  & 18   & 19   & 22   & 24   \\
-    ROADP      & 809  & 18   & 17   & 20   & 21   \\
     SP         &  68  & 26   & 42   & 67   & 96   \\
-    RSP        & 805  & 23   & 31   & 39   & 46   \\
     AP         &  68  & 69   & 107  & 91   & 119  \\
-    RAP        & 805  & 11   & 24   & 29   & 9    \\
     \hline
@@
     OADP       & 15   &  136 &  230 &  350 &  499 \\
-    ROADP      & 15   & 1211 & 1633 & 2023 & 2422 \\
     SP         & 16   &  130 &  222 &  339 &  486 \\
-    RSP        & 15   & 1204 & 1624 & 2012 & 2409 \\
     AP         & 34   &  130 &  222 &  339 &  486 \\
-    RAP        & 18   & 1204 & 1624 & 2012 & 2409 \\
     \hline
@@
     OADP       & 0.23  & 0.15 & 0.11 & 0.08 & 0.07 \\
-    ROADP      & 0.23  & 0.15 & 0.12 & 0.08 & 0.08 \\
     SP         & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
-    RSP        & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
     AP         & 0.46  & 0.37 & 0.31 & 0.26 & 0.22 \\
-    RAP        & 3.02  & 2.93 & 2.54 & 2.15 & 1.91 \\
     \hline
@@
     OADP       & 4.43  & 4.06 & 3.77 & 3.55 & 3.20 \\
-    ROADP      & 4.44  & 4.06 & 3.76 & 3.55 & 3.20 \\
     SP         & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
-    RSP        & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
     AP         & 4.59  & 4.10 & 3.76 & 3.57 & 3.24 \\
-    RAP        & 4.23  & 3.87 & 3.55 & 3.38 & 3.07 \\
     \hline
@@
     OADP       &  73  & 108 &  156 &  189 &  282 \\
-    ROADP      &  73  & 108 &  159 &  183 &  281 \\
     SP         &  44  &  64 &   84 &  103 &  131 \\
-    RSP        &  43  &  63 &   83 &  101 &  129 \\
     AP         &  66  &  81 &  100 &  117 &  146 \\
-    RAP        &  22  &  15 &   15 &   21 &   24 \\
     \hline
@@
 \end{table}
+
+\begin{figure}[p]
+  \centering
+  \includegraphics[width=0.98\textwidth]{ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_sp}
+  \caption{EUR samples in UK Biobank data by SP}
+\end{figure}
+
+\begin{figure}[p]
+  \centering
+  \includegraphics[width=0.98\textwidth]{ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_ap}
+  \caption{EUR samples in UK Biobank data by AP}
+\end{figure}
+
+\begin{figure}[p]
+  \centering
+  \includegraphics[width=0.98\textwidth]{ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_oadp}
+  \caption{EUR samples in UK Biobank data by oadp}
+\end{figure}
+
+\begin{figure}[p]
+  \centering
+  \includegraphics[width=0.98\textwidth]{ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_adp}
+  \caption{EUR samples in UK Biobank data by adp}
+\end{figure}
 

### 2018-06-18 frugalpca.py
Change from multiprocessing to pbs array job

@@ -53,6 +53,37 @@ DELIMITER = '\t'
 DIR_TMP = 'tmp'
 
 
+# def split_bed_indiv(filepref, chunklevel):
+#     fam = np.loadtxt(filepref+'.fam')
+#     n = fam.shape[0]
+#     if n > chunklevel:
+#         fam_0 = fam[:(n/2),:]
+#         fam_1 = fam[(n/2):,:]
+#         split_fam()
+#         extract_indivs(0)
+#         extract_indivs(1)
+#         split_bed_indiv(0)
+#         split_bed_indiv(1)
+
+# def submit_pca(pref_ref, pref_stu, popu_ref_filename=None, popu_ref_k=None, method='oadp', dim_ref=DIM_REF, dim_stu=DIM_STU, dim_stu_high=DIM_STU_HIGH, use_memmap=False, load_saved_ref_decomp=True, log_level='info'):
+#     submit_pca_on_flux
+
+
+# def run_pca_parallel(ref_filepref, stu_filepref, popu_ref_filename=None, popu_ref_k=None, chunksize=None, method='oadp', dim_ref=DIM_REF, dim_stu=DIM_STU, dim_stu_high=DIM_STU_HIGH, use_memmap=False, load_saved_ref_decomp=True, log_level='info'):
+#     stu_filepref_chunk_list = split_bed_indiv(stu_filepref, chunksize)
+#     n_chunks = len(stu_filepref_chunk_list)
+#     stu_pcspopu_filepref_list = np.chararray((n_chunks, 2))
+#     for i,stu_filepref_chunk in enumerate(stu_filepref_chunk_list):
+#         submit_pca(ref_filepref, stu_filepref_chunk, popu_ref_filename=None, popu_ref_k=None, method='oadp', dim_ref=DIM_REF, dim_stu=DIM_STU, dim_stu_high=DIM_STU_HIGH, use_memmap=False, load_saved_ref_decomp=True, log_level='info')
+#         stu_pcspopu_filename_list[i,:2] = [
+#             stu_filepref_chunk+'_stu_'+method+'.pcs',
+#             stu_filepref_chunk+'.popu']
+#     stu_pcs_filename = merge_pcs(stu_pcspopu_filename_list[:,0])
+#     stu_popu_filename = merge_popu(stu_pcspopu_filename_list[:,1])
+#     ref_pcs_filename = ref_filepref + '.pcs'
+#     ref_popu_filename = ref_filepref + '.popu'
+#     return variables, filenames
+
[…]

### 2018-06-18 frugalpca_runner.py

@@ -9,9 +9,10 @@ print('Number of cores: ' + str(NUM_CORES))
 # fpt.test_pca_subpopu(pref_ref, pref_stu, 'EUR', cmp_trace=True)
 
 # fpt.test_pca_ggsim()
-fpt.test_pca_5c()
+# fpt.test_pca_5c()
 # fpt.test_pca_5c_EUR()
 # Parallel(n_jobs=NUM_CORES)(delayed(fpt.convert_ggsim)(i) for i in range(5))
 # fpt.convert_ggsim(4)
 # fpt.test_pca_ggsim()
+fpt.test_merge_array_results()
 

### 2018-06-18 frugalpca_tests.py

@@ -164,15 +164,9 @@ def test_pca(pref_ref, pref_stu, cmp_trace=True, load_results=False, assert_resu
         popu_stu_pred_ap = np.loadtxt(pref_out + '_pred_ap.popu', dtype=np.object)[:,2]
         popu_stu_pred_oadp = np.loadtxt(pref_out + '_pred_ap.popu', dtype=np.object)[:,2]
     else:
-        pcs_ref, pcs_stu_sp, popu_ref, popu_stu_pred_sp =  fp.run_pca(
-            pref_ref, pref_stu, popu_filename_ref = popu_filename_ref,
-            method='sp', use_memmap=use_memmap, load_saved_ref_decomp=load_saved_ref_decomp, log_level=log_level)[:4]
-        pcs_ref, pcs_stu_ap, popu_ref, popu_stu_pred_ap =  fp.run_pca(
-            pref_ref, pref_stu, popu_filename_ref = popu_filename_ref,
-            method='ap', use_memmap=use_memmap, load_saved_ref_decomp=load_saved_ref_decomp, log_level=log_level)[:4]
-        pcs_ref, pcs_stu_oadp, popu_ref, popu_stu_pred_oadp =  fp.run_pca(
-            pref_ref, pref_stu, popu_filename_ref = popu_filename_ref,
-            method='oadp', use_memmap=use_memmap, load_saved_ref_decomp=load_saved_ref_decomp, log_level=log_level)[:4]
+        pcs_ref, pcs_stu_sp, popu_ref, popu_stu_pred_sp = fp.run_pca(pref_ref, pref_stu, method='sp')[:4]
+        pcs_ref, pcs_stu_ap, popu_ref, popu_stu_pred_ap = fp.run_pca(pref_ref, pref_stu, method='ap')[:4]
+        pcs_ref, pcs_stu_oadp, popu_ref, popu_stu_pred_oadp = fp.run_pca(pref_ref, pref_stu, method='oadp')[:4]
 
     method_list = ['sp', 'ap', 'oadp']
     pcs_stu_list = [pcs_stu_sp, pcs_stu_ap, pcs_stu_oadp]
@@ -229,23 +223,16 @@ def test_pca_subpopu(pref_ref, pref_stu, popu_name_this, cmp_trace=True, load_re
 
[…]

### 2018-06-19 frugalpca.py
Finish parallel submission on flux, add uniform plot limits

@@ -33,6 +33,7 @@ DIM_STU = 20
 DIM_STU_HIGH = DIM_STU * 2
 N_NEIGHBORS=5
 HDPCA_N_SPIKES_MAX = 18
+HDPCA_N_SPIKES = DIM_REF
 SAMPLE_CHUNK_SIZE_STU = 5000
 SAMPLE_SPLIT_PREF_LEN = 4
 PROCRUSTES_NITER_MAX = 10000
@@ -112,7 +113,7 @@ def create_logger(prefix='frugalpca', level='info'):
     # create file handler which logs even debug messages
     log_dir = os.path.dirname(prefix) + '/logs/'
     if not os.path.exists(log_dir):
-        os.makedirs(log_dir)
+        os.makedirs(log_dir, exist_ok=True)
     filename = log_dir + os.path.basename(prefix) + '.' + str(round(time.time())) + '.log'
     fh = logging.FileHandler(filename, 'w')
     fh.setLevel(log_level)
@@ -438,17 +439,17 @@ def load_pcs(pref, methods):
         pcs_stu_list += [pcs_stu_this]
     return pcs_ref, pcs_stu_list
 
-def adj_hdpc_shrinkage(U, s, p_ref, n_ref, hdpca_n_spikes_max=HDPCA_N_SPIKES_MAX, dim_ref=DIM_REF):
+def adj_hdpc_shrinkage(U, s, p_ref, n_ref, hdpca_n_spikes_max=HDPCA_N_SPIKES_MAX, hdpca_n_spikes=HDPCA_N_SPIKES, dim_ref=DIM_REF):
     logging.info('Estimating PC shrinkage and adjusting PC loadings...')
-    hdpc_est_result  = hdpc_est(s**2, p_ref, n_ref, n_spikes_max=hdpca_n_spikes_max)
+    # hdpc_est_result  = hdpc_est(s**2, p_ref, n_ref, n_spikes_max=hdpca_n_spikes_max)
+    hdpc_est_result  = hdpc_est(s**2, p_ref, n_ref, n_spikes=hdpca_n_spikes)
     shrinkage = np.array(hdpc_est_result[-1])
     n_pc_adjusted = min(dim_ref, len(shrinkage))
     for i in range(n_pc_adjusted):
-        U[:, i] /= shrinkage[i]
+        U[:, i] /= shrinkage[i] # TODO: Adjust angles too?
 
 
[…]

### 2018-06-19 frugalpca_runner.py

@@ -13,6 +13,8 @@ print('Number of cores: ' + str(NUM_CORES))
 # fpt.test_pca_5c_EUR()
 # Parallel(n_jobs=NUM_CORES)(delayed(fpt.convert_ggsim)(i) for i in range(5))
 # fpt.convert_ggsim(4)
-# fpt.test_pca_ggsim()
-fpt.test_merge_array_results()
+for i in range(5):
+    fpt.test_pca_ggsim(i)
+# fpt.test_merge_array_results()
+# fpt.test_split_bed_indiv()
 

### 2018-06-19 frugalpca_tests.py

@@ -4,23 +4,6 @@ import frugalpca as fp
 import subprocess
 import os.path
 
-# def run_tests():
-#     # pref_ref = '../data/kgn/kgn_bial_orphans_snps_ukb_snpscap_ukb'
-#     # pref_stu = '../data/ukb/ukb_snpscap_kgn_bial_orphans'
-#     # pref_stu = '../data/ukb/ukb_snpscap_kgn_bial_orphans_5c'
-#     # pref_stu = '../data/ukb/old/ukb_snps_kgn' # Old, don't use except for testing plotting
-#     # cmp_trace = False
-#     # test_standardize()
-#     # test_online_svd_procrust()
-#     # test_pca(pref_ref, pref_stu, cmp_trace=cmp_trace)
-#     # test_pca_subpopu(pref_ref, pref_stu, 'EUR')
-#     # test_pca_subpopu(pref_ref, pref_stu, 'EAS')
-#     # test_pca_subpopu(pref_ref, pref_stu, 'SAS')
-#     # test_pca_subpopu(pref_ref, pref_stu, 'AMR')
-#     # test_pca_subpopu(pref_ref, pref_stu, 'AFR')
-#     # plot_results(pref_ref, pref_stu)
-
-
 def test_standardize():
     print('Testing standardize...')
     x = np.array([[0.0,1.0,2.0],
@@ -141,14 +124,16 @@ def test_online_svd_procrust():
     print('Passed!')
 
 
-def test_pca(pref_ref, pref_stu, cmp_trace=True, load_results=False, assert_results=False):
+def test_pca(pref_ref, pref_stu, cmp_trace=True, load_results=False, assert_results=False, pref_out_trace=None, dim_ref=4):
     popu_filename_ref = pref_ref + '.popu'
     subpopu_filename_ref = pref_ref + '_sub.popu'
     base_ref = os.path.basename(pref_ref)
     dir_stu = os.path.dirname(pref_stu)
     pref_out = pref_stu + '_sturef_' + base_ref
-    pcs_trace_ref_filename = pref_out + '.RefPC.coord'
-    pcs_trace_stu_filename = pref_out + '.ProPC.coord'
+    if pref_out_trace is None:
+        pref_out_trace = pref_out
[…]

### 2018-06-20 doc/paper/paper.tex
Finish results for simulation study

@@
 
-For the simulation study
-(\Cref{tbl:nChg-runtimes-ref}, \Cref{tbl:nChg-runtimes-study},
-\Cref{tbl:nChg-accuracy-gold}, \Cref{tbl:nChg-accuracy-ctr},
-\Cref{tbl:mChg-runtimes-study}),
-we see that ADP, which is the most reliable method,
-is the slowest in study runtimes.
-By using the online SVD algorithm,
-OADP achieved the highest accuracy (excluding ADP, which is the golden standard) and the fastest runtime.
-SP is fast, also,
-but its accuracy is the lowest.
-By using the HDPCA package,
-AP reduces the error of SP
-and achieved the second highest in accuracy.
-However, as the reference size increases,
-the accuracies of different methods become closer and closer (\Cref{fig:nChg}).
-As for the speed,
-ADP's study runtime increases so fast with the reference size
-that the study runtimes for the other methods seem to be constant in comparison (\Cref{fig:nChg}).
-On the other hand,
-the study runtimes appear to increase linearly with respect to the study size (\Cref{fig:mChg}).
-This information can be used to predict the runtime for larger study samples.
+For the simulation study,
+the PC scores predicted by AP, OADP, and ADP
+are very close to each other
+for every choice of reference sample size.
+Compared to their results,
+SP exhibits a strong shrinkage
+when the reference sample size is 1000.
+As we gradually increases the reference sample size to 3000
+by an increment of 500,
+the magnitude of the shrinkage decreases
+but is still visible.
+This is consistent with the theoretical result
+that the shrinkage bias decreases
+as the ratio of
+the reference sample dimension to the reference sample size
+decreases \citep{dey2016asymptotic}.
+
+For computational cost,
+we can see that the runtime for SP increases faster than linearly
+as the reference sample size increases.
+This is consistent
+with the theoretical computational complexity of SP,
+which expects the runtime to increase at the rate of $\bO[n^3]$.
+In comparison, the runtimes of SP, AP, and OADP
+increase only slightly.
+The theoretical computational complexity of SP and AP is $\bO[1]$
+and that of OADP is $\bO[n]$,
+when the data dimension and the study sample size are fixed.
 
@@
 
-When we used the global samples,
+When we used the global samples in the UK Biobank data,
 we found that the PC scores predicted by the four methods' are almost identical.
@@
 
-However, when we focus on the European (EUR) samples only,
+However, when we focused on the European (EUR) samples only,
 the results given by SP were clearly unreliable.
@@
 \appendix
-
-\section{Randomized SVD}\label{sec:randsvd}
-
-% In addition to estimating the PC scores of the study individuals,
-% we have another algorithm that aims at improving the decomposition of the
-% training data.
-% For SP, AP, ADP, and OADP, we need to find the covariance matrix of the training
-% data, which has dimension $n \times n$, and conduct eigendecomosition on it.
-% However, when $n$ is exceedingly large,
-% computation and memory cost of PCA on the reference samples can be enormous.
-% A randomized SVD (RSVD) algorithm has been developed to handle this issue.
-% When the reference sample size is small, RSVD can be slower than standard SVD,
-% but it is scalable to large sample size.
-% For the details of the RSVD algorithm, see \citet{halko2011finding}.
-
-% Notice that the randomized SVD method on the training data
-% only calculates the first $k$ eigenvalues.
-% However, in order to use adjusted projection,
-% all the eigenvalues of the training data are required.
-% Thus for the missing eigenvalues,
-% we predict them by regressing their log-scale values on the ranks
-% and using the first $k$ eigenvalues for training the linear regression.
-
-\newpage
-
 \bibliographystyle{plainnat}
@@
 \newpage
-
 \section{Tables and Figures}
 
-\begin{table} 
+\begin{table}[h]
   \centering
@@
 
-\begin{table} 
-  \centering
-  \begin{tabular}{|l|l|l|l|l|}
-    \hline
-    Method & Reference Runtimes (n=2492) & Study Runtimes (m=1000)& Golden Standard Error \\ 
-    \hline
-    ADP & 411 & 29813 & 0.00 \\
-    \hline
-    OADP & 2518 & 103 & 0.40 \\
-    \hline
-    SP & 411 & 425 & 1.95 \\
-    \hline
-    AP & 411 & 482 & 1.12 \\
-    \hline
-  \end{tabular}
-  \caption{
-    Comparison of runtimes and accuracy for different methods applied to the real-life datasets.
-    The reference individuals are from the 1000 Genomes dataset with 2492 unrelated individuals.
-    The study sample contain 1000 individuals from the UKBioBank dataset.
-    There are about 125,000 loci shared by the reference and study data.
-    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADP.
-  }
-  \label{tbl:ukb}
-\end{table}
-
-\begin{table}
+\begin{table}[h]
   \centering
@@
     \hline
-    ADP        &  68  & 1263 & 2871 & 5720 & 9737 \\
-    OADP       &  72  & 18   & 19   & 22   & 24   \\
-    SP         &  68  & 26   & 42   & 67   & 96   \\
-    AP         &  68  & 69   & 107  & 91   & 119  \\
+    \multicolumn{6}{|c|}{Study Runtime} \\
+    \hline
+    SP         &  17  & 17   & 16   & 18   & 18   \\
+    AP         &  18  & 16   & 18  & 16   & 17  \\
+    OADP       &  145  & 146   & 147   & 148   & 149   \\
+    ADP        &  379  & 809 & 1526 & 2556 & 4062 \\
+    \hline
+    \multicolumn{6}{|c|}{Euclidean Error} \\
+    \hline
+    SP         &  188  & 139   & 112   & 91   & 84   \\
+    AP         &  31  & 27   & 12  & 11   & 14  \\
+    OADP       &  21  & 16   & 12   & 9   & 10   \\
+    ADP        &  0  & 0 & 0 & 0 & 0 \\
     \hline
@@
   \caption{
-    Reference runtimes for the simulated data as the reference size increases.
-    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
-    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+    The runtimes and errors of the four methods
+    as the reference size increases.
+    Errors are measured as the Euclidean distance of
+    The number of variants is 100,000,
+    and the study sample size is 200.
+    Only the top 4 PCs are calculated.
+    The data is simulated by the GGS software \citep{mathieson2012differential}.
     }
-  \label{tbl:nChg-runtimes-ref}
 \end{table}
 
-\begin{table}
+\begin{table}[h]
   \centering
-  \begin{tabular}{|l|l|l|l|l|l|}
+  \begin{tabular}{|r|r|r|r|r|}
     \hline
-    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
+    Method & \multicolumn{2}{|c|}{Runtime} & \multicolumn{2}{|c|}{Error} \\
+    & Global & European & Global & European \\
     \hline
-    ADP        & 391  &  130 &  222 &  339 &  486 \\
-    OADP       & 15   &  136 &  230 &  350 &  499 \\
-    SP         & 16   &  130 &  222 &  339 &  486 \\
-    AP         & 34   &  130 &  222 &  339 &  486 \\
+    SP         &  62   & 48  &     & 450 \\ 
+    AP         &  65   & 48  &     & 33  \\
+    OADP       &  363  & 349 &     & 32  \\
+    ADP        &  6745 & 518 &     & 0   \\
     \hline
@@
   \caption{
-    Study runtimes for simulated data as the reference size increases.
-    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
-    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+    The study runtimes (in seconds) and erros for the four methods
+    for 500 randomly selected study samples
+    in the UK Biobank data.
+    Errors are measured as the Euclidean distance of
+    the PC scores predicted by each method
+    to those predicted by ADP.
+    The 2492 samples from the 1000 Genomes data
+    are used as the reference.
+    The number of variants in common is 145,282.
+    Only the top 4 PCs are calculated.
     }
-  \label{tbl:nChg-runtimes-study}
 \end{table}
 
-\begin{table}
+\begin{figure}[h]
   \centering
-  \begin{tabular}{|l|l|l|l|l|l|}
-    \hline
-    Study Size & 1000  & 1500 & 2000 & 2500 & 3000 \\
-    \hline
-    ADP        & 0.00  & 0.00 & 0.00 & 0.00 & 0.00 \\
-    OADP       & 0.23  & 0.15 & 0.11 & 0.08 & 0.07 \\
-    SP         & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
-    AP         & 0.46  & 0.37 & 0.31 & 0.26 & 0.22 \\
-    \hline
-  \end{tabular}
+  \includegraphics[width=0.98\textwidth]{img/ggsim_runtimes.png}
+  \includegraphics[width=0.98\textwidth]{img/ggsim_errors.png}
   \caption{
-    Golden standard error for the simulated data as the reference size increases.
-    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
-    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+    The runtimes (in seconds) and errors (Euclidean distance from ADP's PC scores) for the four methods
+    as the reference size increases.
+    The number of variants is 100,000,
+    and the study sample size is 200.
+    Only the top 4 PCs are calculated.
+    The data is simulated by the GGS software \citep{mathieson2012differential}.
   }
-  \label{tbl:nChg-accuracy-gold}
-\end{table}
+\end{figure}
 
-\begin{table}
+\begin{figure}[h]
   \centering
-  \begin{tabular}{|l|l|l|l|l|l|}
-    \hline
-    Reference Size & 1000  & 1500 & 2000 & 2500 & 3000 \\
-    \hline
-    ADP        & 4.53  & 4.13 & 3.81 & 3.59 & 3.23 \\
-    OADP       & 4.43  & 4.06 & 3.77 & 3.55 & 3.20 \\
-    SP         & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
-    AP         & 4.59  & 4.10 & 3.76 & 3.57 & 3.24 \\
-    \hline
-  \end{tabular}
-  \caption{
-    Geographic center error for the simulated data as the reference size increases.
-    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
-    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
-  }
-  \label{tbl:nChg-accuracy-ctr}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|l|l|l|l|l|l|}
-    \hline
-    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
-    \hline
-    ADP        & 507  & 758 & 1011 & 1262 & 1549 \\
-    OADP       &  73  & 108 &  156 &  189 &  282 \\
-    SP         &  44  &  64 &   84 &  103 &  131 \\
-    AP         &  66  &  81 &  100 &  117 &  146 \\
-    \hline
-  \end{tabular}
-  \caption{
-    Study runtimes for the simulated data as the study size increases.
-    The reference size is fixed to 600 and the number of SNPs is 100,000 (1000 per genealogy). 
-    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
-  }
-  \label{tbl:mChg-runtimes-study}
-\end{table}
-
-\begin{figure}[p]
-  \centering
-  \includegraphics[width=0.98\textwidth]{ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_sp}
-  \caption{EUR samples in UK Biobank data by SP}
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_sp}
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_ap}
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_oadp}
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_adp}
+  \caption{PC scores of 500 randomly selected global samples in UK Biobank data (circles) predicted by the four methods with the global samples in the 1000 Genomes data (squares) as the reference.}
 \end{figure}
@@
   \centering
-  \includegraphics[width=0.98\textwidth]{ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_ap}
-  \caption{EUR samples in UK Biobank data by AP}
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_sp}
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_ap}
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_oadp}
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_adp}
+  \caption{PC scores of the European in the 500 randomly selected global samples UK Biobank data (circles) predicted by the four methods with the European samples in the 1000 Genomes data (squares) as the reference.}
 \end{figure}
 
-\begin{figure}[p]
-  \centering
-  \includegraphics[width=0.98\textwidth]{ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_oadp}
-  \caption{EUR samples in UK Biobank data by oadp}
-\end{figure}
+\end{document}
 
-\begin{figure}[p]
-  \centering
-  \includegraphics[width=0.98\textwidth]{ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_adp}
-  \caption{EUR samples in UK Biobank data by adp}
-\end{figure}
+% \begin{figure}[p]
+%   \centering
+%   \includegraphics[width=0.98\textwidth]{n1000}
+%   \caption{
+%     Predicted PC scores on top of reference PC scores.
+%     The four populations in the $2 \times 2$ grid simulated by the GGS software can be clearly identified.
+%     Reference individuals are in color,
+%     while predicted PC scores for the study individuals are in gray.
+%     Two measurements of deviations (accuracy) are used.
+%     The reference center error is the square root of the average squared distance between the study individuals and the population centers, which are the means of the PC scores of the reference individuals by the populations.
+%     The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADP.
+%     The reference size is 1000 and the study size is 200.
+%     There are 100,000 loci (1000 per genealogy) and the migration rate is 100.
+%   }
+%   \label{fig:n1000}
+% \end{figure}
 
-\begin{figure}[p]
-  \centering
-  \includegraphics[width=0.98\textwidth]{n1000}
-  \caption{
-    Predicted PC scores on top of reference PC scores.
-    The four populations in the $2 \times 2$ grid simulated by the GGS software can be clearly identified.
-    Reference individuals are in color,
-    while predicted PC scores for the study individuals are in gray.
-    Two measurements of deviations (accuracy) are used.
-    The reference center error is the square root of the average squared distance between the study individuals and the population centers, which are the means of the PC scores of the reference individuals by the populations.
-    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADP.
-    The reference size is 1000 and the study size is 200.
-    There are 100,000 loci (1000 per genealogy) and the migration rate is 100.
-  }
-  \label{fig:n1000}
-\end{figure}
+% \begin{figure}[p]
+%   \centering
+%   \includegraphics[width=0.98\textwidth]{ukb}
+%   \caption{
+%     The reference individuals are from the 1000 Genomes dataset with 2492 unrelated individuals from 5 superpopulations: African, American, East Asian, European, and South Asian. 
+%     The study sample contain 1000 individuals from the UKBioBank dataset.
+%     There are about 125,000 loci shared by the reference and study data.
+%     The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADP.
+%   }
+%   \label{fig:ukb}
+% \end{figure}
 
-\begin{figure}[p]
-  \centering
-  \includegraphics[width=0.98\textwidth]{ukb}
-  \caption{
-    The reference individuals are from the 1000 Genomes dataset with 2492 unrelated individuals from 5 superpopulations: African, American, East Asian, European, and South Asian. 
-    The study sample contain 1000 individuals from the UKBioBank dataset.
-    There are about 125,000 loci shared by the reference and study data.
-    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADP.
-  }
-  \label{fig:ukb}
-\end{figure}
+% \begin{figure}[p]
+%   \centering
+%   \includegraphics[width=0.98\textwidth]{img/nChg}
+%   \caption{
+%     Runtimes and deviations for different methods as the reference sample size increases.
+%     The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
+%     Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+%   }
+%   \label{fig:nChg}
+% \end{figure}
 
-\begin{figure}[p]
-  \centering
-  \includegraphics[width=0.98\textwidth]{nChg}
-  \caption{
-    Runtimes and deviations for different methods as the reference sample size increases.
-    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
-    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
-  }
-  \label{fig:nChg}
-\end{figure}
-
-\begin{figure}[p]
-  \centering
-  \includegraphics[width=0.98\textwidth]{mChg}
-  \caption{
-    Runtimes and deviations for different methods as the study sample size increases.
-    The reference sample size is fixed to 600 and the number of SNPs is 100,000 (1000 per genealogy).
-    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
-  }
-  \label{fig:mChg}
-\end{figure}
+% \begin{figure}[p]
+%   \centering
+%   \includegraphics[width=0.98\textwidth]{img/mChg}
+%   \caption{
+%     Runtimes and deviations for different methods as the study sample size increases.
+%     The reference sample size is fixed to 600 and the number of SNPs is 100,000 (1000 per genealogy).
+%     Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+%   }
+%   \label{fig:mChg}
+% \end{figure}
 
@@
 
-\newpage
-
 % \includegraphics[width=0.98\textwidth]{kgn_allChr_ukb_orphans_ukb_1k_rand}
 
-\newpage
+% \begin{table} 
+%   \centering
+%   \begin{tabular}{|l|l|l|l|l|}
+%     \hline
+%     Method & Reference Runtimes (n=2492) & Study Runtimes (m=1000)& Golden Standard Error \\ 
+%     \hline
+%     ADP & 411 & 29813 & 0.00 \\
+%     \hline
+%     OADP & 2518 & 103 & 0.40 \\
+%     \hline
+%     SP & 411 & 425 & 1.95 \\
+%     \hline
+%     AP & 411 & 482 & 1.12 \\
+%     \hline
+%   \end{tabular}
+%   \caption{
+%     Comparison of runtimes and accuracy for different methods applied to the real-life datasets.
+%     The reference individuals are from the 1000 Genomes dataset with 2492 unrelated individuals.
+%     The study sample contain 1000 individuals from the UKBioBank dataset.
+%     There are about 125,000 loci shared by the reference and study data.
+%     The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADP.
+%   }
+%   \label{tbl:ukb}
+% \end{table}
 
-\end{document}
+% \begin{table}
+%   \centering
+%   \begin{tabular}{|l|l|l|l|l|l|}
+%     \hline
+%     Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
+%     \hline
+%     ADP        &  68  & 1263 & 2871 & 5720 & 9737 \\
+%     OADP       &  72  & 18   & 19   & 22   & 24   \\
+%     SP         &  68  & 26   & 42   & 67   & 96   \\
+%     AP         &  68  & 69   & 107  & 91   & 119  \\
+%     \hline
+%   \end{tabular}
+%   \caption{
+%     Reference runtimes for the simulated data as the reference size increases.
+%     The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
+%     Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+%   }
+%   \label{tbl:nChg-runtimes-ref}
+% \end{table}
+
+% \begin{table}
+%   \centering
+%   \begin{tabular}{|l|l|l|l|l|l|}
+%     \hline
+%     Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
+%     \hline
+%     ADP        & 391  &  130 &  222 &  339 &  486 \\
+%     OADP       & 15   &  136 &  230 &  350 &  499 \\
+%     SP         & 16   &  130 &  222 &  339 &  486 \\
+%     AP         & 34   &  130 &  222 &  339 &  486 \\
+%     \hline
+%   \end{tabular}
+%   \caption{
+%     Study runtimes for simulated data as the reference size increases.
+%     The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
+%     Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+%   }
+%   \label{tbl:nChg-runtimes-study}
+% \end{table}
+
+% \begin{table}
+%   \centering
+%   \begin{tabular}{|l|l|l|l|l|l|}
+%     \hline
+%     Study Size & 1000  & 1500 & 2000 & 2500 & 3000 \\
+%     \hline
+%     ADP        & 0.00  & 0.00 & 0.00 & 0.00 & 0.00 \\
+%     OADP       & 0.23  & 0.15 & 0.11 & 0.08 & 0.07 \\
+%     SP         & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
+%     AP         & 0.46  & 0.37 & 0.31 & 0.26 & 0.22 \\
+%     \hline
+%   \end{tabular}
+%   \caption{
+%     Golden standard error for the simulated data as the reference size increases.
+%     The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
+%     Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+%   }
+%   \label{tbl:nChg-accuracy-gold}
+% \end{table}
+
+% \begin{table}
+%   \centering
+%   \begin{tabular}{|l|l|l|l|l|l|}
+%     \hline
+%     Reference Size & 1000  & 1500 & 2000 & 2500 & 3000 \\
+%     \hline
+%     ADP        & 4.53  & 4.13 & 3.81 & 3.59 & 3.23 \\
+%     OADP       & 4.43  & 4.06 & 3.77 & 3.55 & 3.20 \\
+%     SP         & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
+%     AP         & 4.59  & 4.10 & 3.76 & 3.57 & 3.24 \\
+%     \hline
+%   \end{tabular}
+%   \caption{
+%     Geographic center error for the simulated data as the reference size increases.
+%     The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
+%     Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+%   }
+%   \label{tbl:nChg-accuracy-ctr}
+% \end{table}
+
+% \begin{table}
+%   \centering
+%   \begin{tabular}{|l|l|l|l|l|l|}
+%     \hline
+%     Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
+%     \hline
+%     ADP        & 507  & 758 & 1011 & 1262 & 1549 \\
+%     OADP       &  73  & 108 &  156 &  189 &  282 \\
+%     SP         &  44  &  64 &   84 &  103 &  131 \\
+%     AP         &  66  &  81 &  100 &  117 &  146 \\
+%     \hline
+%   \end{tabular}
+%   \caption{
+%     Study runtimes for the simulated data as the study size increases.
+%     The reference size is fixed to 600 and the number of SNPs is 100,000 (1000 per genealogy). 
+%     Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
+%   }
+%   \label{tbl:mChg-runtimes-study}
+% \end{table}
+
+% \section{Randomized SVD}\label{sec:randsvd}
+
+% In addition to estimating the PC scores of the study individuals,
+% we have another algorithm that aims at improving the decomposition of the
+% training data.
+% For SP, AP, ADP, and OADP, we need to find the covariance matrix of the training
+% data, which has dimension $n \times n$, and conduct eigendecomosition on it.
+% However, when $n$ is exceedingly large,
+% computation and memory cost of PCA on the reference samples can be enormous.
+% A randomized SVD (RSVD) algorithm has been developed to handle this issue.
+% When the reference sample size is small, RSVD can be slower than standard SVD,
+% but it is scalable to large sample size.
+% For the details of the RSVD algorithm, see \citet{halko2011finding}.
+
+% Notice that the randomized SVD method on the training data
+% only calculates the first $k$ eigenvalues.
+% However, in order to use adjusted projection,
+% all the eigenvalues of the training data are required.
+% Thus for the missing eigenvalues,
+% we predict them by regressing their log-scale values on the ranks
+% and using the first $k$ eigenvalues for training the linear regression.

### 2018-06-20 doc/paper/plot.py

@@ -1,0 +1,23 @@
+import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+
+methods = ['sp', 'ap', 'oadp', 'adp']
+n = np.arange(5, dtype=int) * 500 + 1000
+
+ggsim = np.loadtxt('ggsim_runtimes.dat')
+for i in range(4):
+    plt.plot(n, ggsim[i], label=methods[i])
+plt.xticks(n)
+plt.legend()
+plt.savefig('img/ggsim_runtimes.png', dpi=300)
+plt.close('all')
+
+ggsim = np.loadtxt('ggsim_errors.dat')
+for i in range(4):
+    plt.plot(n, ggsim[i], label=methods[i])
+plt.xticks(n)
+plt.legend()
+plt.savefig('img/ggsim_errors.png', dpi=300)
+plt.close('all')
+

### 2018-06-20 frugalpca.py
Add plot_dim

@@ -449,7 +449,7 @@ def adj_hdpc_shrinkage(U, s, p_ref, n_ref, hdpca_n_spikes_max=HDPCA_N_SPIKES_MAX
         U[:, i] /= shrinkage[i] # TODO: Adjust angles too?
 
 
-def plot_pcs(pcs_ref, pcs_stu_list, popu_ref, popu_stu_list, method_list, out_pref, markers=PLOT_MARKERS, alpha_ref=PLOT_ALPHA_REF, alpha_stu=PLOT_ALPHA_STU, plot_lim=None):
+def plot_pcs(pcs_ref, pcs_stu_list, popu_ref, popu_stu_list, method_list, out_pref, markers=PLOT_MARKERS, alpha_ref=PLOT_ALPHA_REF, alpha_stu=PLOT_ALPHA_STU, plot_lim=None, plot_dim=float('inf')):
     if type(pcs_stu_list) is not list:
         pcs_stu_list = [pcs_stu_list]
     if type(popu_stu_list) is not list:
@@ -461,7 +461,7 @@ def plot_pcs(pcs_ref, pcs_stu_list, popu_ref, popu_stu_list, method_list, out_pr
     popu_n = len(popu_unique)
     plot_colors_unique = plt.rcParams['axes.prop_cycle'].by_key()['color']
     dim_ref = pcs_ref.shape[1]
-    n_subplot = dim_ref // 2
+    n_subplot = int(min(dim_ref, plot_dim) / 2)
     fig, ax = plt.subplots(ncols=n_subplot)
     for j in range(n_subplot):
         plt.subplot(1, n_subplot, j+1)

### 2018-06-20 frugalpca_runner.py

@@ -8,13 +8,14 @@ print('Number of cores: ' + str(NUM_CORES))
 # fpt.convert()
 # fpt.test_pca_subpopu(pref_ref, pref_stu, 'EUR', cmp_trace=True)
 
-# fpt.test_pca_ggsim()
-# fpt.test_pca_5c()
-# fpt.test_pca_5c_EUR()
 # Parallel(n_jobs=NUM_CORES)(delayed(fpt.convert_ggsim)(i) for i in range(5))
 # fpt.convert_ggsim(4)
-for i in range(5):
-    fpt.test_pca_ggsim(i)
+# for i in range(5):
+#     fpt.test_pca_ggsim(i)
 # fpt.test_merge_array_results()
 # fpt.test_split_bed_indiv()
 
+
+fpt.test_pca_ggsim()
+# fpt.test_pca_5c()
+# fpt.test_pca_5c_EUR()

### 2018-06-20 frugalpca_tests.py

@@ -124,7 +124,7 @@ def test_online_svd_procrust():
     print('Passed!')
 
 
-def test_pca(pref_ref, pref_stu, cmp_trace=True, load_results=False, assert_results=False, pref_out_trace=None, dim_ref=4):
+def test_pca(pref_ref, pref_stu, cmp_trace=True, load_results=False, assert_results=False, pref_out_trace=None, dim_ref=4, plot_dim=2):
     popu_filename_ref = pref_ref + '.popu'
     subpopu_filename_ref = pref_ref + '_sub.popu'
     base_ref = os.path.basename(pref_ref)
@@ -156,9 +156,9 @@ def test_pca(pref_ref, pref_stu, cmp_trace=True, load_results=False, assert_resu
     pcs_min = np.vstack((pcs_ref, pcs_stu_sp, pcs_stu_ap, pcs_stu_oadp)).min(axis=0)
     pcs_max = np.vstack((pcs_ref, pcs_stu_sp, pcs_stu_ap, pcs_stu_oadp)).max(axis=0)
     plot_lim = np.vstack((pcs_min, pcs_max)) * 1.20
-    fp.plot_pcs(pcs_ref, pcs_stu_sp, popu_ref, popu_stu_pred_sp, 'sp', out_pref=pref_out, plot_lim=plot_lim)
-    fp.plot_pcs(pcs_ref, pcs_stu_ap, popu_ref, popu_stu_pred_ap, 'ap', out_pref=pref_out, plot_lim=plot_lim)
-    fp.plot_pcs(pcs_ref, pcs_stu_oadp, popu_ref, popu_stu_pred_oadp, 'oadp', out_pref=pref_out, plot_lim=plot_lim)
+    fp.plot_pcs(pcs_ref, pcs_stu_sp, popu_ref, popu_stu_pred_sp, 'sp', out_pref=pref_out, plot_lim=plot_lim, plot_dim=plot_dim)
+    fp.plot_pcs(pcs_ref, pcs_stu_ap, popu_ref, popu_stu_pred_ap, 'ap', out_pref=pref_out, plot_lim=plot_lim, plot_dim=plot_dim)
+    fp.plot_pcs(pcs_ref, pcs_stu_oadp, popu_ref, popu_stu_pred_oadp, 'oadp', out_pref=pref_out, plot_lim=plot_lim, plot_dim=plot_dim)
 
     method_list = ['sp', 'ap', 'oadp']
[…]

### 2018-06-20 doc/paper/paper.tex
Add fine-tuned plots to paper

@@
 
-\usepackage{epsfig} \usepackage[margin=1.5in]{geometry}
+\usepackage{epsfig}
+\usepackage[margin=1in]{geometry}
 \usepackage[round]{natbib}
@@
 \usepackage{graphicx}
+\usepackage{subcaption}
 
@@
 
+\begin{figure}[p]
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim1000_100000_200_2_1_100_1_sturef_ggsim1000_100000_1000_2_1_100_0_sp}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim1000_100000_200_2_1_100_1_sturef_ggsim1000_100000_1000_2_1_100_0_ap}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim1000_100000_200_2_1_100_1_sturef_ggsim1000_100000_1000_2_1_100_0_oadp}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim1000_100000_200_2_1_100_1_sturef_ggsim1000_100000_1000_2_1_100_0_adp}
+\end{subfigure}
+\hfill
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim1500_100000_200_2_1_100_1_sturef_ggsim1500_100000_1500_2_1_100_0_sp}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim1500_100000_200_2_1_100_1_sturef_ggsim1500_100000_1500_2_1_100_0_ap}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim1500_100000_200_2_1_100_1_sturef_ggsim1500_100000_1500_2_1_100_0_oadp}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim1500_100000_200_2_1_100_1_sturef_ggsim1500_100000_1500_2_1_100_0_adp}
+\end{subfigure}
+\hfill
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim2000_100000_200_2_1_100_1_sturef_ggsim2000_100000_2000_2_1_100_0_sp}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim2000_100000_200_2_1_100_1_sturef_ggsim2000_100000_2000_2_1_100_0_sp}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim2000_100000_200_2_1_100_1_sturef_ggsim2000_100000_2000_2_1_100_0_oadp}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim2000_100000_200_2_1_100_1_sturef_ggsim2000_100000_2000_2_1_100_0_adp}
+\end{subfigure}
+\hfill
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim2500_100000_200_2_1_100_1_sturef_ggsim2500_100000_2500_2_1_100_0_sp}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim2500_100000_200_2_1_100_1_sturef_ggsim2500_100000_2500_2_1_100_0_ap}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim2500_100000_200_2_1_100_1_sturef_ggsim2500_100000_2500_2_1_100_0_oadp}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim2500_100000_200_2_1_100_1_sturef_ggsim2500_100000_2500_2_1_100_0_adp}
+\end{subfigure}
+\hfill
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_sp}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_sp}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_oadp}
+\end{subfigure}%
+\begin{subfigure}{.25\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_adp}
+\end{subfigure}
+\caption{
+  The PC scores predicted by each of the four methods
+  as the reference size varies.
+  Reference samples are in colors
+  and study samples are in black.
+  The number of variants is 100,000,
+  and the study sample size is 200.
+  Only the top 4 PCs are calculated.
+  The data is simulated by the GGS software \citep{mathieson2012differential}.
+}
+\end{figure}
+
 \begin{table}[h]
@@
     \hline
-    \multicolumn{6}{|c|}{Euclidean Error} \\
+    \multicolumn{6}{|c|}{Error form ADP} \\
     \hline
@@
 
+\begin{figure}[h]
+
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \includegraphics[width=.98\linewidth]{img/ggsim_runtimes.png}
+  \end{subfigure}
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \includegraphics[width=.98\linewidth]{img/ggsim_errors.png}
+  \end{subfigure}
+  \caption{
+    The runtimes (in seconds) and errors (Euclidean distance from ADP's PC scores) for the four methods
+    as the reference size increases.
+    The number of variants is 100,000,
+    and the study sample size is 200.
+    Only the top 4 PCs are calculated.
+    The data is simulated by the GGS software \citep{mathieson2012differential}.
+  }
+\end{figure}
+
 \begin{table}[h]
@@
     Method & \multicolumn{2}{|c|}{Runtime} & \multicolumn{2}{|c|}{Error} \\
+    \hline
     & Global & European & Global & European \\
     \hline
-    SP         &  62   & 48  &     & 450 \\ 
-    AP         &  65   & 48  &     & 33  \\
-    OADP       &  363  & 349 &     & 32  \\
-    ADP        &  6745 & 518 &     & 0   \\
+    SP         &  62   & 48  & 92  & 450 \\ 
+    AP         &  65   & 48  & 48  & 33  \\
+    OADP       &  363  & 349 & 10  & 32  \\
+    ADP        &  6739 & 518 & 0   & 0   \\
     \hline
@@
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.98\textwidth]{img/ggsim_runtimes.png}
-  \includegraphics[width=0.98\textwidth]{img/ggsim_errors.png}
-  \caption{
-    The runtimes (in seconds) and errors (Euclidean distance from ADP's PC scores) for the four methods
-    as the reference size increases.
-    The number of variants is 100,000,
-    and the study sample size is 200.
-    Only the top 4 PCs are calculated.
-    The data is simulated by the GGS software \citep{mathieson2012differential}.
-  }
-\end{figure}
-
-\begin{figure}[h]
+\begin{figure}[p]
   \centering

### 2018-06-20 doc/paper/paper.tex
Add all the plots to paper

@@
 \begin{figure}[p]
-\begin{subfigure}{.25\textwidth}
+\begin{subfigure}{.5\textwidth}
   \centering
@@
 \end{subfigure}%
-\begin{subfigure}{.25\textwidth}
+\begin{subfigure}{.5\textwidth}
   \centering
   \includegraphics[width=.98\linewidth]{img/ggsim1000_100000_200_2_1_100_1_sturef_ggsim1000_100000_1000_2_1_100_0_ap}
-\end{subfigure}%
-\begin{subfigure}{.25\textwidth}
+\end{subfigure}
+\hfill
+\begin{subfigure}{.5\textwidth}
   \centering
@@
 \end{subfigure}%
-\begin{subfigure}{.25\textwidth}
+\begin{subfigure}{.5\textwidth}
   \centering
@@
 \end{subfigure}
-\hfill
-\begin{subfigure}{.25\textwidth}
+\end{figure}
+
+\begin{figure}[p]
+\begin{subfigure}{.5\textwidth}
   \centering
@@
 \end{subfigure}%
-\begin{subfigure}{.25\textwidth}
+\begin{subfigure}{.5\textwidth}
   \centering
   \includegraphics[width=.98\linewidth]{img/ggsim1500_100000_200_2_1_100_1_sturef_ggsim1500_100000_1500_2_1_100_0_ap}
-\end{subfigure}%
-\begin{subfigure}{.25\textwidth}
+\end{subfigure}
+\hfill
+\begin{subfigure}{.5\textwidth}
   \centering
@@
 \end{subfigure}%
-\begin{subfigure}{.25\textwidth}
-  \centering
-  \includegraphics[width=.98\linewidth]{img/ggsim2000_100000_200_2_1_100_1_sturef_ggsim2000_100000_2000_2_1_100_0_sp}
-\end{subfigure}%
-\begin{subfigure}{.25\textwidth}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim2000_100000_200_2_1_100_1_sturef_ggsim2000_100000_2000_2_1_100_0_ap}
+\end{subfigure}
+\hfill
+\begin{subfigure}{.5\textwidth}
   \centering
@@
 \end{subfigure}
-\hfill
-\begin{subfigure}{.25\textwidth}
+\end{figure}
+
+\begin{figure}
+\begin{subfigure}{.5\textwidth}
   \centering
@@
 \end{subfigure}%
-\begin{subfigure}{.25\textwidth}
+\begin{subfigure}{.5\textwidth}
   \centering
   \includegraphics[width=.98\linewidth]{img/ggsim2500_100000_200_2_1_100_1_sturef_ggsim2500_100000_2500_2_1_100_0_ap}
-\end{subfigure}%
-\begin{subfigure}{.25\textwidth}
+\end{subfigure}
+\hfill
+\begin{subfigure}{.5\textwidth}
   \centering
@@
 \end{subfigure}%
-\begin{subfigure}{.25\textwidth}
+\begin{subfigure}{.5\textwidth}
   \centering
   \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_sp}
-\end{subfigure}%
-\begin{subfigure}{.25\textwidth}
+\end{subfigure}
+\hfill
+\begin{subfigure}{.5\textwidth}
   \centering
@@
   and study samples are in black.
+  The stars mark the centers of mass (geographic centers) of each population for the reference and study samples.
   The number of variants is 100,000,
@@
     \hline
-    \multicolumn{6}{|c|}{Error form ADP} \\
+    \multicolumn{6}{|c|}{Distance from ADP's results} \\
     \hline
@@
     \hline
+    \multicolumn{6}{|c|}{Distance from reference centers} \\
+    \hline
+    SP         &  335 & 161  & 113 & 71  & 60  \\
+    AP         &  31  & 19   & 17  & 25  & 15  \\
+    OADP       &  12  & 17   & 11  & 20  & 12  \\
+    ADP        &  20  & 14   & 14  & 23  & 14  \\
+    \hline
   \end{tabular}
@@
     as the reference size increases.
-    Errors are measured as the Euclidean distance of
+    Distances are measured by the Euclidean distance.
+    Reference centers are the centers of mass of the populations
+    in the reference samples.
     The number of variants is 100,000,
@@
 \begin{figure}[h]
-
-  \begin{subfigure}{.5\textwidth}
   \centering
   \includegraphics[width=.98\linewidth]{img/ggsim_runtimes.png}
-  \end{subfigure}
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \includegraphics[width=.98\linewidth]{img/ggsim_errors.png}
-  \end{subfigure}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim_errors_adp.png}
+  \centering
+  \includegraphics[width=.98\linewidth]{img/ggsim_errors_ctr.png}
   \caption{
-    The runtimes (in seconds) and errors (Euclidean distance from ADP's PC scores) for the four methods
+    The runtimes and errors of the four methods
     as the reference size increases.
+    Runtimes are measured in seconds.
+    Errors from ADP are the Euclidean distance
+    from each method's result to ADP's result.
+    Errors between centers are the Euclidean distance
+    from the centers of mass of the reference populations
+    to the corresponding centers of mass of the study populations.
     The number of variants is 100,000,
@@
 
-\begin{table}[h]
-  \centering
-  \begin{tabular}{|r|r|r|r|r|}
-    \hline
-    Method & \multicolumn{2}{|c|}{Runtime} & \multicolumn{2}{|c|}{Error} \\
-    \hline
-    & Global & European & Global & European \\
-    \hline
-    SP         &  62   & 48  & 92  & 450 \\ 
-    AP         &  65   & 48  & 48  & 33  \\
-    OADP       &  363  & 349 & 10  & 32  \\
-    ADP        &  6739 & 518 & 0   & 0   \\
+\begin{table}[H]
+  \centering
+  \begin{tabular}{|r|r|r|r|r|r|r|}
+    \hline
+    Method & \multicolumn{2}{|c|}{Runtime} & \multicolumn{2}{|c|}{Error (ADP)} & \multicolumn{2}{|c|}{Error (centers)} \\
+    \hline
+    & Global & European & Global & European & Global & European \\
+    \hline
+    SP         &  62   & 48  & 92  & 450 & 3156 & 2304 \\ 
+    AP         &  65   & 48  & 48  & 33  & 2486 & 706  \\
+    OADP       &  363  & 349 & 10  & 32  & 2575 & 665  \\
+    ADP        &  6739 & 518 & 0   & 0   & 2578 & 649  \\
     \hline
@@
     in the UK Biobank data.
-    Errors are measured as the Euclidean distance of
+    Errors (ADP) are the Euclidean distances from
     the PC scores predicted by each method
     to those predicted by ADP.
-    The 2492 samples from the 1000 Genomes data
-    are used as the reference.
+    Errors (centers) are the Euclidean distances
+    from the centers of mass of the reference samples
+    to the centers of mass of the study samples.
+    For the global study,
+    2492 samples from the 1000 Genomes data are used as the reference.
+    For the European study,
+    only the 498 European samples among them are used.
     The number of variants in common is 145,282.
@@
   \caption{PC scores of the European in the 500 randomly selected global samples UK Biobank data (circles) predicted by the four methods with the European samples in the 1000 Genomes data (squares) as the reference.}
+\end{figure}
+
+\begin{figure}[p]
+  \centering
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_nchunks100_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_sp.png}
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_nchunks100_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_ap.png}
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_nchunks100_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_oadp.png}
+  \caption{All the 488366 samples in the UK Biobank data.}
+\end{figure}
+
+\begin{figure}[p]
+  \centering
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_pred_EUR_nchunks100_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_sp.png}
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_pred_EUR_nchunks100_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_ap.png}
+  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_pred_EUR_nchunks100_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_oadp.png}
+  \caption{All the 462284 European samples in the UK Biobank data.}
 \end{figure}

### 2018-06-20 doc/paper/plot.py

@@ -10,14 +10,31 @@ for i in range(4):
     plt.plot(n, ggsim[i], label=methods[i])
 plt.xticks(n)
 plt.legend()
+plt.title('Runtime', fontsize=20)
+fig = plt.gcf()
+fig.set_size_inches((12,5))
 plt.savefig('img/ggsim_runtimes.png', dpi=300)
 plt.close('all')
 
-ggsim = np.loadtxt('ggsim_errors.dat')
+ggsim = np.loadtxt('ggsim_errors_adp.dat')
 for i in range(4):
     plt.plot(n, ggsim[i], label=methods[i])
 plt.xticks(n)
 plt.legend()
-plt.savefig('img/ggsim_errors.png', dpi=300)
+plt.title('Error (from ADP)', fontsize=20)
+fig = plt.gcf()
+fig.set_size_inches((12,5))
+plt.savefig('img/ggsim_errors_adp.png', dpi=300)
+plt.close('all')
+
+ggsim = np.loadtxt('ggsim_errors_ctr.dat')
+for i in range(4):
+    plt.plot(n, ggsim[i], label=methods[i])
+plt.xticks(n)
+plt.legend()
+plt.title('Error (between centers)', fontsize=20)
+fig = plt.gcf()
+fig.set_size_inches((12,5))
+plt.savefig('img/ggsim_errors_ctr.png', dpi=300)
 plt.close('all')
 

### 2018-06-20 frugalpca.py
Add geocenter similarity

@@ -43,7 +43,7 @@ NUM_CORES = mp.cpu_count()
 
 PLOT_ALPHA_REF=0.05
 PLOT_ALPHA_STU=0.40
-PLOT_MARKERS = ['.', '+', 'x', '*', 'd', 's']
+PLOT_MARKERS = ['.', '+', 'x', 'd', '*', 's']
 LOG_LEVEL = 'info'
 
 # DIM_SVDRAND = DIM_STU * 4
@@ -52,6 +52,22 @@ NP_OUTPUT_FMT = '%.4f'
 DELIMITER = '\t'
 # DIR_TMP = mkdtemp()
 DIR_TMP = 'tmp'
+
+# def pca_stu_io(
+#         i, stu_filepref_list, ref_filepref, path_tmp, method,
+#         X_mean, X_std, U, s, V, pcs_ref,
+#         dim_ref=DIM_REF, dim_stu=DIM_STU, dim_stu_high=DIM_STU_HIGH):
+#     stu_filepref = stu_filepref_list[i]
+#     W = read_bed(stu_filepref, bed_store='memory', dtype=np.int8)[0]
+#     pcs_stu = pca_stu(
+#         W, X_mean, X_std, method=method, path_tmp=path_tmp,
+#         U=U, s=s, V=V, pcs_ref=pcs_ref,
+#         dim_ref=dim_ref, dim_stu=dim_stu, dim_stu_high=dim_stu_high)
+#     ref_basepref = os.path.basename(ref_filepref)
+#     out_filepref = stu_filepref + '_sturef_' + ref_basepref
+#     pcs_stu_filename = out_filepref + '_stu_' + method +'.pcs'
+#     np.savetxt(pcs_stu_filename, pcs_stu, fmt=NP_OUTPUT_FMT, delimiter='\t')
+#     return pcs_stu
 
 
 # def split_bed_indiv(filepref, chunklevel):
@@ -183,6 +199,25 @@ def procrustes(Y_mat, X_mat, return_transformed=False):
         return R, rho, c, X_new
     else:
         return R, rho, c
+
+def geocenter_coordinate(X, X_ctr):
+    p = X.shape[1]
+    X_ctr_unique = np.unique(X_ctr)
+    X_ctr_unique_n = len(X_ctr_unique)
+    X_ctr_unique_coord = np.zeros((X_ctr_unique_n, p))
+    for i,ctr in enumerate(X_ctr_unique):
[…]

### 2018-06-20 frugalpca_runner.py

@@ -12,10 +12,10 @@ print('Number of cores: ' + str(NUM_CORES))
 # fpt.convert_ggsim(4)
 # for i in range(5):
 #     fpt.test_pca_ggsim(i)
-# fpt.test_merge_array_results()
 # fpt.test_split_bed_indiv()
 
 
-fpt.test_pca_ggsim()
+# fpt.test_pca_ggsim()
 # fpt.test_pca_5c()
 # fpt.test_pca_5c_EUR()
+fpt.test_merge_array_results()

### 2018-06-20 frugalpca_tests.py

@@ -124,7 +124,9 @@ def test_online_svd_procrust():
     print('Passed!')
 
 
-def test_pca(pref_ref, pref_stu, cmp_trace=True, load_results=False, assert_results=False, pref_out_trace=None, dim_ref=4, plot_dim=2):
+def test_pca(pref_ref, pref_stu, cmp_trace=True, load_results=False, assert_results=False, pref_out_trace=None, dim_ref=4, plot_dim=4, plot_size=None, plot_title=None, plot_color_stu=None, plot_legend=True, alpha_stu=0.4, plot_centers=False):
+    print('Reference: ' + pref_ref)
+    print('Study: ' + pref_stu)
     popu_filename_ref = pref_ref + '.popu'
     subpopu_filename_ref = pref_ref + '_sub.popu'
     base_ref = os.path.basename(pref_ref)
@@ -156,9 +158,9 @@ def test_pca(pref_ref, pref_stu, cmp_trace=True, load_results=False, assert_resu
     pcs_min = np.vstack((pcs_ref, pcs_stu_sp, pcs_stu_ap, pcs_stu_oadp)).min(axis=0)
     pcs_max = np.vstack((pcs_ref, pcs_stu_sp, pcs_stu_ap, pcs_stu_oadp)).max(axis=0)
     plot_lim = np.vstack((pcs_min, pcs_max)) * 1.20
-    fp.plot_pcs(pcs_ref, pcs_stu_sp, popu_ref, popu_stu_pred_sp, 'sp', out_pref=pref_out, plot_lim=plot_lim, plot_dim=plot_dim)
-    fp.plot_pcs(pcs_ref, pcs_stu_ap, popu_ref, popu_stu_pred_ap, 'ap', out_pref=pref_out, plot_lim=plot_lim, plot_dim=plot_dim)
-    fp.plot_pcs(pcs_ref, pcs_stu_oadp, popu_ref, popu_stu_pred_oadp, 'oadp', out_pref=pref_out, plot_lim=plot_lim, plot_dim=plot_dim)
+    fp.plot_pcs(pcs_ref, pcs_stu_sp, popu_ref, popu_stu_pred_sp, 'sp', out_pref=pref_out, plot_lim=plot_lim, plot_dim=plot_dim, plot_size=plot_size, plot_title=plot_title, plot_color_stu=plot_color_stu, plot_legend=plot_legend, alpha_stu=alpha_stu, plot_centers=plot_centers)
[…]

### 2018-06-20 doc/paper/paper.tex
Finish writing results

@@
 \usepackage{subcaption}
+\usepackage{color}
 
@@
 \newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg\;min}}\;}
+\newcommand{\todo}[1]{{\color{red} (#1)}}
 
@@
 \begin{abstract}
-    Population stratification (PS) is an important confounder
-    for the association between genotypes and phenotypes.
-    To adjust for PS,
-    we need to accurately estimate the ancestral structure
-    in the study samples.
-    This can be done effectively
-    by using principle component (PC) analysis
-    to predict the PC scores of the study genotypes
-    from a set of reference genotypes.
-    There are two major existing methods for predicting PC scores:
-    simple projection (SP) and augmentation-decomposition-transformation (ADP).
-    However, when the reference size is large,
-    the former is computationally efficient but inaccurate
-    (PC scores shrinking toward zero),
-    while the latter is accurate but computationally costly ($\bO[n^3]$).
-    To solve these problems,
-    we created the genoPCA software
-    that implemented 
-    the adjusted projection (AP) method to remove the bias in SP, and
-    the online ADP (OADP) method to speed up ADP.
-    We compared the accuracy and computational efficiency of these four methods
-    by conducting a simulation study
-    with the GGS software.
-    We also used them to study the UKbiobank data by using the 1000Genomes data as the reference.
-    We found that AP achieved the accuracy of ADP without adding much runtime,
-    while OADP reduced the runtime of ADP by $x$-fold without losing accuracy.
-    Furturmore, when studying a large number of samples
-    from a narrower geographic area
-    where the signal-to-noise ratio is low,
-    AP and OADP performed reasonably well
-    while SP frequently misclassified the populations
-    and the runtime of ADP became impractical.
+
+ Population stratification (PS) is a major confounder in genome-wide association studies (GWAS) and can lead to false positive associations.
+To adjust for PS, principal component analysis (PCA)-based ancestry prediction has been widely used.
+This approach first performs PCA on the reference samples with detailed ancestry information and then uses the reference sample PCA results to predict the PC scores of the study samples.
+Simple Projection (SP) and Augmentation-Decomposition-Transformation (ADP), such as LASER and TRACE, are popular methods for predicting PC scores.
+However, they are either biased or computationally expensive.
+For example, the predicted PC scores from SP can be biased toward NULL.
+On the other hand, since ADP requires running PCA separately for each study sample on the augmented dataset, its computational cost is high.
+
+To address these problems, we propose two alternative approaches, Adjusted Projection (AP) and Online ADP (OADP), implemented in the genoPCA software.
+Using random matrix theory results, AP asymptotically estimates and adjusts for the bias of SP.
+Moreover, OADP uses online singular value decomposition, which can greatly reduce the computation cost of ADP.
+We carried out extensive simulation studies to show that these alternative approaches are unbiased and that the computation times are 10-100 times faster than ADP when 2000 reference samples are used.
+We also applied our approaches to 500,000 study samples from the UK biobank data with 2,500 samples from the 1000 Genomes data as the reference.
+AP and OADP required 16 and 92 CPU hours, respectively, while the projected computation time of ADP is 1,933 CPU hours.
+Furthermore, when we only used the European reference samples in the 1000 Genomes to infer sub-European ancestry, SP clearly showed bias, unlike the proposed approaches.
+By using AP and OADP, we can infer ancestry and adjust for PS in GWAS robustly and efficiently
 
@@
 We used the Grid Genotype Simulator (GGS)
-by \citet{ mathieson2012differential}
+by \citet{mathieson2012differential}
 to simulate population migration.
@@
 to predict the PC scores for the study samples.
-Then we used the k-nearest neighbors method
+We only calculated the top 4 PCs,
+and for OADP and ADP,
+we calculated the top 20 PC scores for the study samples
+and projected them to the dim-4 reference PC score space
+with a Procrustes transformation.
+Finally, we used the k-nearest neighbors method
 to predict which ethnic group a study sample comes from.
+
+We used two measures for the accuracy of PC score prediction.
+First, we measured the ADP error,
+which is the Euclidean distance
+from a method's predicted PC scores
+to ADP's PC scores,
+since ADP is believed to be the currently most reliable method.
+Next, for the center error,
+we calculated the centers of mass of each population's PC scores.
+This was done for both the reference and the study samples.
+Then the center error
+was set to be the Euclidean distance
+from the reference centers to the corresponding study centers.
+
+For the comparison of computational cost,
+we recorded each method's runtime for analyzing the study samples.
+(The runtime for PCA on the reference samples
+and for reading or writing files are not included.)
+All the methods were run on a single CPU core.
 
@@
 to study the UK Biobank data \citep{sudlow2015uk},
-which consists of 490,000 middle- or elderly-aged individuals
+which consists of 488,000 middle- or elderly-aged individuals
 in the United Kingdom.
@@
 we applied SP, AP, and OADP to all the study samples.
+The same measurements for accuracy and computational cost
+from the simulation study
+was applied here.
 
@@
 
-\subsection{Simulation study}
-
-% Runtimes vs ref size
-% Distance to ADP vs ref size
-% Distance to geo center vs ref size
-% PC1 vs PC2, 5 x 3 grid (uncolored study samples)
-% The predicted PC scores are shown in \Cref{fig:n1000} and \Cref{fig:ukb}.
+\subsection{Simulation studies}
 
@@
 
+The accuracy of the four methods were compared
+by using their ADP errors and center errors.
+Compared to SP,
+the other three methods' errors are very small
+and close to each other.
+OADP had smaller errors than AP
+for both the ADP errors and the center errors.
+Interestingly, the center errors of ADP were slightly higher
+than that of OADP.
+As the reference size increased,
+the errors of all the four methods decreased.
+SP exhibited the fastest decrease in both errors,
+and the four methods' difference in accuracy
+became smaller and smaller.
+
 For computational cost,
-we can see that the runtime for SP increases faster than linearly
+we observed that the runtime for ADP increases faster than linearly
 as the reference sample size increases.
 This is consistent
-with the theoretical computational complexity of SP,
+with the theoretical computational complexity of ADP,
 which expects the runtime to increase at the rate of $\bO[n^3]$.
+When the reference sample size reached 3000,
+ADP's runtime was 27 times of OADP's
+and 225 times of SP's and AP's.
 In comparison, the runtimes of SP, AP, and OADP
-increase only slightly.
-The theoretical computational complexity of SP and AP is $\bO[1]$
-and that of OADP is $\bO[n]$,
-when the data dimension and the study sample size are fixed.
+increased only slightly,
+which is consistent with their $\bO[1]$ and $\bO[n]$ computational complexity when the data dimension and study sample size are fixed.
+SP and AP were the most computational efficient methods.
+Their runtimes were approximately only one-eighth of OADP's.
 
@@
 
-When we used the global samples in the UK Biobank data,
+When we used the 500 randomly selected global samples
+in the UK Biobank data,
 we found that the PC scores predicted by the four methods' are almost identical.
@@
 that the shrinkage bias decreases
-as the signal-to-noise ratio increases.
-
-However, when we focused on the European (EUR) samples only,
-the results given by SP were clearly unreliable.
+as the signal-to-noise ratio increases \todo{Find citation}.
+
+However, when we focused on the European (EUR) samples
+among these 500 UK Biobank samples,
+(of which there are 475 many,)
+the results given by SP were clearly inaccurate.
 The samples from British in England and Scotland (GBR)
@@
 drifted very close to NULL.
-This causes SP to incorrectly classified almost all of them as CEU,
-since most GBR reference samples are far away from NULL.
+This caused SP to incorrectly classified almost all of them as CEU,
+since most GBR reference samples are farther away from NULL.
 A few of the study samples predicted by ADP to be GBR or CEU
@@
 were classified by SP as IBS.
-
-For OADP, its predicted PC1 and PC2 scores
+The same patterns for the accuracy
+of SP, AP, and OADP
+discussed above
+were observed
+when we used all the 488,000 study samples in the UK Biobank dataset
+and those predicted to be Europeans among them.
+
+For AP and OADP, their predicted PC1 and PC2 scores
 for the European study samples
-are almost identical
+were almost identical
 to those done by ADP.
-OADP's results for PC3 and PC4
-are also very close to ADP's results,
-though we observed a very slight decrease in the magnitude of the PC scores.
-This minor shrinkage did not affect
-OADP's classification for the IBS and TSI samples,
-since their PC1 and PC2 scores were accurately predicted by OADP.
-Neither was any CEU/GBR sample classified as IBS/TSI. 
-For the CEU and GBR samples,
-their predicted PC scores are tightly mixed together
-for all the 4 PCs,
-which were even difficult for ADP to separate them.
-Thus the slight difference in PC3 and PC4 scores
-between OADP and ADP
-caused the classification of many CEU and GBR samples
-to be different on the two methods.
-
-Furthermore, AP successfully adjusted for the shrinkage
-in PC1 and PC2 for the European samples,
-so its predicted PC scores are almost identical to ADP's.
-However, it did not detect the shrinkage in PC 3 and PC4,
-so its predicted PC scores are the same as SP's for these two PCs.
-This still allowed AP to correctly separate
-the CEU/GBR samples from the IBS/TSI samples.
-However, its separation of the IBS and TSI samples
-did not agree completely with that done by ADP,
-most likely caused by the shrinkage on PC3.
-For the classification between CEU and GBR,
-the AP's results differ far from ADP's,
-as AP classified many more CEU/GBR samples as CEU.
-
-The patterns for the accuracy of SP, AP, and OADP
-discussed above
-were also observed
-when we used all the 490,000 study samples in the UK Biobank dataset
-and those predicted to be Europeans among them.
-See \Cref{tbl:ukb} for details.
+The three methods' classification
+agreed on most of the samples,
+except for within the cluster of CEU and GBR.
+Since many residents with northern and western European ancestry living in Utah
+could originally migrated from Britain,
+these two groups' ancestral history could be very similar
+and thus difficult if not impossible to be separated by PCA.
+Hence any slightest change
+in the PC scores for these two groups' study samples
+could easily change the classification from one group to the other.
+This is a problem inherent to the population structure
+and is not caused by the limitations of our methods.
+
+Just as in the simulation study,
+SP's errors were the greatest.
+Its ADP error
+was twice of SP's and 9 times of OADP's for the global samples,
+and 14 times of SP's and OADP's for the European samples.
+For the center errors,
+SP's error was still much greater than the other three methods.
+For the European samples,
+ADP's center error was the lowest,
+followed by OADP and AP.
+Interestingly, this order of the three methods
+was reversed for the global samples.
+
+For the computational cost,
+ADP's runtime was 1.5 times of OADP's and 10 times of AP's and SP's
+when only the European samples
+of the 500 random global study samples
+and the 498 European reference samples of the 1000 Genomes data
+were used.
+This difference increased to 18 times, 103 times, and 108 times
+when we studied the all the 500 random global samples
+with all the 2492 1000 Genomes reference samples.
+As we included all the 488,000 study samples in the UK Biobank data,
+SP, AP, and OADP only took 13, 16, and 90 CPU hours,
+while ADP was estimated (from its 500-sample runtime)
+to take 1828 CPU hours,
+which makes ADP very computational costly to use in practice.
 
@@
     \hline
-    \multicolumn{6}{|c|}{Distance from ADP's results} \\
+    \multicolumn{6}{|c|}{Error (ADP)} \\
     \hline
@@
     \hline
-    \multicolumn{6}{|c|}{Distance from reference centers} \\
+    \multicolumn{6}{|c|}{Error (centers)} \\
     \hline
@@
     as the reference size increases.
-    Distances are measured by the Euclidean distance.
-    Reference centers are the centers of mass of the populations
-    in the reference samples.
+    Runtimes are measured in seconds.
+    Errors (ADP) are the Euclidean distances from
+    the PC scores predicted by each method
+    to those predicted by ADP.
+    Errors (centers) are the Euclidean distances
+    from the centers of mass of the reference samples
+    to the centers of mass of the study samples.
     The number of variants is 100,000,
@@
   \centering
-  \begin{tabular}{|r|r|r|r|r|r|r|}
-    \hline
-    Method & \multicolumn{2}{|c|}{Runtime} & \multicolumn{2}{|c|}{Error (ADP)} & \multicolumn{2}{|c|}{Error (centers)} \\
-    \hline
-    & Global & European & Global & European & Global & European \\
-    \hline
-    SP         &  62   & 48  & 92  & 450 & 3156 & 2304 \\ 
-    AP         &  65   & 48  & 48  & 33  & 2486 & 706  \\
-    OADP       &  363  & 349 & 10  & 32  & 2575 & 665  \\
-    ADP        &  6739 & 518 & 0   & 0   & 2578 & 649  \\
+  \begin{tabular}{|r|r|r|r|r|r|r|r|}
+    \hline
+    & \multicolumn{3}{|c|}{Runtime} & \multicolumn{2}{|c|}{Error (ADP)} & \multicolumn{2}{|c|}{Error (centers)} \\
+    \hline
+    Samples & All (hrs) & Global (sec) & European (sec) & Global & European & Global & European \\
+    \hline
+    Sample size & 488,000 & 500 & 475 & 500 & 475 & 500 & 475 \\
+    \hline
+    SP   & 13    &  62   & 48  & 92  & 450 & 3156 & 2304 \\ 
+    AP   & 16    &  65   & 48  & 48  & 33  & 2486 & 706  \\
+    OADP & 90    &  363  & 349 & 10  & 32  & 2575 & 665  \\
+    ADP  & 1828*  &  6739 & 518 & 0   & 0   & 2578 & 649  \\
     \hline
@@
   \caption{
-    The study runtimes (in seconds) and erros for the four methods
-    for 500 randomly selected study samples
+    The study runtimes and erros for the four methods
+    for 1) all the samples and 2) 500 randomly selected study samples
     in the UK Biobank data.
@@
     Errors (centers) are the Euclidean distances
-    from the centers of mass of the reference samples
-    to the centers of mass of the study samples.
+    from the centers of mass of the reference populations
+    to the corresponding centers of mass of the study populations
+    predicted by the $k$-nearest neighbor method.
     For the global study,
@@
     only the 498 European samples among them are used.
-    The number of variants in common is 145,282.
+    The number of variants
+    in common between the reference and study samples is 145,282.
     Only the top 4 PCs are calculated.
+    *Estimated from the runtime of ADP for the 500 global samples.
     }

### 2018-06-21 frugalpca.py
Reduce dpi of plots

@@ -550,7 +550,7 @@ def plot_pcs(pcs_ref, pcs_stu_list, popu_ref, popu_stu_list, method_list, out_pr
     if plot_size is not None:
         fig.set_size_inches(plot_size)
     fig_filename = out_pref+'_'.join([''] + method_list)+'.png'
-    plt.savefig(fig_filename, dpi=300)
+    plt.savefig(fig_filename, dpi=75)
     plt.close('all')
     logging.info('PC plots saved to ' + fig_filename)
 

### 2018-06-21 frugalpca_runner.py

@@ -15,7 +15,7 @@ print('Number of cores: ' + str(NUM_CORES))
 # fpt.test_split_bed_indiv()
 
 
-# fpt.test_pca_ggsim()
-# fpt.test_pca_5c()
-# fpt.test_pca_5c_EUR()
+fpt.test_pca_ggsim()
+fpt.test_pca_5c()
+fpt.test_pca_5c_EUR()
 fpt.test_merge_array_results()

### 2018-06-21 doc/paper/paper.tex

@@
 \newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg\;min}}\;}
-\newcommand{\todo}[1]{{\color{red} (#1)}}
-
-\title{Computational Methods for Principle Component Analysis Applied to
-  Genotype Data}
+\newcommand{\todo}[1]{{\color{red} (TODO: #1)}}
+
+\title{
+  Fast and robust ancestry inference
+  using principal component analysis
+}
+\author{Daiwei Zhang, Rounak Dey, Seunggeun Lee}
 \date{}
@@
 We also applied our approaches to 500,000 study samples from the UK biobank data with 2,500 samples from the 1000 Genomes data as the reference.
-AP and OADP required 16 and 92 CPU hours, respectively, while the projected computation time of ADP is 1,933 CPU hours.
+AP and OADP required 16 and 90 CPU hours, respectively, while the projected computation time of ADP is 1,828 CPU hours.
 Furthermore, when we only used the European reference samples in the 1000 Genomes to infer sub-European ancestry, SP clearly showed bias, unlike the proposed approaches.
@@
 
-ADP is regarded as the most accurate method for predicting study PC scores.
-It is does not have SP's shrinkage bias.
+ADP is regarded as the most accurate method for predicting study PC scores \todo{Find citation}.
+It does not have SP's shrinkage bias.
 However, the major disadvantage of ADP
@@
   (Computational complexity: $\bO[n^2p]$.)  
-\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
+\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$,
+  where $D$ and $V$ only need their first $k$ columns.
   ($\bO[n^3]$.)
-\item Calculate $U_3 = X V_3 D_3^{-1}$,
-  where $V_3$ is the first $k_3$ columns of $V$,
-  and $D_3$ is the diagonal matrix of the first $k_3$ diagonal entries of $D$.
-  ($\bO[k_3 n p]$.)
+\item Calculate $U = X V D^{-1}$ but for the first $k$ columns only.
+  ($\bO[k n p]$.)
 \item Calculate 
   \[
-    L = U_3^T y \quad \text{and} \quad K = y^T H,
+    L = U^T y \quad \text{and} \quad K = y^T H,
   \]
-  where $H$ is the normalized  $y - U_3L$
-  ($\bO[k_3 p]$.)
+  where $H$ is the normalized  $y - UL$
+  ($\bO[k p]$.)
 \item Calculate $Q^T Q$, where
@@
     \begin{bmatrix}
-      D_3 & L \\
+      D & L \\
       0 & K
@@
   \]
-  ($\bO[k_3^3]$.)
-\item Apply eigendecomosition to $Q^T Q$ to get $Q^T Q = \ddot{V}_3 \ddot{D}^2_3 \ddot{V}^T_3$.
-  ($\bO[k_3^3]$.)
+  ($\bO[k^3]$.)
+\item Apply eigendecomosition to $Q^T Q$ to get $Q^T Q = \ddot{V} \ddot{D}^2 \ddot{V}^T$.
+  ($\bO[k^3]$.)
 \item Calculate
   \[
-    \tilde{V}_3 =
+    \tilde{V} =
     \begin{bmatrix}
-      V_3 & 0 \\
+      V & 0 \\
       0 & 1
     \end{bmatrix}
-    \ddot{V}_3.
+    \ddot{V}.
   \]
-  ($\bO[nk_3^2]$.)
-\item Project the first $k_2$ columns of $\tilde{V}_3$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual,
-  where $k_1 < k_2  < k_3 \ll n$.
-  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
+  ($\bO[nk^2]$.)
+\item Transform the first $k_2$ columns of $\tilde{V}$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual,
+  where $k_1 < k_2  < k \ll n$.
+  (Computational complexity is negligible if $k_2$ is small compared to $n$.)
   \item Go to step 4 for the next study individual.
@@
 
-Total computational complexity: $\bO[n^2p + n^3 + k_3np + m(k_3p + k_3^3 + nk_3^2]) \approx \bO[n^2p + m(k_3p + k_3^2n)]$, provided $k_3 \ll n \ll p$.
-
-The computational complexity of OADP for the study individuals increases linearly with respect to the reference sample size, which makes the computation time of OADP close to those of SP.
-The closeness between the results given by OADP and ADP will be empirically shown later.
+The total computational complexity is $\bO[n^2p + m(kp + k^2n)]$
+provided $k \ll n \ll p$.
+The computational complexity of OADP for the study individuals increases linearly with respect to the reference sample size,
+which is much more efficient than ADP's cubic increasing rate.
+The closeness between the results given by OADP and ADP
+will be empirically shown in \Cref{sec:results}.
 
@@
 and for reading or writing files are not included.)
-All the methods were run on a single CPU core.
+We used our genoPCA software,
+which implemented SP, AP, and OADP in Python.
+and TRACE by \citet{wang2015improved},
+which implemented ADP in C++.
+All the programs were run on a single CPU core.
 
@@
 
-\section{Results}
+\section{Results}\label{sec:results}.
 
@@
 are very close to each other
-for every choice of reference sample size.
+for every choice of reference sample size (\Cref{fig:ggsim}).
 Compared to their results,
@@
 The accuracy of the four methods were compared
-by using their ADP errors and center errors.
+by using their ADP errors and center errors (\Cref{tbl:ggsim}).
 Compared to SP,
@@
 and the four methods' difference in accuracy
-became smaller and smaller.
-
-For computational cost,
+became smaller and smaller (\Cref{fig:ggsim_trend}).
+
+For computational cost (\Cref{tbl:ggsim}),
 we observed that the runtime for ADP increases faster than linearly
@@
 with the theoretical computational complexity of ADP,
-which expects the runtime to increase at the rate of $\bO[n^3]$.
+which expects the runtime to increase at the rate of $\bO[n^3]$ (\Cref{tbl:cplx}).
 When the reference sample size reached 3000,
@@
 in the UK Biobank data,
-we found that the PC scores predicted by the four methods' are almost identical.
+we found that the PC scores predicted by the four methods' are almost identical (\Cref{fig:ukb_500}).
 The separation of the ethnic groups was clear,
@@
 (of which there are 475 many,)
-the results given by SP were clearly inaccurate.
+the results given by SP were clearly inaccurate (\Cref{fig:ukb_500_eur}).
 The samples from British in England and Scotland (GBR)
@@
 were classified by SP as IBS.
-The same patterns for the accuracy
-of SP, AP, and OADP
-discussed above
-were observed
-when we used all the 488,000 study samples in the UK Biobank dataset
-and those predicted to be Europeans among them.
 
@@
 and is not caused by the limitations of our methods.
+The patterns discussed above
+were still observed
+when we used all the 488,000 study samples in the UK Biobank dataset
+and those predicted to be Europeans among them (\Cref{fig:ukb,fig:ukb_eur}).
 
 Just as in the simulation study,
-SP's errors were the greatest.
+SP's errors were the greatest (\Cref{tbl:ukb}).
 Its ADP error
@@
 and the 498 European reference samples of the 1000 Genomes data
-were used.
+were used (\Cref{tbl:ukb}).
 This difference increased to 18 times, 103 times, and 108 times
@@
 
+From the simulation study and the study on the UK Biobank data,
+we see that AP and OADP have achieved the accuracy of ADP
+and the computational efficiency of SP.
+The advantage of AP's and OADP's accuracy
+is the most clear
+when the ratio of the reference sample size
+to the number of features is low
+or the signal-to-noise ratio is low.
+On the other hand,
+the advantage of AP's and OADP's computational efficiency
+is the most observable when the reference sample size is large.
+After the comparison of the four methods,
+we conclude that for a very small reference size,
+SP is unreliable since the shrinkage would be great,
+and ADP is a decent method to use
+because its computational inefficiency is negligible,
+but AP and OADP would still be much faster than ADP
+and their accuracy would be very close to ADP's.
+As the reference sample size increases
+(with respect to the number of variants),
+all the four methods' accuracy converge together,
+and SP becomes more and more accurate,
+although for genotype data,
+due to its high dimensions,
+the reference size must be extremely large
+in order for SP to achieve an accuracy
+that is at the same level as the other three methods.
+Therefore, for the sample sizes
+of the data sets that we have currently,
+OADP and SP are the most robust and fast methods
+for ancestry inference by PCA.
+
 For the future, we would like
-to improve the parallelization of our software.
-Currently it uses the poor-man’s parallelism,
-that is,
-dividing the study dataset into multiple smaller files on the hard drive,
-applying the analysis to each of them,
-and merge the results.
-This parallelism saves both the memory usage and the runtime.
-However, we would like to utilize
-more sophisticated parallel computation techniques
-to further improve the performance of our software.
+to improve the parallelism of our genoPCA software.
+We are also interested in applying randomized SVD algorithms
+such as the one by \citet{halko2011finding}
+for PCA on the reference data,
+since as the reference sample size becomes larger and larger,
+it becomes impractical to load such a large matrix into memory
+to compute its SVD or ED.
 
@@
   The data is simulated by the GGS software \citep{mathieson2012differential}.
+  \label{fig:ggsim}
 }
@@
     \hline
-    \multicolumn{6}{|c|}{Study Runtime} \\
+    \multicolumn{6}{|c|}{Runtime (sec)} \\
     \hline
@@
   \caption{
-    The runtimes and errors of the four methods
+    The study runtimes and errors of the four methods
     as the reference size increases.
@@
     }
+    \label{tbl:ggsim}
 \end{table}
@@
     The data is simulated by the GGS software \citep{mathieson2012differential}.
+    \label{fig:ggsim_trend}
   }
@@
     }
+    \label{tbl:ukb}
 \end{table}
@@
   \caption{PC scores of 500 randomly selected global samples in UK Biobank data (circles) predicted by the four methods with the global samples in the 1000 Genomes data (squares) as the reference.}
+  \label{fig:ukb_500}
 \end{figure}
@@
   \caption{PC scores of the European in the 500 randomly selected global samples UK Biobank data (circles) predicted by the four methods with the European samples in the 1000 Genomes data (squares) as the reference.}
+  \label{fig:ukb_500_eur}
 \end{figure}
@@
   \caption{All the 488366 samples in the UK Biobank data.}
+  \label{fig:ukb}
 \end{figure}
@@
   \caption{All the 462284 European samples in the UK Biobank data.}
+  \label{fig:ukb_eur}
 \end{figure}

### 2018-06-21 doc/paper/paper.tex
Correct typos in paper

@@
 For example, the predicted PC scores from SP can be biased toward NULL.
-On the other hand, since ADP requires running PCA separately for each study sample on the augmented dataset, its computational cost is high.
+On the other hand, since ADP requires running PCA separately for each study sample on the augmented data set, its computational cost is high.
 
@@
 We carried out extensive simulation studies to show that these alternative approaches are unbiased and that the computation times are 10-100 times faster than ADP when 2000 reference samples are used.
-We also applied our approaches to 500,000 study samples from the UK biobank data with 2,500 samples from the 1000 Genomes data as the reference.
+We also applied our approaches to 500,000 study samples from the UK Biobank data with 2,500 samples from the 1000 Genomes data as the reference.
 AP and OADP required 16 and 90 CPU hours, respectively, while the projected computation time of ADP is 1,828 CPU hours.
@@
 This prediction-based approach allows not only adjustment for PS
-but also infererence of the ancestry memberships of the study samples.
+but also inference of the ancestry memberships of the study samples.
 
@@
 However, though computationally efficient,
-a major disadvantae of SP is the loss of accuracy when the number of features $p$ greatly exceeds the reference sample size $n$,
+a major disadvantage of SP is the loss of accuracy when the number of features $p$ greatly exceeds the reference sample size $n$,
 a situation that is not uncommon for genotype data.
@@
 A minor disadvantage of AP is that
-while the computatinal complexity and memory usage of SP
+while the computational complexity and memory usage of SP
 can be further reduced
@@
 by \citet{halko2011finding}
-as discussed in \cref{sec:randsvd})
 to compute the SVD for only the top $k$ PCs of the reference matrix,
@@
   (CC: $\bO[n^3]$.)
-\item Find the Procrutes transformation
+\item Find the Procrustes transformation
 from the first $lk$ columns of $\tilde{V}$
@@
 the computational cost for a study sample
-increases cubically.
+increases cubicly.
 Fortunately, this disadvantage can be mitigated by taking some computational shortcuts,
@@
   (Computational complexity: $\bO[n^2p]$.)  
-\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$,
+\item Apply EGD to $X^T X$ to get $X^T X = V D^2 V^T$,
   where $D$ and $V$ only need their first $k$ columns.
@@
   ($\bO[k^3]$.)
-\item Apply eigendecomosition to $Q^T Q$ to get $Q^T Q = \ddot{V} \ddot{D}^2 \ddot{V}^T$.
+\item Apply EGD to $Q^T Q$ to get $Q^T Q = \ddot{V} \ddot{D}^2 \ddot{V}^T$.
   ($\bO[k^3]$.)
@@
 were still observed
-when we used all the 488,000 study samples in the UK Biobank dataset
+when we used all the 488,000 study samples in the UK Biobank data set
 and those predicted to be Europeans among them (\Cref{fig:ukb,fig:ukb_eur}).
@@
 it becomes impractical to load such a large matrix into memory
-to compute its SVD or ED.
+to compute its SVD or EGD.
 
@@
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=.98\linewidth]{img/ggsim_runtimes.png}
-  \centering
-  \includegraphics[width=.98\linewidth]{img/ggsim_errors_adp.png}
-  \centering
-  \includegraphics[width=.98\linewidth]{img/ggsim_errors_ctr.png}
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=.90\linewidth]{img/ggsim_runtimes.png}
+  \centering
+  \includegraphics[width=.90\linewidth]{img/ggsim_errors_adp.png}
+  \centering
+  \includegraphics[width=.90\linewidth]{img/ggsim_errors_ctr.png}
   \caption{
@@
   \centering
-  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_sp}
-  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_ap}
-  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_oadp}
-  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_adp}
+  \includegraphics[width=0.90\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_sp}
+  \includegraphics[width=0.90\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_ap}
+  \includegraphics[width=0.90\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_oadp}
+  \includegraphics[width=0.90\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_adp}
   \caption{PC scores of 500 randomly selected global samples in UK Biobank data (circles) predicted by the four methods with the global samples in the 1000 Genomes data (squares) as the reference.}
@@
   \centering
-  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_sp}
-  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_ap}
-  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_oadp}
-  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_adp}
+  \includegraphics[width=0.9\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_sp}
+  \includegraphics[width=0.9\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_ap}
+  \includegraphics[width=0.9\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_oadp}
+  \includegraphics[width=0.9\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_adp}
   \caption{PC scores of the European in the 500 randomly selected global samples UK Biobank data (circles) predicted by the four methods with the European samples in the 1000 Genomes data (squares) as the reference.}

### 2018-06-22 doc/paper/paper.tex
Edit author list

@@
 }
-\author{Daiwei Zhang, Rounak Dey, Seunggeun Lee}
+\author{Daiwei Zhang, Rounak Dey, Sehee Kim, Seunggeun Lee}
 \date{}
@@
 \begin{document}
-
 \maketitle
@@
 \begin{abstract}
-
 Population stratification (PS) is a major confounder in genome-wide association studies (GWAS) and can lead to false positive associations.
